---
title: "Week 2: Regression, Variance, Covariance, Correlation, & Equation Systems Review"
date: today
description: '<a href="lectures/w02_regression_review.pdf" download="w02_regression_review.pdf">Download PDF</a>'
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    self-contained-math: true
release: week 2
draft: true
---


## Agenda

1. Methods walk through expectations & sign up
2. Updates: lecture updated
3. Variance, covariance, correlation, regression, multiple regression, and interactions

::: {.content-visible when-profile="teacher"}
::: {.fragment}
1: 10 min
:::
:::


```{r}
#| include: false
library(tidyverse)

pisa <- read_csv("rlabs/w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)

theme_set(theme_minimal(base_size = 16))

# BOOKSHOME is coded 1, 2, 3, 4, 6, 7 in this extract -- there is no 5.
# Derive the levels from the data so plots never invent a category.
books_levels <- sort(unique(na.omit(pisa$BOOKSHOME)))

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")

coef_table <- function(model) {
  broom::tidy(model) |>
    mutate(
      term = recode(term,
        "(Intercept)"        = "Intercept",
        "PAREDINT"           = "Parent education",
        "BOOKSHOME"          = "Books at home",
        "PAREDINT:BOOKSHOME" = "Parent education x Books"
      ),
      across(c(estimate, std.error, statistic), ~ round(.x, 2)),
      p.value = ifelse(p.value < .001, "< .001", fmt(p.value, 3))
    ) |>
    rename(Term = term, Estimate = estimate, SE = std.error,
           t = statistic, p = p.value) |>
    knitr::kable()
}
```

## Variance

```{r}
#| echo: false
#| fig-height: 3.2
#| fig-width: 9
set.seed(625)
samp <- pisa |>
  filter(!is.na(READ_AVG)) |>
  slice_sample(n = 60) |>
  mutate(case = row_number())

read_mean <- mean(samp$READ_AVG)

ggplot(samp, aes(x = case, y = READ_AVG)) +
  geom_segment(aes(xend = case, yend = read_mean),
               colour = "grey65", linewidth = 0.4) +
  geom_point(size = 2.4, alpha = 0.8, colour = "grey20") +
  geom_hline(yintercept = read_mean, colour = "firebrick", linewidth = 1.2) +
  annotate("text", x = 0, y = read_mean,
           label = paste0("mean = ", fmt(read_mean, 1)),
           colour = "firebrick", hjust = 0, vjust = -0.6, size = 5) +
  labs(x = "Student", y = "Reading score")
```

Each grey line is one student's distance from the mean.
For all `r fmt(sum(!is.na(pisa$READ_AVG)), 0)` students, $s^2$ = `r fmt(var(pisa$READ_AVG, na.rm = TRUE), 1)` and $s$ = `r fmt(sd(pisa$READ_AVG, na.rm = TRUE), 1)`.

$$s^2 = \frac{\sum_{i=1}^{n}(y_i - \bar{y})^2}{n - 1}$$

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: a measure of how scattered the data is around the mean.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
The grey segments are the deviations that get squared in the formula. Squaring
is why the variance (11,296) is on an uninterpretable scale while the SD (106)
is back in reading-score points.

Original plan: Students should know this well. Show a scatterplot of the pisa
reading out come with a mean bar and cases on the bottom (do a random selection
that visualizes nicely) and ask them to describe what variance is. Target
answer: measure of how scattered data is around the mean.
:::
:::

## Covariance

```{r}
#| echo: false
#| fig-height: 3.4
#| fig-width: 9
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick", linewidth = 1.2) +
  labs(x = "Parent education (years)", y = "Reading score")
```

Covariance of reading score and parent education = **`r fmt(cov(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 2)`**

$$\mathrm{cov}(x, y) = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{n - 1}$$

How would you describe what covariance is? What are its units?

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how two variables change in relation to each other. Not
standardized. Units are reading-score points x years of schooling, which is
why the number itself is hard to interpret.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
Original plan: Also old hat to students. Show pisa reading outcome x PAREDINT
and ask them to describe what covariance is. Show the line & covariance number.
Target answer: how two variables change in relation to eachother. Not
standarized. Units of ??
:::
:::

## Correlation

:::: {.columns}

::: {.column width="50%"}
**Raw units**

```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.5
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Parent education (years)", y = "Reading score")
```

Covariance = **`r fmt(cov(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 2)`**
:::

::: {.column width="50%"}
**Standardized (z scores)**

```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.5
#| warning: false
pisa |>
  mutate(z_pared = scale(PAREDINT)[, 1], z_read = scale(READ_AVG)[, 1]) |>
  ggplot(aes(x = z_pared, y = z_read)) +
  geom_jitter(width = 0.15, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Parent education (SD)", y = "Reading score (SD)")
```

Correlation = **`r fmt(cor(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 3)`**
:::

::::

$$r = \frac{\mathrm{cov}(x, y)}{s_x \, s_y}$$

Same data, same shape. What is the difference between the two numbers?

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: correlation is standardized, so it ranges between -1 and 1 and
is much easier to interpret. Covariance carries the original units.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
The point clouds are identical apart from the axis labels. Dividing by the two
SDs (`r fmt(sd(pisa$PAREDINT, na.rm = TRUE), 2)` and
`r fmt(sd(pisa$READ_AVG, na.rm = TRUE), 2)`) is the only difference:
49.54 / (2.67 x 106.28) = 0.174.

Original plan: Also very well known to students. Show pisa reading outcome x
PAREDINT correlation and covariance side x side and ask them to describe the
difference between the two. Target answer: correlation is standardized, ranges
between -1 and 1. easier to interpret.
:::
:::

## Regression

$$\hat{y} = b_0 + b_1 x_1$$

```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 9
#| warning: false
m1 <- lm(READ_AVG ~ PAREDINT, data = pisa)

ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick", linewidth = 1.2) +
  labs(x = "Parent education (years)", y = "Reading score")
```

```{r}
#| echo: false
coef_table(m1)
```

$R^2$ = `r fmt(summary(m1)$r.squared, 3)`, residual SE = `r fmt(sigma(m1), 1)`

How do you interpret this?

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: we are predicting reading outcome from parent education. Each
additional year of parent education predicts about
`r fmt(coef(m1)[["PAREDINT"]], 2)` more reading-score points. $R^2$ and the
residual SE are the indicators of how good those predictions are.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.fragment}
The reminder you asked for: the slope is just the covariance rescaled.

$$b_1 = \frac{\mathrm{cov}(x, y)}{s^2_x} = \frac{49.54}{7.11} = 6.97
\qquad
r = b_1 \frac{s_x}{s_y} = 6.97 \times \frac{2.67}{106.28} = 0.174$$

Same information as the covariance and the correlation, in prediction units.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
$R^2$ is only 0.03 here, so parent education alone explains 3% of the variance
in reading. Worth flagging that a real, significant effect can still be a weak
predictor.

Original plan: show a regression of pisa reading outcome ~ PAREDINT. Show the
plot and also the output. Ask students to interpret. Target answer: Predicting
reading outcome from PAREDINT; indicators of how good the predictions are.
Coefficient the same as correlation/covariance (Remind me). show the structural
equation y = b0 + b1x1 in nice latex formatting.
:::
:::

## Multiple Regression

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2$$

```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 9
#| warning: false
m2 <- lm(READ_AVG ~ PAREDINT + BOOKSHOME, data = pisa)

grid2 <- expand_grid(
  PAREDINT  = seq(3, 16, length.out = 50),
  BOOKSHOME = books_levels
) |>
  mutate(READ_AVG = predict(m2, newdata = pick(everything())))

pisa |>
  filter(!is.na(BOOKSHOME)) |>
  ggplot(aes(x = PAREDINT, y = READ_AVG, colour = factor(BOOKSHOME, levels = books_levels))) +
  geom_jitter(width = 0.45, alpha = 0.10) +
  geom_line(data = grid2, aes(group = factor(BOOKSHOME, levels = books_levels)), linewidth = 1.2) +
  scale_colour_viridis_d(end = 0.9) +
  labs(x = "Parent education (years)", y = "Reading score",
       colour = "Books at home")
```

```{r}
#| echo: false
coef_table(m2)
```

$R^2$ = `r fmt(summary(m2)$r.squared, 3)` (was `r fmt(summary(m1)$r.squared, 3)` with parent education alone)

How do you interpret this?

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how parent education predicts reading outcome controlling for
books at home, and vice versa. The lines are parallel because each predictor
gets one slope that holds at every level of the other.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
The parent-education slope drops from `r fmt(coef(m1)[["PAREDINT"]], 2)` to
`r fmt(coef(m2)[["PAREDINT"]], 2)` once books at home is in the model. Much of
what looked like a parent-education effect was shared with books at home. This
is the setup for the mediation problem at the end.

Original plan: Show a regression of pisa reading outcome ~ PAREDINT +
BOOKSHOME. Show a plot and also output. Ask students to interpret. Target
answer: how PAREDINT predicts reading outcome controlling for BOOKSHOME & vice
versa. show the structural equation y = b0 + b1x1 + b2x2 in nice latex
formatting.
:::
:::

## Interaction in Multiple Regression

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2 + b_3 (x_1 \times x_2)$$

```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 9
#| warning: false
m3 <- lm(READ_AVG ~ PAREDINT * BOOKSHOME, data = pisa)

grid3 <- expand_grid(
  PAREDINT  = seq(3, 16, length.out = 50),
  BOOKSHOME = books_levels
) |>
  mutate(READ_AVG = predict(m3, newdata = pick(everything())))

pisa |>
  filter(!is.na(BOOKSHOME)) |>
  ggplot(aes(x = PAREDINT, y = READ_AVG, colour = factor(BOOKSHOME, levels = books_levels))) +
  geom_jitter(width = 0.45, alpha = 0.10) +
  geom_line(data = grid3, aes(group = factor(BOOKSHOME, levels = books_levels)), linewidth = 1.2) +
  scale_colour_viridis_d(end = 0.9) +
  labs(x = "Parent education (years)", y = "Reading score",
       colour = "Books at home")
```

```{r}
#| echo: false
coef_table(m3)
```

$R^2$ = `r fmt(summary(m3)$r.squared, 3)`

How do you interpret this? Are the main effects more or less important than the interaction effect?

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how parent education predicts reading controlling for books at
home and vice versa, AND how the two together predict reading. The lines are
no longer parallel: the parent-education slope depends on how many books are
in the home.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Here the main effects do the work. Adding the interaction moves $R^2$ from
`r fmt(summary(m2)$r.squared, 3)` to `r fmt(summary(m3)$r.squared, 3)` and the
interaction term is not significant (p = `r fmt(summary(m3)$coefficients["PAREDINT:BOOKSHOME", 4], 3)`).
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Key point: an interaction says the two predictors act *jointly* on reading.
That is a different claim from a SEM-style model where parent education
predicts books at home, and books at home predicts reading.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
The main-effect coefficients look wild here (parent education goes slightly
negative) because with an interaction in the model they are conditional on the
other predictor being 0, which is outside the data range. If asked: centering
fixes the interpretation, not the fit.

Original plan: Show a regression of pisa reading outcome ~ PAREDINT + BOOKSHOME
+ PAREDINT x BOOKSHOME. Ask students to interpret. Target answer: how PAREDINT
predicts reading outcome controlling for BOOKSHOME & vice versa AND how PAREDINT
x BOOKSHOME together predict reading. Are main effects more or less important
than the interaction effect? A key point is to distinguish this
combined/interaction effect from SEM approaches where PAREDINT predicts
BOOKSHOME and BOOKSHOME predicts reading (for example) show the structural
equation y = b0 + b1x1 + B1x1*b2x2 in nice latex formatting.
:::
:::

## Problem

- We know there's a relationship between parent education & books in the home. We know that together they predict reading score. 
- It is possible that parent education is one cause of the number of books in the home
  - This is a mediation effect 

$$\widehat{\text{Books}} = a_0 + a_1 (\text{ParentEd})$$

$$\widehat{\text{Reading}} = b_0 + b_1 (\text{Books}) + b_2 (\text{ParentEd})$$

Books at home is an *outcome* in the first equation and a *predictor* in the second. That cannot be captured by regression, multiple regression, or interaction effects.

::: {.content-visible when-profile="teacher"}
::: {.fragment}
A single regression forces every variable to be either a predictor or an
outcome, never both. Two equations estimated separately also cannot share
information or test the indirect path. This is what systems of equations, and
then SEM, are for.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
Original plan: show the two equations for this. ^ cannot be captured by
regression, multiple regression or interaction effects.
:::
:::
