---
title: "Week 2: Review & SEM FUNdamentals"
date: today
description: '<a href="lectures/w02_regression_review.pdf" download="w02_regression_review.pdf">Download PDF</a>'
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    self-contained-math: true
release: week 2
---

## Agenda

1. Methods walk through expectations & sign up
2. Updates: lecture updated, office hrs updated
3. Variance, covariance, correlation, regression, multiple regression, and interactions
4. Systems of Equations
5. Solving 2 ways
4. R lab: review 
5. Homework: Example paper of Measured Variable Path Analysis

::: {.content-visible when-profile="teacher"}
::: {.fragment}
The point of this week is to draw a contrast between models you've studied in the past and SEM. Understand the key thing that distinguishes SEM from regression. 
:::
:::


```{r}
#| include: false
library(tidyverse)

pisa <- read_csv("../rlabs/w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)

theme_set(theme_minimal(base_size = 16))

# BOOKSHOME is coded 1, 2, 3, 4, 6, 7 in this extract -- there is no 5.
# Derive the levels from the data so plots never invent a category.
books_levels <- sort(unique(na.omit(pisa$BOOKSHOME)))

# Mean-centred versions for the interaction model. Books at home starts at 1
# and parent education at 3, so in an uncentred interaction the main effects
# are slopes at "the other predictor = 0" -- a point off both scales, which
# flips the parent-education coefficient negative for purely arithmetic
# reasons. Centring moves that reference point to the mean, so the main
# effects read as average slopes. The fit is untouched.
pared_mean <- mean(pisa$PAREDINT, na.rm = TRUE)
books_mean <- mean(pisa$BOOKSHOME, na.rm = TRUE)

pisa <- pisa |>
  mutate(
    PAREDINT_c  = PAREDINT - pared_mean,
    BOOKSHOME_c = BOOKSHOME - books_mean
  )

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")

coef_table <- function(model) {
  broom::tidy(model) |>
    mutate(
      term = recode(term,
        "(Intercept)"            = "Intercept",
        "PAREDINT"               = "Parent education",
        "BOOKSHOME"              = "Books at home",
        "PAREDINT:BOOKSHOME"     = "Parent education x Books",
        "PAREDINT_c"             = "Parent education",
        "BOOKSHOME_c"            = "Books at home",
        "PAREDINT_c:BOOKSHOME_c" = "Parent education x Books"
      ),
      across(c(estimate, std.error, statistic), ~ round(.x, 2)),
      p.value = ifelse(p.value < .001, "< .001", fmt(p.value, 3))
    ) |>
    rename(Term = term, Estimate = estimate, SE = std.error,
           t = statistic, p = p.value) |>
    knitr::kable()
}

# Overall model test. The coefficient table gives one t-test per predictor;
# this is the omnibus F comparing the whole model against an intercept-only
# model, so students see model-level and coefficient-level significance
# together.
model_fit <- function(model) {
  s <- summary(model)
  f <- s$fstatistic
  p <- pf(f[["value"]], f[["numdf"]], f[["dendf"]], lower.tail = FALSE)
  sprintf(
    "$R^2$ = %s, adj. $R^2$ = %s, $F$(%d, %s) = %s, $p$ %s, residual SE = %s",
    fmt(s$r.squared, 3),
    fmt(s$adj.r.squared, 3),
    as.integer(f[["numdf"]]),
    formatC(f[["dendf"]], format = "d", big.mark = ","),
    fmt(f[["value"]], 1),
    if (p < .001) "< .001" else paste("=", fmt(p, 3)),
    fmt(sigma(model), 1)
  )
}
```

## Variance

```{r}
#| echo: false
#| fig-height: 3.2
#| fig-width: 9
set.seed(625)
samp <- pisa |>
  filter(!is.na(READ_AVG)) |>
  slice_sample(n = 60) |>
  mutate(case = row_number())

read_mean <- mean(pisa$READ_AVG, na.rm = TRUE)

ggplot(samp, aes(x = case, y = READ_AVG)) +
  geom_segment(aes(xend = case, yend = read_mean),
               colour = "grey65", linewidth = 0.4) +
  geom_point(size = 2.4, alpha = 0.8, colour = "grey20") +
  geom_hline(yintercept = read_mean, colour = "magenta", linewidth = 1.2) +
  annotate("text", x = 0, y = read_mean,
           label = paste0("mean = ", fmt(read_mean, 1)),
           colour = "magenta", hjust = 0, vjust = -0.6, size = 5) +
  labs(x = "Student", y = "Reading score")
```

$$s^2 = \frac{\sum_{i=1}^{n}(y_i - \bar{y})^2}{n - 1}$$

$s^2$ = `r fmt(var(pisa$READ_AVG, na.rm = TRUE), 1)` 
$s$ = `r fmt(sd(pisa$READ_AVG, na.rm = TRUE), 1)`

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: a measure of how scattered the data is around the mean. This is what we want to explain. WHY are students getting different scores. Causally: what causes differences in student scores. 

The grey segments are the deviations that get squared in the formula. Squaring
is why the variance (11,296) is on an uninterpretable scale while the SD (106)
is back in reading-score points.

Original plan: Students should know this well. Show a scatterplot of the pisa
reading out come with a mean bar and cases on the bottom (do a random selection
that visualizes nicely) and ask them to describe what variance is. Target
answer: measure of how scattered data is around the mean.
:::
:::

## Covariance

```{r}
#| echo: false
#| fig-height: 3.4
#| fig-width: 9
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "magenta", linewidth = 1.2) +
  labs(x = "Parent education (years)", y = "Reading score")
```

$$\mathrm{cov}(x, y) = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{n - 1}$$

cov = `r fmt(cov(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 2)`


::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how two variables change in relation to each other. Not
standardized. Units are reading-score points x years of schooling, which is
why the number itself is hard to interpret.


Original plan: Also old hat to students. Show pisa reading outcome x PAREDINT
and ask them to describe what covariance is. Show the line & covariance number.
Target answer: how two variables change in relation to eachother. Not
standarized. Units of ??
:::
:::

## Correlation $$r = \frac{\mathrm{cov}(x, y)}{s_x \, s_y}$$

:::: {.columns}

::: {.column width="50%"}
Covariance

```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.5
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "magenta") +
  labs(x = "Parent education (years)", y = "Reading score")
```

Covariance = **`r fmt(cov(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 2)`**
:::

::: {.column width="50%"}
Correlation

```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.5
#| warning: false
pisa |>
  mutate(z_pared = scale(PAREDINT)[, 1], z_read = scale(READ_AVG)[, 1]) |>
  ggplot(aes(x = z_pared, y = z_read)) +
  geom_jitter(width = 0.15, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "magenta") +
  labs(x = "Parent education", y = "Reading score")
```

Correlation = **`r fmt(cor(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 3)`**
:::

::::

$$r = \frac{\mathrm{cov}(x, y)}{s_x \, s_y}$$


::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: correlation is standardized, so it ranges between -1 and 1 and
is much easier to interpret. Covariance carries the original units.

The point clouds are identical apart from the axis labels. Dividing by the two
SDs (`r fmt(sd(pisa$PAREDINT, na.rm = TRUE), 2)` and
`r fmt(sd(pisa$READ_AVG, na.rm = TRUE), 2)`) is the only difference:
49.54 / (2.67 x 106.28) = 0.174.

Original plan: Also very well known to students. Show pisa reading outcome x
PAREDINT correlation and covariance side x side and ask them to describe the
difference between the two. Target answer: correlation is standardized, ranges
between -1 and 1. easier to interpret.
:::
:::

## Regression

$$\hat{y} = b_0 + b_1 x_1$$

```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 9
#| warning: false
m1 <- lm(READ_AVG ~ PAREDINT_c, data = pisa)

# Plotted on the raw scale: centring shifts the intercept, not the line.
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = TRUE, colour = "magenta", linewidth = 1.2) +
  labs(x = "Parent education (years)", y = "Reading score")
```

```{r}
#| echo: false
coef_table(m1)
```

`r model_fit(m1)`


::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: we are predicting reading outcome from parent education. Each
additional year of parent education predicts about
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` more reading-score points. $R^2$ and the
residual SE are the indicators of how good those predictions are.

Parent education is mean-centred, so the intercept
(`r fmt(coef(m1)[["(Intercept)"]], 1)`) is the predicted reading score at
average parent education (`r fmt(pared_mean, 1)` years) rather than at zero
years, which is off the scale. The slope is unchanged by centring.

The reminder you asked for: the slope is just the covariance rescaled.

$$b_1 = \frac{\mathrm{cov}(x, y)}{s^2_x} = \frac{49.54}{7.11} = 6.97
\qquad
r = b_1 \frac{s_x}{s_y} = 6.97 \times \frac{2.67}{106.28} = 0.174$$

Same information as the covariance and the correlation, in prediction units.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
$R^2$ is only 0.03 here, so parent education alone explains 3% of the variance
in reading. Worth flagging that a real, significant effect can still be a weak
predictor.

Original plan: show a regression of pisa reading outcome ~ PAREDINT. Show the
plot and also the output. Ask students to interpret. Target answer: Predicting
reading outcome from PAREDINT; indicators of how good the predictions are.
Coefficient the same as correlation/covariance (Remind me). show the structural
equation y = b0 + b1x1 in nice latex formatting.
:::
:::

## Multiple Regression 1

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2$$

```{r}
#| echo: false
pisa |>
  select(Reading = READ_AVG, `Parent ed` = PAREDINT, Books = BOOKSHOME) |>
  cor(use = "pairwise.complete.obs") |>
  round(2) |>
  knitr::kable()
```



## Multiple Regression 2

```{r}
#| echo: false
m2 <- lm(READ_AVG ~ PAREDINT_c + BOOKSHOME_c, data = pisa)

coef_table(m2)
```

`r model_fit(m2)`

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how parent education predicts reading outcome controlling for
books at home, and vice versa. The lines are parallel because each predictor
gets one slope that holds at every level of the other.

The parent-education slope drops from `r fmt(coef(m1)[["PAREDINT_c"]], 2)` to
`r fmt(coef(m2)[["PAREDINT_c"]], 2)` once books at home is in the model. Much of
what looked like a parent-education effect was shared with books at home. This
is the setup for the mediation problem at the end.

Both predictors are mean-centred, so the intercept
(`r fmt(coef(m2)[["(Intercept)"]], 1)`) is the predicted reading score for a
student at average parent education and average books at home. Note the books
mean is `r fmt(books_mean, 2)`, which is not one of the response options --
worth saying out loud if a student asks.

Original plan: Show a regression of pisa reading outcome ~ PAREDINT +
BOOKSHOME. Show a plot and also output. Ask students to interpret. Target
answer: how PAREDINT predicts reading outcome controlling for BOOKSHOME & vice
versa. show the structural equation y = b0 + b1x1 + b2x2 in nice latex
formatting.
:::
:::

## Interaction in Multiple Regression 1

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2 + b_3 (x_1 \times x_2)$$

## Interaction in Multiple Regression 2

```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 9
#| warning: false
m3 <- lm(READ_AVG ~ PAREDINT_c * BOOKSHOME_c, data = pisa)

# Build the grid on the raw scales so the axes stay readable, then centre
# to match the model before predicting.
grid3 <- expand_grid(
  PAREDINT  = seq(3, 16, length.out = 50),
  BOOKSHOME = books_levels
) |>
  mutate(
    PAREDINT_c  = PAREDINT - pared_mean,
    BOOKSHOME_c = BOOKSHOME - books_mean,
    READ_AVG    = predict(m3, newdata = pick(everything()))
  )

pisa |>
  filter(!is.na(BOOKSHOME)) |>
  ggplot(aes(x = PAREDINT, y = READ_AVG, colour = factor(BOOKSHOME, levels = books_levels))) +
  geom_jitter(width = 0.45, alpha = 0.10) +
  geom_line(data = grid3, aes(group = factor(BOOKSHOME, levels = books_levels)), linewidth = 1.2) +
  scale_colour_viridis_d(end = 0.9) +
  labs(x = "Parent education (years)", y = "Reading score",
       colour = "Books at home")
```

```{r}
#| echo: false
coef_table(m3)
```

`r model_fit(m3)`

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how parent education predicts reading controlling for books at
home and vice versa, AND how the two together predict reading. The lines are
no longer parallel: the parent-education slope depends on how many books are
in the home.

Here the main effects do the work. Adding the interaction moves $R^2$ from
`r fmt(summary(m2)$r.squared, 3)` to `r fmt(summary(m3)$r.squared, 3)` and the
interaction term is not significant (p = `r fmt(summary(m3)$coefficients["PAREDINT_c:BOOKSHOME_c", 4], 3)`).

Predictors are mean-centred here, so the main effects read as average slopes
rather than slopes at "the other predictor = 0". Centring changes only the
interpretation, never the fit.

Key point: an interaction says the two predictors act *jointly* on reading.
That is a different claim from a SEM-style model where parent education
predicts books at home, and books at home predicts reading.

Where this is heading: none of these models cleanly separates parent education
from books at home. The two correlate
`r fmt(cor(pisa$PAREDINT, pisa$BOOKSHOME, use = "pairwise.complete.obs"), 2)`,
so adding books cuts the parent-education slope from
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` to `r fmt(coef(m2)[["PAREDINT_c"]], 2)`.
Multiple regression assigns that shared variance to one predictor or the
other, but it cannot tell us the two are related, or in which direction. The
interaction does not rescue this: it tests whether the slopes differ, not how
the predictors are linked. That gap is what SEM and mediation are for.

Original plan: Show a regression of pisa reading outcome ~ PAREDINT + BOOKSHOME
+ PAREDINT x BOOKSHOME. Ask students to interpret. Target answer: how PAREDINT
predicts reading outcome controlling for BOOKSHOME & vice versa AND how PAREDINT
x BOOKSHOME together predict reading. Are main effects more or less important
than the interaction effect? A key point is to distinguish this
combined/interaction effect from SEM approaches where PAREDINT predicts
BOOKSHOME and BOOKSHOME predicts reading (for example) show the structural
equation y = b0 + b1x1 + B1x1*b2x2 in nice latex formatting.
:::
:::

## Problem

- No relationship between parent education & books? Does that seem right? 
- It is possible that parent education is one cause of the number of books in the home
  - This is a mediation effect 

$$\widehat{\text{Books}} = a_0 + a_1 (\text{ParentEd})$$

$$\widehat{\text{Reading}} = b_0 + b_1 (\text{Books}) + b_2 (\text{ParentEd})$$

Books at home is an *outcome* in the first equation and a *predictor* in the second. That cannot be captured by regression, multiple regression, or interaction effects.

::: {.content-visible when-profile="teacher"}
::: {.fragment}
A single regression forces every variable to be either a predictor or an
outcome, never both. Two equations estimated separately also cannot share
information or test the indirect path. This is what systems of equations, and
then SEM, are for.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
Original plan: show the two equations for this. ^ cannot be captured by
regression, multiple regression or interaction effects.
:::
:::


## Solving Systems of Equqtions

Algebraic approach. Demonstrate how to solve this simple example above (parent education predicts books in home; books in home predicts reading scores). Break over slides to do it step x step. Show the parameters to be solved for. Show how we solve for each. 

Problem: We can't algebraically solve all problems. More on that when we talk about model identification.

Solution: Maximum likelihood. The maximum likelihood that the observed (model-predicted outputs) match the real data. We maximize the likelihood by iteratively changing the model parameters. Leave as a placeholder. I have slides in which we covered this. 


R Lab
Focus on interpretation of the regression models. We will compare this to SEM results next week. 




## Up Next

Measured Variable Path Analysis Specification

Prep: read the article
