---
title: "Week 2: Review & SEM FUNdamentals"
date: today
description: '<a href="lectures/w02_regression_review.pdf" download="w02_regression_review.pdf">Download PDF</a>'
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    self-contained-math: true
release: week 2
draft: true
---

## Agenda

1. Methods walk through expectations & sign up
2. Logistics Updates
3. Data discussion
4. Variance, covariance, correlation, regression, multiple regression, and interactions
5. R lab 
6. Systems of Equations
7. Homework: Example paper of Measured Variable Path Analysis

::: {.content-visible when-profile="teacher"}
::: {.fragment}
The point of this week is to draw a contrast between models you've studied in the past and SEM. Understand the key thing that distinguishes SEM from regression. 

:::
:::


```{r}
#| include: false
library(tidyverse)

pisa <- read_csv("../rlabs/w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)

theme_set(theme_minimal(base_size = 16))

# BOOKSHOME is coded 1, 2, 3, 4, 6, 7 in this extract -- there is no 5.
# Derive the levels from the data so plots never invent a category.
books_levels <- sort(unique(na.omit(pisa$BOOKSHOME)))

# Mean-centred versions for the interaction model. Books at home starts at 1
# and parent education at 3, so in an uncentred interaction the main effects
# are slopes at "the other predictor = 0" -- a point off both scales, which
# flips the parent-education coefficient negative for purely arithmetic
# reasons. Centring moves that reference point to the mean, so the main
# effects read as average slopes. The fit is untouched.
pared_mean <- mean(pisa$PAREDINT, na.rm = TRUE)
books_mean <- mean(pisa$BOOKSHOME, na.rm = TRUE)

pisa <- pisa |>
  mutate(
    PAREDINT_c  = PAREDINT - pared_mean,
    BOOKSHOME_c = BOOKSHOME - books_mean
  )

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")

coef_table <- function(model) {
  broom::tidy(model) |>
    mutate(
      term = recode(term,
        "(Intercept)"            = "Intercept",
        "PAREDINT"               = "Parent education",
        "BOOKSHOME"              = "Books at home",
        "PAREDINT:BOOKSHOME"     = "Parent education x Books",
        "PAREDINT_c"             = "Parent education",
        "BOOKSHOME_c"            = "Books at home",
        "PAREDINT_c:BOOKSHOME_c" = "Parent education x Books"
      ),
      across(c(estimate, std.error, statistic), ~ round(.x, 2)),
      p.value = ifelse(p.value < .001, "< .001", fmt(p.value, 3))
    ) |>
    rename(Term = term, Estimate = estimate, SE = std.error,
           t = statistic, p = p.value) |>
    knitr::kable()
}

# Overall model test. The coefficient table gives one t-test per predictor;
# this is the omnibus F comparing the whole model against an intercept-only
# model, so students see model-level and coefficient-level significance
# together.
model_fit <- function(model) {
  s <- summary(model)
  f <- s$fstatistic
  p <- pf(f[["value"]], f[["numdf"]], f[["dendf"]], lower.tail = FALSE)
  sprintf(
    "$R^2$ = %s, adj. $R^2$ = %s, $F$(%d, %s) = %s, $p$ %s, residual SE = %s",
    fmt(s$r.squared, 3),
    fmt(s$adj.r.squared, 3),
    as.integer(f[["numdf"]]),
    formatC(f[["dendf"]], format = "d", big.mark = ","),
    fmt(f[["value"]], 1),
    if (p < .001) "< .001" else paste("=", fmt(p, 3)),
    fmt(sigma(model), 1)
  )
}
```

## Updates

1. Downloadable deck for w01 updated
2. AI Fair Use for class added to syllabus
  - living doc - open to future discussion
3. Office hours: Changed to Mon 2:30-3:30 or by apt. Sign up link added

## Data Discussions

- < 1 min overview of the datasets you found 
- What questions can you answer with them
- Share with the class


## Methods walk through

Assignment: Lamaku > Assignments > methods walkthrough

Template: Course website: https://srdevore.github.io/edep_625_sem/

## Variance

```{r}
#| echo: false
#| fig-height: 3.2
#| fig-width: 9
set.seed(625)
samp <- pisa |>
  filter(!is.na(READ_AVG)) |>
  slice_sample(n = 60) |>
  mutate(case = row_number())

read_mean <- mean(pisa$READ_AVG, na.rm = TRUE)

ggplot(samp, aes(x = case, y = READ_AVG)) +
  geom_segment(aes(xend = case, yend = read_mean),
               colour = "grey65", linewidth = 0.4) +
  geom_point(size = 2.4, alpha = 0.8, colour = "grey20") +
  geom_hline(yintercept = read_mean, colour = "magenta", linewidth = 1.2) +
  annotate("text", x = 0, y = read_mean,
           label = paste0("mean = ", fmt(read_mean, 1)),
           colour = "magenta", hjust = 0, vjust = -0.6, size = 5) +
  labs(x = "Student", y = "Reading score")
```

$$s^2 = \frac{\sum_{i=1}^{n}(y_i - \bar{y})^2}{n - 1}$$

$s^2$ = `r fmt(var(pisa$READ_AVG, na.rm = TRUE), 1)` 
$s$ = `r fmt(sd(pisa$READ_AVG, na.rm = TRUE), 1)`

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: a measure of how scattered the data is around the mean. This is what we want to explain. WHY are students getting different scores. Causally: what causes differences in student scores. 

The grey segments are the deviations that get squared in the formula. Squaring
is why the variance (11,296) is on an uninterpretable scale while the SD (106)
is back in reading-score points.

Original plan: Students should know this well. Show a scatterplot of the pisa
reading out come with a mean bar and cases on the bottom (do a random selection
that visualizes nicely) and ask them to describe what variance is. Target
answer: measure of how scattered data is around the mean.
:::
:::

## Covariance

```{r}
#| echo: false
#| fig-height: 3.4
#| fig-width: 9
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "magenta", linewidth = 1.2) +
  labs(x = "Parent education (years)", y = "Reading score")
```

$$\mathrm{cov}(x, y) = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{n - 1}$$

cov = `r fmt(cov(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 2)`


::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how two variables change in relation to each other. Not
standardized. Units are reading-score points x years of schooling, which is
why the number itself is hard to interpret.


Original plan: Also old hat to students. Show pisa reading outcome x PAREDINT
and ask them to describe what covariance is. Show the line & covariance number.
Target answer: how two variables change in relation to eachother. Not
standarized. Units of ??
:::
:::

## Correlation 

:::: {.columns}

::: {.column width="50%"}
Covariance

```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.5
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "magenta") +
  labs(x = "Parent education (years)", y = "Reading score")
```

Covariance = **`r fmt(cov(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 2)`**
:::

::: {.column width="50%"}
Correlation

```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.5
#| warning: false
pisa |>
  mutate(z_pared = scale(PAREDINT)[, 1], z_read = scale(READ_AVG)[, 1]) |>
  ggplot(aes(x = z_pared, y = z_read)) +
  geom_jitter(width = 0.15, alpha = 0.12) +
  geom_smooth(method = "lm", se = FALSE, colour = "magenta") +
  labs(x = "Parent education", y = "Reading score")
```

Correlation = **`r fmt(cor(pisa$READ_AVG, pisa$PAREDINT, use = "pairwise.complete.obs"), 3)`**
:::

::::
$$r = \frac{\mathrm{cov}(x, y)}{s_x \, s_y}$$


::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: correlation is standardized, so it ranges between -1 and 1 and
is much easier to interpret. Covariance carries the original units.

The point clouds are identical apart from the axis labels. Dividing by the two
SDs (`r fmt(sd(pisa$PAREDINT, na.rm = TRUE), 2)` and
`r fmt(sd(pisa$READ_AVG, na.rm = TRUE), 2)`) is the only difference:
49.54 / (2.67 x 106.28) = 0.174.

Original plan: Also very well known to students. Show pisa reading outcome x
PAREDINT correlation and covariance side x side and ask them to describe the
difference between the two. Target answer: correlation is standardized, ranges
between -1 and 1. easier to interpret.
:::
:::

## Correlation p2

```{r}
#| echo: false
pisa |>
  select(Reading = READ_AVG, `Parent ed` = PAREDINT, Books = BOOKSHOME) |>
  cor(use = "pairwise.complete.obs") |>
  round(2) |>
  knitr::kable()
```

## Regression

$$\hat{y} = b_0 + b_1 x_1$$

```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 9
#| warning: false
m1 <- lm(READ_AVG ~ PAREDINT_c, data = pisa)

# Plotted on the raw scale: centring shifts the intercept, not the line.
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.45, alpha = 0.12) +
  geom_smooth(method = "lm", se = TRUE, colour = "magenta", linewidth = 1.2) +
  labs(x = "Parent education (years)", y = "Reading score")
```

```{r}
#| echo: false
coef_table(m1)
```

`r model_fit(m1)`


::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: we are predicting reading outcome from parent education. Each
additional year of parent education predicts about
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` more reading-score points. $R^2$ and the
residual SE are the indicators of how good those predictions are.

Parent education is mean-centred, so the intercept
(`r fmt(coef(m1)[["(Intercept)"]], 1)`) is the predicted reading score at
average parent education (`r fmt(pared_mean, 1)` years) rather than at zero
years, which is off the scale. The slope is unchanged by centring.

The reminder you asked for: the slope is just the covariance rescaled.

$$b_1 = \frac{\mathrm{cov}(x, y)}{s^2_x} = \frac{49.54}{7.11} = 6.97
\qquad
r = b_1 \frac{s_x}{s_y} = 6.97 \times \frac{2.67}{106.28} = 0.174$$

Same information as the covariance and the correlation, in prediction units.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
$R^2$ is only 0.03 here, so parent education alone explains 3% of the variance
in reading. Worth flagging that a real, significant effect can still be a weak
predictor.

Original plan: show a regression of pisa reading outcome ~ PAREDINT. Show the
plot and also the output. Ask students to interpret. Target answer: Predicting
reading outcome from PAREDINT; indicators of how good the predictions are.
Coefficient the same as correlation/covariance (Remind me). show the structural
equation y = b0 + b1x1 in nice latex formatting.
:::
:::

## Multiple Regression 1

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2$$

```{r}
#| echo: false
m2 <- lm(READ_AVG ~ PAREDINT_c + BOOKSHOME_c, data = pisa)

coef_table(m2)
```


`r model_fit(m2)`

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how parent education predicts reading outcome controlling for
books at home, and vice versa. The lines are parallel because each predictor
gets one slope that holds at every level of the other.

The parent-education slope drops from `r fmt(coef(m1)[["PAREDINT_c"]], 2)` to
`r fmt(coef(m2)[["PAREDINT_c"]], 2)` once books at home is in the model. Much of
what looked like a parent-education effect was shared with books at home. This
is the setup for the mediation problem at the end.

Both predictors are mean-centred, so the intercept
(`r fmt(coef(m2)[["(Intercept)"]], 1)`) is the predicted reading score for a
student at average parent education and average books at home. Note the books
mean is `r fmt(books_mean, 2)`, which is not one of the response options --
worth saying out loud if a student asks.

Original plan: Show a regression of pisa reading outcome ~ PAREDINT +
BOOKSHOME. Show a plot and also output. Ask students to interpret. Target
answer: how PAREDINT predicts reading outcome controlling for BOOKSHOME & vice
versa. show the structural equation y = b0 + b1x1 + b2x2 in nice latex
formatting.
:::
:::

## Interaction in Multiple Regression 1

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2 + b_3 (x_1 \times x_2)$$

## Interaction in Multiple Regression 2

```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 9
#| warning: false
m3 <- lm(READ_AVG ~ PAREDINT_c * BOOKSHOME_c, data = pisa)

coef_table(m3)
```

`r model_fit(m3)`

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Target answer: how parent education predicts reading controlling for books at
home and vice versa, AND how the two together predict reading. The lines are
no longer parallel: the parent-education slope depends on how many books are
in the home.

Here the main effects do the work. Adding the interaction moves $R^2$ from
`r fmt(summary(m2)$r.squared, 3)` to `r fmt(summary(m3)$r.squared, 3)` and the
interaction term is not significant (p = `r fmt(summary(m3)$coefficients["PAREDINT_c:BOOKSHOME_c", 4], 3)`).

Predictors are mean-centred here, so the main effects read as average slopes
rather than slopes at "the other predictor = 0". Centring changes only the
interpretation, never the fit.

Key point: an interaction says the two predictors act *jointly* on reading.
That is a different claim from a SEM-style model where parent education
predicts books at home, and books at home predicts reading.

Where this is heading: none of these models cleanly separates parent education
from books at home. The two correlate
`r fmt(cor(pisa$PAREDINT, pisa$BOOKSHOME, use = "pairwise.complete.obs"), 2)`,
so adding books cuts the parent-education slope from
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` to `r fmt(coef(m2)[["PAREDINT_c"]], 2)`.
Multiple regression assigns that shared variance to one predictor or the
other, but it cannot tell us the two are related, or in which direction. The
interaction does not rescue this: it tests whether the slopes differ, not how
the predictors are linked. That gap is what SEM and mediation are for.

Original plan: Show a regression of pisa reading outcome ~ PAREDINT + BOOKSHOME
+ PAREDINT x BOOKSHOME. Ask students to interpret. Target answer: how PAREDINT
predicts reading outcome controlling for BOOKSHOME & vice versa AND how PAREDINT
x BOOKSHOME together predict reading. Are main effects more or less important
than the interaction effect? A key point is to distinguish this
combined/interaction effect from SEM approaches where PAREDINT predicts
BOOKSHOME and BOOKSHOME predicts reading (for example) show the structural
equation y = b0 + b1x1 + B1x1*b2x2 in nice latex formatting.
:::
:::

## Problem

- There is a correlation between books at home & parent education (r = 0.27)
- What is the real-world relationship between them? 

## Mediation Effect

- parent education > more books in home > reading score 

$$\widehat{\text{Books}} = a_0 + a_1(\text{ParentEd})$$
$$\widehat{\text{Reading}} = b_0 + b_1(\text{Books})$$

- Books at home is an *outcome* of parental education in the first equation and a *predictor* of reading score in the second. 
- Cannot be captured by regression, multiple regression, or interaction effects.

::: {.content-visible when-profile="teacher"}
::: {.fragment}
A single regression forces every variable to be either a predictor or an
outcome, never both. Two equations estimated separately also cannot share
information or test the indirect path. This is what systems of equations, and
then SEM, are for.
:::
:::

::: {.content-visible when-profile="teacher"}
::: {.notes}
Original plan: show the two equations for this. ^ cannot be captured by
regression, multiple regression or interaction effects.
:::
:::

## Rlab 2

- Write a short interpretation of each of the models. 
- Compare the models: Which is the best? Why?  
- Save somewhere you can easily find it again. We will come back to this in future weeks. 


## Solving Systems of Equqtions 1

```{r}
#| include: false
cov_pb <- cov(pisa$PAREDINT, pisa$BOOKSHOME, use = "pairwise.complete.obs")
var_p  <- var(pisa$PAREDINT, na.rm = TRUE)
a1 <- cov_pb / var_p
a0 <- books_mean - a1 * pared_mean

cov_br <- cov(pisa$BOOKSHOME, pisa$READ_AVG, use = "pairwise.complete.obs")
var_b  <- var(pisa$BOOKSHOME, na.rm = TRUE)
b1 <- cov_br / var_b
b0 <- read_mean - b1 * books_mean
```

Two equations, four unknowns

$$\text{Books} = a_0 + a_1(\text{ParentEd}) + d_B$$
$$\text{Reading} = b_0 + b_1(\text{Books}) + d_R$$

Unknowns: $a_0$, $a_1$, $b_0$, $b_1$.

The SEM idea: set the covariances the model *implies* equal to the covariances
we actually *observe*, then solve for the parameters.

## First Equation — Step 1: The model

$$\text{Books} = a_0 + a_1(\text{ParentEd}) + d_B$$

$d_B$ is the disturbance (what ParentEd does not explain). In SEM we assume it
is uncorrelated with the predictor:

$$\mathrm{cov}(\text{ParentEd},\, d_B) = 0$$

## First Equation — Step 2: Covary both sides with ParentEd

Take the covariance of each side with ParentEd:

$$\mathrm{cov}(\text{ParentEd},\, \text{Books})
= \mathrm{cov}\big(\text{ParentEd},\; a_0 + a_1\text{ParentEd} + d_B\big)$$

## First Equation — Step 3: Distribute the covariance

$$\mathrm{cov}(\text{ParentEd}, \text{Books}) =
\mathrm{cov}(\text{ParentEd}, a_0)
+ a_1\,\mathrm{cov}(\text{ParentEd}, \text{ParentEd})
+ \mathrm{cov}(\text{ParentEd}, d_B)$$

## First Equation — Step 4: Simplify each term

- $\mathrm{cov}(\text{ParentEd}, a_0) = 0$ &nbsp; (covariance with a constant)
- $\mathrm{cov}(\text{ParentEd}, \text{ParentEd}) = s^2_{\text{ParentEd}}$
- $\mathrm{cov}(\text{ParentEd}, d_B) = 0$ &nbsp; (our assumption)

$$\mathrm{cov}(\text{ParentEd}, \text{Books}) = a_1\, s^2_{\text{ParentEd}}$$

## First Equation — Step 5: Solve for $a_1$

Divide both sides by $s^2_{\text{ParentEd}}$:

$$a_1 = \frac{\mathrm{cov}(\text{ParentEd}, \text{Books})}{s^2_{\text{ParentEd}}}$$

## First Equation — Step 6: Plug in the numbers

$$a_1 = \frac{`r fmt(cov_pb, 3)`}{`r fmt(var_p, 2)`} = `r fmt(a1, 3)`$$

## First Equation — Step 7: Back out the intercept $a_0$

The line passes through the means, so:

$$a_0 = \overline{\text{Books}} - a_1 \times \overline{\text{ParentEd}}
= `r fmt(books_mean, 2)` - `r fmt(a1, 3)` \times `r fmt(pared_mean, 2)`
= `r fmt(a0, 2)`$$

## Second Equation — Step 1: The model

$$\text{Reading} = b_0 + b_1(\text{Books}) + d_R$$

Same assumption, one link later in the chain:

$$\mathrm{cov}(\text{Books},\, d_R) = 0$$

## Second Equation — Step 2: Covary both sides with Books

$$\mathrm{cov}(\text{Books},\, \text{Reading})
= \mathrm{cov}\big(\text{Books},\; b_0 + b_1\text{Books} + d_R\big)$$

## Second Equation — Step 3: Distribute the covariance

$$\mathrm{cov}(\text{Books}, \text{Reading}) =
\mathrm{cov}(\text{Books}, b_0)
+ b_1\,\mathrm{cov}(\text{Books}, \text{Books})
+ \mathrm{cov}(\text{Books}, d_R)$$

## Second Equation — Step 4: Simplify each term

- $\mathrm{cov}(\text{Books}, b_0) = 0$ &nbsp; (covariance with a constant)
- $\mathrm{cov}(\text{Books}, \text{Books}) = s^2_{\text{Books}}$
- $\mathrm{cov}(\text{Books}, d_R) = 0$ &nbsp; (our assumption)

$$\mathrm{cov}(\text{Books}, \text{Reading}) = b_1\, s^2_{\text{Books}}$$

## Second Equation — Step 5: Solve for $b_1$

$$b_1 = \frac{\mathrm{cov}(\text{Books}, \text{Reading})}{s^2_{\text{Books}}}$$

## Second Equation — Step 6: Plug in the numbers

$$b_1 = \frac{`r fmt(cov_br, 2)`}{`r fmt(var_b, 3)`} = `r fmt(b1, 2)`$$

## Second Equation — Step 7: Back out the intercept $b_0$

$$b_0 = \overline{\text{Reading}} - b_1 \times \overline{\text{Books}}
= `r fmt(read_mean, 1)` - `r fmt(b1, 2)` \times `r fmt(books_mean, 2)`
= `r fmt(b0, 1)`$$

## The Solved System

$$\widehat{\text{Books}} = `r fmt(a0, 2)` + `r fmt(a1, 3)`(\text{ParentEd})$$
$$\widehat{\text{Reading}} = `r fmt(b0, 1)` + `r fmt(b1, 2)`(\text{Books})$$

- Merely ~14 steps 
- simplest possible MVPA model

::: {.content-visible when-profile="teacher"}
::: {.notes}
Original plan: Algebraic approach. Demonstrate how to solve this simple
example of parent education predicts books in home; books in home predicts
reading scores. Break over slides to do it step x step. Show the parameters
to be solved for. Show how we solve for each.
:::
:::

## Solving Systems of Equations: Maximum Likelihood

Objective: 

- Find the $\beta$ values & other parameters that **maximize** the **likelihood** of the observed data.
- Given a candidate set of parameters, how likely is it that I'd see exactly this pattern of real values? 

## Maximum Likelihood Estimation (MLE)

- Initialize the parameters ($\theta$) with reasonable start values (e.g., from the observed covariances)

::: notes
Maximum likelihood: find parameters that make the observed data most probable.

NOTE FOR STUDENTS WHO TOOK LOG REG WITH ME: This is a slightly different
application of ML than what we did with logistic regression. There, we were
fitting individual 0/1 outcomes with predicted probabilities. In SEM we are
fitting the observed covariance matrix (S) with a model-implied covariance
matrix. Same big idea (guess parameters, iteratively improve the fit), but the
"data" we're trying to reproduce is different.

Start with reasonable guesses, then iteratively adjust the parameters to
improve fit.
:::

## Maximum Likelihood: Continued

After we make our first guess at the parameters

- compute the model-implied covariance matrix $\Sigma(\theta)$
- compare it to the observed covariance matrix $S$
- use the ML fit function to see how far off the model is from the data.
  - we want this to be as small as possible

## ML fit function

$$F_{ML} = \log|\Sigma(\theta)| + \text{tr}(S\,\Sigma(\theta)^{-1}) - \log|S| - p$$

Where:

- $S$ = observed covariance matrix (from our data)
- $\Sigma(\theta)$ = model-implied covariance matrix
- $\theta$ = parameters
- $p$ = number of observed variables

::: notes
NOTE FOR STUDENTS WHO TOOK LOG REG WITH ME: In log reg we maximized a
log-likelihood built from individual 0/1 outcomes. Here we MINIMIZE a fit
function built from covariance matrices. It's the same ML logic, just applied
to reproducing S instead of reproducing individual outcomes.

- $F_{ML}$ is a discrepancy (misfit) between what the model implies and what we
  actually observed.
- When $\Sigma(\theta) = S$ (model perfectly reproduces the observed
  covariances), the fit function equals 0.
- The further $\Sigma(\theta)$ is from $S$, the larger the value.
- So 0 is the best outcome, and larger values are worse. (Opposite direction
  from the log reg log-likelihood, where less-negative was better.)
:::

## Pause & Recap

Maximum likelihood estimation represents the objective of what we want to do. 

Conceptually, what is our objective? 

## Gradient Descent: Finding the Valley

The search: Iterative optemizer

```{r gradient-descent-plot, echo=FALSE, fig.width=10, fig.height=6}
# Create visualization of gradient descent (minimizing the fit function)
theta_vals <- seq(-3, 3, by = 0.05)
# Simulate a misfit curve (upright parabola) - minimized at theta = 1
misfit_vals <- (theta_vals - 1)^2

gradient_df <- data.frame(
  theta = theta_vals,
  misfit = misfit_vals
)

# Points showing gradient descent steps
descent_points <- data.frame(
  step = 0:5,
  theta = c(-2, -1, 0, 0.5, 0.8, 1),
  misfit = c((-2 - 1)^2, (-1 - 1)^2, (0 - 1)^2,
             (0.5 - 1)^2, (0.8 - 1)^2, (1 - 1)^2)
)

ggplot(gradient_df, aes(x = theta, y = misfit)) +
  geom_line(size = 1.5, color = "steelblue") +
  geom_point(data = descent_points, aes(x = theta, y = misfit),
             color = "red", size = 4) +
  geom_path(data = descent_points, aes(x = theta, y = misfit),
            color = "red", size = 1, arrow = arrow(length = unit(0.3, "cm"))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "darkgreen", size = 1) +
  annotate("text", x = 1, y = 8, label = "Optimal θ",
           color = "darkgreen", size = 5, fontface = "bold") +
  annotate("text", x = -2, y = 9.5, label = "Start: θ = -2",
           color = "red", size = 4) +
  annotate("text", x = descent_points$theta[6] + 0.6, y = descent_points$misfit[6] + 0.5,
           label = "Converged!", color = "red", size = 4) +
  labs(
    title = "Gradient Descent: Sliding Down to Minimum Misfit",
    subtitle = "Iteratively adjust θ to reduce the fit function",
    x = "θ (Parameter)",
    y = "Fit Function (Misfit)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

::: notes
NOTE FOR STUDENTS WHO TOOK LOG REG WITH ME: In log reg we pictured this as
gradient ASCENT - climbing UP to the peak of the log-likelihood. In SEM we
MINIMIZE misfit, so it's gradient DESCENT - sliding DOWN into the valley. Same
idea (follow the slope to the optimum), just flipped upside down because we're
minimizing instead of maximizing.

Start at arbitrary point
Calculate slope (gradient) at current position
Move in the direction that reduces misfit (downhill)
Repeat until we reach the bottom of the valley

**At any point on the curve:**

Calculate the derivative (gradient) = slope at that point. Conceptually, the
gradient tells us how the fit function changes as we nudge each parameter.

**Interpretation:**

- The gradient points uphill; we step in the opposite direction to reduce misfit
- Steep slope → we're far from the optimum, take a bigger step
- Gradient near 0 → we're at the bottom of the valley (best fit)!

Gradient tells us which direction to move
When gradient = 0, we've found the optimal parameters
:::

## Pause & Recap

Gradient Descent represents a search for the best paramters. 

Conceptually, how does it work? 

## Updating Parameters: The Learning Process

**Update Rule:**

$$\theta_j^{\text{(new)}} = \theta_j^{\text{(old)}} - \text{step size} \times \text{gradient}$$

Why a step size?

- Gradient tells us the direction & steepness of the curve
- Step size controls how far we move (a small increment, not a full unit)
- Too large → might overshoot the valley
- Too small → takes forever to converge

::: notes
This is the simple version. SEM software actually uses a more sophisticated
version of this "follow the slope" idea - typically Newton-Raphson or Fisher
scoring. Instead of just using the slope (gradient), these methods also use the
curvature of the fit surface (the second derivative / information matrix) to
figure out both the direction AND how big a step to take. That lets them take
much smarter steps and converge in far fewer iterations.
:::

## Pause & Recap

Objective: Maximum likelihood
Search: Gradient descent
How to update parameters: Update rule 

Conceptually, how do these three work together? 


::: notes

- Step size matters; SEM software chooses it smartly using curvature
- Small, well-chosen steps ensure we don't overshoot the minimum
- Typical stopping criterion: change in the fit function < 0.0001
- Usually converges in 10-100 iterations
:::


## Up Next

Measured Variable Path Analysis Specification

Prep: read the article

