---
title: "Week 3: MVPA: Specification"
date: today
description: '<a href="lectures/w03_mvpa_specification.pdf" download="w03_mvpa_specification.pdf">Download PDF</a>'
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    self-contained-math: true
release: week 3
draft: true
---

# Model specification

Which variables are included in the model
Which variables are related to which 
How are variables related to each other
Based on the relationships, what is the model-implied covariance? 

# Agenda

- Paper Discussion
- Terminology
- Systems of Equations as Path Diagrams
- Model-implied covariance
- Mini Project 1 
- R/lavaan gui


# Paper Discussion

- Theoretical motivation for the variables in the model? 
- Theoretical motivation for the relationships between variables in the model?
- Were any variables left out? 

# Terminology

## Measured vs. latent variables

**Measured** (**observed**) variables are recorded directly in the data.

- Our variables: **reading score**, **books at home**, **parent education**.

::: {.fragment}
**Latent** variables are *not* observed directly 
  
  - they are inferred from patterns among measured variables
  - typically not easily captured by a single variable

- Examples: **academic performance**, **intelligence**, **problem-solving ability**.
:::



::: {.content-visible when-profile="teacher"}
::: {.notes}
"Measured" and "observed" For this unit, measured. Later, we'll get into latent variables
:::
:::

## Review

- Input/independent variables & output/dependent variables

## Exogenous vs. endogenous variables

- **Exogenous variables:** nothing in the model explains it, but it explains other things.

::: {.fragment}
- **Endogenous variables:** are explained by another variable in the model. 
:::

::: {.fragment}
$$\text{Books} = a_0 + a_1(\text{ParentEd}) + \zeta_B$$
$$\text{Reading} = b_0 + b_1(\text{Books}) + \zeta_R$$
:::

## Review

Residuals: 

::: {.fragment}
difference between an observed value and the value your model predicts 
:::

## Disturbance

* Case-level: Actual value of the endogenous variable minus what's predicted by whatever the model says explains it

* Model level: Var() the part of an endogenous variable's variance that isn't explained by the model

* **Exogenous** variables have no disturbance — nothing in the model predicts them.

In our example, books at home and reading score each get a disturbance; parent education does not.

::: {.content-visible when-profile="teacher"}
::: {.notes}
$\zeta_B$ = everything besides parent education driving books at home;
$\zeta_R$ = everything besides books at home driving reading. Same idea as a
regression residual. 

As in Week 2, we assume the disturbance is uncorrelated with the predictors: cov(ParentEd, ζ_B) = 0.
:::
:::

## Review

**Parameters** are the quantities the model estimates: 

* What parameters do we estimate in a regression model? 

## SEM Parameters

**Parameters** are the quantities the model estimates: 

* path coefficients, 
* variances of exogenous variables, 
* disturbance variances, 
* Covariances among exogenous variables

## Parameters p2a

In our example, the **coefficients** are parameters the model estimates:

$$\text{Books} = \color{#C81E7A}{a_0} + \color{#C81E7A}{a_1}(\text{ParentEd}) + \zeta_B$$
$$\text{Reading} = \color{#C81E7A}{b_0} + \color{#C81E7A}{b_1}(\text{Books}) + \zeta_R$$

## Pause & Recap

Contrast these terms from regression and their counterparts in SEM

1. Input/Output (dependent/independent) vs exogenous/endogenous
2. Residuals vs disturbance

## Pause & Recap 2

1. What are parameters we esitmate in regression? 
2. What additional ones do we have in SEM?

## Systems of Equations as Diagrams

Same model, two representations — matching colors show what maps to what.

:::: {.columns}

::: {.column width="46%"}
::: {style="font-size: 0.82em"}
$$\text{Books} = a_0 + \color{#E07B1A}{a_1}\,\text{ParentEd} + \color{#C81E7A}{\zeta_B}$$

$$\text{Reading} = b_0 + \color{#2C6FB5}{b_1}\,\text{Books} + \color{#6FA80F}{\zeta_R}$$
:::
:::

::: {.column width="54%"}
```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.8, yend = 0.6, colour = MAG, arrow = ar) +
  geom_path(data = circle(4, 2.3), aes(x, y), colour = MAG, linewidth = 1) +
  annotate("text", x = 4, y = 2.3, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.8, yend = 0.6, colour = LIM, arrow = ar) +
  geom_path(data = circle(8, 2.3), aes(x, y), colour = LIM, linewidth = 1) +
  annotate("text", x = 8, y = 2.3, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
  theme_void()
```
:::

::::


## Key rules for Diagrams

* Measured Variables: rectangles/squares
* Latent Variables: Ovals/circles
* Path Coefficients: Straight, single-headed arrows from predictor to predicted (theory-driven)
* Disturbance: Arrow from Disturbance pointing to the endogenous variable (mandatory)
* Covariance: Curved, double-headed arrow (between two exogenous variables)

## Examples

Read each of the following models. For each, answer the same three questions.

1. What relationships does it predict? 
2. Which variables are Exogenous vs. endogenous? 
3. What parameters does it estimate? 


```{r}
#| include: false
library(ggplot2)

dk <- "#3A3A3A"; MAG <- "#C81E7A"; ORG <- "#E07B1A"; GRY <- "#9AA0A6"
AR  <- grid::arrow(length = grid::unit(0.20, "cm"), type = "closed")
AR2 <- grid::arrow(length = grid::unit(0.16, "cm"), ends = "both", type = "closed")

ecirc <- function(x, y, r = 0.45) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
node <- function(x, y, label, w = 2.6, h = 1.1) list(
  annotate("rect", xmin = x - w/2, xmax = x + w/2, ymin = y - h/2, ymax = y + h/2, fill = dk),
  annotate("text", x = x, y = y, label = label, colour = "white", size = 4.5, lineheight = 0.9)
)
path <- function(x1, y1, x2, y2) annotate("segment", x = x1, y = y1, xend = x2, yend = y2,
                                          colour = MAG, linewidth = 1.2, arrow = AR)
dist <- function(x, ycirc, ytarget, label) list(
  geom_path(data = ecirc(x, ycirc), aes(x, y), colour = GRY, linewidth = 0.9),
  annotate("segment", x = x, y = ycirc - 0.45, xend = x, yend = ytarget, colour = GRY, arrow = AR),
  annotate("text", x = x, y = ycirc, label = label, parse = TRUE, colour = GRY, size = 3.8)
)
canvas <- list(coord_equal(clip = "off"), theme_void())
```

## Example 1

::: {style="font-size: 0.85em"}
$$\text{Motivation} = a_0 + a_1(\text{Support}) + \zeta_M$$
$$\text{Grades} = b_0 + b_1(\text{Motivation}) + \zeta_G$$
:::

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 2.4
ggplot() +
  path(1.3, 0, 2.7, 0) +
  path(5.3, 0, 6.7, 0) +
  dist(4, 2.0, 0.6, "zeta[M]") +
  dist(8, 2.0, 0.6, "zeta[G]") +
  node(0, 0, "Teacher\nsupport") +
  node(4, 0, "Academic\nmotivation") +
  node(8, 0, "Course\ngrades") +
  canvas +
  expand_limits(x = c(-1.5, 9.5), y = c(-0.7, 2.6))
```

::: {.content-visible when-profile="teacher"}
::: {.notes}
Exogenous: teacher support. Endogenous: motivation, grades. Parameters: 2
paths, var(teacher support), 2 disturbance variances. Ask: is "motivation"
measured (a scale score → rectangle) or latent (a construct → oval)?
:::
:::

## Example 2

::: {style="font-size: 0.8em"}
$$\text{Engagement} = a_0 + a_1(\text{Involvement}) + a_2(\text{SES}) + \zeta_E$$
$$\text{Homework} = b_0 + b_1(\text{Engagement}) + \zeta_H$$
$$\text{Grade} = c_0 + c_1(\text{Homework}) + \zeta_G$$
:::

```{r}
#| echo: false
#| fig-width: 9.5
#| fig-height: 3.0
ggplot() +
  annotate("curve", x = -1.4, y = 1.1, xend = -1.4, yend = -1.1,
           colour = ORG, linewidth = 1, curvature = 0.6, arrow = AR2) +
  annotate("text", x = -2.7, y = 0, label = "cov", colour = ORG, size = 3.8, fontface = "italic") +
  path(1.4, 1.35, 2.4, 0.35) +
  path(1.4, -1.35, 2.4, -0.35) +
  path(5.2, 0, 6.2, 0) +
  path(9.0, 0, 10.0, 0) +
  dist(3.8, 2.0, 0.6, "zeta[E]") +
  dist(7.6, 2.0, 0.6, "zeta[H]") +
  dist(11.4, 2.0, 0.6, "zeta[G]") +
  node(0, 1.7, "Parental\ninvolvement", w = 2.8) +
  node(0, -1.7, "SES", w = 2.8) +
  node(3.8, 0, "Student\nengagement", w = 2.8) +
  node(7.6, 0, "Homework\ncompletion", w = 2.8) +
  node(11.4, 0, "Final\ngrade", w = 2.8) +
  canvas +
  expand_limits(x = c(-3.2, 12.9), y = c(-2.5, 2.6))
```

::: {.content-visible when-profile="teacher"}
::: {.notes}
Exogenous: parental involvement, SES (correlated → the curved covariance
arrow). Endogenous: engagement, homework completion, final grade. Parameters:
4 paths, 2 exogenous variances, 1 covariance, 3 disturbance variances.
:::
:::

## Example 3

::: {style="font-size: 0.8em"}
$$\text{Home resources} = a_0 + a_1(\text{SES}) + \zeta_H$$
$$\text{Reading skill} = b_0 + b_1(\text{Home resources}) + \zeta_R$$
$$\text{GPA} = c_0 + c_1(\text{Reading skill}) + \zeta_G$$
:::

```{r}
#| echo: false
#| fig-width: 9.5
#| fig-height: 2.2
ggplot() +
  path(1.3, 0, 2.7, 0) +
  path(5.3, 0, 6.7, 0) +
  path(9.3, 0, 10.7, 0) +
  dist(4, 2.0, 0.6, "zeta[H]") +
  dist(8, 2.0, 0.6, "zeta[R]") +
  dist(12, 2.0, 0.6, "zeta[G]") +
  node(0, 0, "SES") +
  node(4, 0, "Home\nresources") +
  node(8, 0, "Reading\nskill") +
  node(12, 0, "GPA") +
  canvas +
  expand_limits(x = c(-1.5, 13.5), y = c(-0.7, 2.6))
```

::: {.content-visible when-profile="teacher"}
::: {.notes}
Exogenous: SES. Endogenous: home resources, reading skill, GPA. Parameters: 3
paths, var(SES), 3 disturbance variances.
:::
:::

# Model Implied Covariance

## Observed Covariance

```{r}
#| include: false
library(tidyverse)
pisa <- read_csv("../rlabs/w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)
```

:::: {.columns}

::: {.column width="55%"}
Observed covariance matrix

```{r}
#| echo: false
pisa |>
  select(Reading = READ_AVG, Books = BOOKSHOME, `Parent ed` = PAREDINT) |>
  cov(use = "pairwise.complete.obs") |>
  round(2) |>
  knitr::kable()
```
:::

::: {.column width="45%"}
How each entry is computed

Variances (diagonal):

$$s^2_x = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}$$

Covariances (off-diagonal):

$$\mathrm{cov}(x, y) = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{n-1}$$
:::

::::

In our model, we want to maximize the likelihood of getting this covariance matrix.

## Model-implied covariance matrix

- Model-implied covariance is determined by tracing rules
- Tracing rules are derived algebraically. 


## Review

- Why do we need the model-implied covariance matrix? 
- Conceptually, how do we minimize distance between the observed & model-predicted matrices? 

Key concepts: maximum likelihood function, gradient descent, update rule


## Example: Books as a mediator

Here, I'd like to show the system of equations on the top, as it was in lectures/w02_regression_review.qmd for the measured variable path analysis parentEd > books > reading score. I'd also like the model diagram on the bottom. Key point: these are two ways of representing the exact same thing. 

## Drawing the diagrams

## The tracing rules

We get the model-implied covariance (or variance) between two variables by using these tracing rules:

1. Find every legitimate route connecting the two variables (or the same variable to itself, for a variance).
2. A route may go backward against an arrowhead, then forward with an arrowhead — but never forward-then-backward. You can reverse direction only once, and only at a common cause.
3. A route may pass through at most one curved double-headed (covariance) arrow, and only at the route's turning point.
4. Never pass through the same variable twice in one route (no loops).
5. Multiply all the coefficients along a route.
6. Sum the products across all legitimate routes — that sum is the model-implied covariance (or variance, if tracing a variable to itself).

##Getting the model-implied covariance matrix

I want to show the observed matrix again on the right, then on the left, the observed matrix but not the numbers, the equations used to get the numbers (even diagonals). It's ok if the second is larger than the first. I just want to seque into how we calculate things. I'll emphasie that for this, we have known values, so we just plug numbners into the equations.

## Getting the model-implied covariance matrix 2
Here, we have unknown parameters, and theoretically predicted relationships between variables. We draw a model based on those theoretically predicted relationships to get the model-implied covariance matrix. I want to show the model diagram on the left side and the covariance matrix on the right. I want to click to fill each cell of the model-impleid covariance matrix, and highlight the area of the model from which it is derived. 


# Mini-project 1

# Lavaan GUI

- Specifying a model in the GUI

# Rlab 

Brainstorm for Mini-project 1

- Use your data and a research question you submitted last week
- Familiarize with the Lavaan GUI
- How would you specify your own measured variable path analysis model? 
- Specify in the GUI
- Submit 
