---
title: "Week 4: MVPA: Identification & Fit"
date: today
description: '<a href="lectures/w04_mvpa_identification_fit.pdf" download="w04_mvpa_identification_fit.pdf">Download PDF</a>'
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    self-contained-math: true
release: week 4
---

# Agenda

- Recursive & Non-recursive models
- Degrees of Freedom: Review & in SEM
- Identification & Degrees of Freedom in SEM
- Other assumptions Assumptions
- Model-level Fit & Comparing models
- Rlab: Lavaan Package

# Review


# Recursive & Nonrecursive models

## Recursive

- Causal/predictive direction Flows in one direction; 
- disturbance terms are not correlated 
  - no hidden shared causes of unexplained variance.   

## Nonrecursive


:::: {.columns}
::: {.column width="50%"}

- Causal/predictive direction can flow in either direction

```{r}
#| echo: false
#| fig-width: 4.2
#| fig-height: 5
library(ggplot2)

BLU <- "#2C6FB5"; oNG <- "#E07B1A"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")

box <- function(y, label) list(
  annotate("rect", xmin = -1.15, xmax = 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = 0, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # b: Books -> Reading (left, pointing down)
  annotate("segment", x = -0.5, y = 1.6, xend = -0.5, yend = 0.6,
           colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = -0.85, y = 1.1, label = "b", parse = TRUE,
           colour = BLU, size = 6, fontface = "italic") +
  # b*: Reading -> Books (right, pointing up)
  annotate("segment", x = 0.5, y = 0.6, xend = 0.5, yend = 1.6,
           colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 0.85, y = 1.1, label = "b^'*'", parse = TRUE,
           colour = oNG, size = 6, fontface = "italic") +
  box(2.2, "Books at\nhome") + box(0, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.9, 1.9), y = c(-0.8, 3)) +
  theme_void()
```
:::

::: {.column width="50%"}
::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Covariance matrix**

$$
\begin{array}{c|cc}
 & \text{Books} & \text{Reading} \\ \hline
\text{Books}   & \sigma^2_{B} &  \\
\text{Reading} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::
:::
:::

::::


# Degrees of Freedom

## DF in Regression

How do we calculate DF in Regression models? 

::: {.content-visible when-profile="teacher"}
::: {.fragment}

$$df = n - k - 1$$

- n: number of cases in the study
- k: number of predictors in the regression model
- 1: subtract one for the intercept

:::

:::

::: {.content-visible when-profile="teacher"}
::: {.fragment}

$$df = \underbrace{\text{information}}_{\text{n cases}} - \underbrace{\text{constraints}}_{\text{n parameters}}$$

:::

:::

:::
:::

## DF in SEM

$$df = \underbrace{\text{information}}_{\textcolor{magenta}{\text{n variances/covariances}}} - \underbrace{\text{constraints}}_{\text{n parameters}}$$

## SEM information available

:::: {.columns}

::: {.column width="50%"}

- n variances & covariances in the matrix
- Information $= \dfrac{p(p+1)}{2}$


::: {.content-visible when-profile="teacher"}

::: {.fragment}

- Information $= \dfrac{3(4)}{2} = 6$

:::

:::

:::

::: {.column width="50%"}

$$
\begin{array}{c|cc}
 & \text{Books} & \text{Reading} \\ \hline
\text{Books}   & \sigma^2_{B} &  \\
\text{Reading} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::

::::


## DF Parameters

:::: {.columns}

::: {.column width="50%"}

* path coefficients,
* variances of exogenous variables,
* disturbance of endogenous variables,
* Covariances among exogenous variables

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
  theme_void()
```

:::

::::

Paramters: a1, b1, var(parent ed), Disturbance of books, disturbance of reading score

## Putting it together

$$df = \underbrace{\text{information}}_{\text{n variances/covariances}} - \underbrace{\text{constraints}}_{\text{n parameters}}$$

$$df = \dfrac{p(p+1)}{2} - q$$

- p: n variables
- q: n parameters to be estimated

$$df = \left(\dfrac{3(4)}{2}\right) - 5 = 1$$

# Identification

Algebraic Illustration:

- Treat A & B as parameters to estimate. 
- Treat Outcome as a piece of information (cases)

## Under-Identification

:::: {.columns}

::: {.column width="50%"}
**Algebra**

Solve Example 1:

$$A + B = 6$$

::: {.content-visible when-profile="teacher"}
::: {.fragment}
More parameters than information

:::
:::
:::

::: {.column width="50%"}
::: {.content-visible when-profile="teacher"}
::: {.fragment}
**In SEM**

```{r}
#| echo: false
#| fig-width: 4.2
#| fig-height: 5
library(ggplot2)

BLU <- "#2C6FB5"; oNG <- "#E07B1A"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")

box <- function(y, label) list(
  annotate("rect", xmin = -1.15, xmax = 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = 0, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # b: Books -> Reading (left, pointing down)
  annotate("segment", x = -0.5, y = 1.6, xend = -0.5, yend = 0.6,
           colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = -0.85, y = 1.1, label = "b", parse = TRUE,
           colour = BLU, size = 6, fontface = "italic") +
  # b*: Reading -> Books (right, pointing up)
  annotate("segment", x = 0.5, y = 0.6, xend = 0.5, yend = 1.6,
           colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 0.85, y = 1.1, label = "b^'*'", parse = TRUE,
           colour = oNG, size = 6, fontface = "italic") +
  box(2.2, "Books at\nhome") + box(0, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.9, 1.9), y = c(-0.8, 3)) +
  theme_void()
```


:::
:::
:::

::::


## Under-identification Summary

$$df < 0$$

- Infinite correct solutions
- No way to select a single solution over others


## Over-identification

:::: {.columns}

::: {.column width="50%"}

Solve Example 2:

$$\begin{aligned}
A + B &= 6 \\
2A + B &= 10 \\
3A + B &= 12
\end{aligned}$$

:::

::: {.column width="50%"}

::: {.fragment}

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
  theme_void()
```

$$
\begin{array}{c|ccc}
 & \text{Parent Ed} & \text{Books} & \text{Reading} \\ \hline
\text{Parent Ed} & \sigma^2_{P} & & \\
\text{Books}     & \sigma_{PB} & \sigma^2_{B} & \\
\text{Reading}   & \sigma_{PR} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::

:::

::::

## Over-idetnification Summary

$$df > 0$$

More pieces of information than parameters

- no perfect solution that satisfies all equations
- Add a critereon: get as close as possible to solving all three (OLS/ML)


## Just-identification

:::: {.columns}

::: {.column width="50%"}

::: {.content-visible when-profile="teacher"}
::: {.fragment}

Solve Example 3:

$$\begin{aligned}
A + B &= 6 \\
2A + B &= 10
\end{aligned}$$

:::
:::

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; PUR <- "#7A3FA0"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # direct path c1: ParentEd -> Reading (straight, below the boxes, purple)
  annotate("segment", x = 0.2, xend = 7.8, y = -0.75, yend = -0.75, colour = PUR, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 4, y = -1.1, label = "c[1]", parse = TRUE, colour = PUR, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-1.5, 3)) +
  theme_void()
```

$$
\begin{array}{c|ccc}
 & \text{Parent Ed} & \text{Books} & \text{Reading} \\ \hline
\text{Parent Ed} & \sigma^2_{P} & & \\
\text{Books}     & \sigma_{PB} & \sigma^2_{B} & \\
\text{Reading}   & \sigma_{PR} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::

::::


## Just-identified Summary

$$df = 0$$

- Always fits the data perfectly, regardless of whether the causal structure you specified is actually correct
- Not falsifiable 

## Pause & Recap

How do degrees of freedom differ between Regression & SEM? 
How are degrees of freedom used to determine how a model is identified? 


# MVPA Assumptions

## Correct Specification
## Identification

## Multicolinearity

## One last example

Solve Example 4: 

$$\begin{aligned}
A + B &= 6 \\
3A + 3B &= 18
\end{aligned}$$

::: {.content-visible when-profile="teacher"}
::: {.fragment}

Same as Example 1

No new information is provided by the second observation

:::
:::

## Sample Size

- Adequte sample size. ~20 cases per parameter. 

## Additional Assumptions you can adjust for

- Multivariate normality: required for ML; other estimators can handle violations
- Continuous predictors: set ordinal in Lavaan
- Linearity: Ways to handle non-linear relationships
- Uncorrelated disturnances: Applies only to recursive models

## Pause & Recap

What assumptions should be met or adjusted for for MVPA?

# Model Fit

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(lavaan)

pisa <- read_csv("../rlabs/w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")

# Model 1: fully mediated (ParentEd -> Books -> Reading)
model1 <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"
fit1 <- sem(model1, data = pisa)

# Model 2: partially mediated (adds direct ParentEd -> Reading)
model2 <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME + PAREDINT
"
fit2 <- sem(model2, data = pisa)
```

Q: to what extent is the relationship between parental education and children's reading scores mediated by the number of books in the home?

Develop a models and determine how well it fits the data


## Model

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
  theme_void()
```

- What theoretical claims does this model make?
- How many degrees of freedom? 
- How to describe its identification? 

## How well does the model fit the data?

```{r}
#| echo: false
round(fitMeasures(fit1, c("chisq", "df", "pvalue", "cfi", "tli",
                          "rmsea", "srmr", "aic", "bic")), 3)
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Over-identified (df = `r as.integer(fitMeasures(fit1, "df"))`), so fit is
testable. $\chi^2$ = `r fmt(fitMeasures(fit1, "chisq"), 1)`, $p$ =
`r fmt(fitMeasures(fit1, "pvalue"), 3)` -- the omitted direct path leaves
covariance unexplained (CFI = `r fmt(fitMeasures(fit1, "cfi"), 3)`, RMSEA =
`r fmt(fitMeasures(fit1, "rmsea"), 3)`, SRMR =
`r fmt(fitMeasures(fit1, "srmr"), 3)`).
:::
:::

## Model 2

Q: To what extent is the relationship between parental education and children's reading scores mediated by technology (ICTRES) and books in the home?




- What theoretical claims does this make?
-  How many degrees of freedom? 
-  How to describe its identification? 

## How well does model 2 fit the data?

## Model comparison: 

Does adding the additional variable actually improve the model



# R Lab p1: 

Run the code

Use code from this analysis > change 

# R Lab P2

The Lavaan Package




