---
title: "Week 4: MVPA: Identification & Fit"
date: today
description: '<a href="lectures/w04_mvpa_identification_fit.pdf" download="w04_mvpa_identification_fit.pdf">Download PDF</a>'
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    self-contained-math: true
release: week 4
draft: true
---

# Agenda

- Review
- Recursive & Non-recursive models
- Degrees of Freedom: Review & in SEM
- Identification & Degrees of Freedom in SEM
- Other Assumptions
- Model-level Fit & Comparing models
- Rlab: Lavaan Package

# Review

- Lamaku > Quiz > W4

Answer the question. 
3 minute time limit


# Recursive & Nonrecursive models

## Recursive

- Causal/predictive direction Flows in one direction; 
- disturbance terms are not correlated 
  - no hidden shared causes of unexplained variance.   

## Nonrecursive

:::: {.columns}
::: {.column width="50%"}

- Causal/predictive direction can flow in either direction

```{r}
#| echo: false
#| fig-width: 4.2
#| fig-height: 5
library(ggplot2)

BLU <- "#2C6FB5"; oNG <- "#E07B1A"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")

box <- function(y, label) list(
  annotate("rect", xmin = -1.15, xmax = 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = 0, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # b: Books -> Reading (left, pointing down)
  annotate("segment", x = -0.5, y = 1.6, xend = -0.5, yend = 0.6,
           colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = -0.85, y = 1.1, label = "b", parse = TRUE,
           colour = BLU, size = 6, fontface = "italic") +
  # b*: Reading -> Books (right, pointing up)
  annotate("segment", x = 0.5, y = 0.6, xend = 0.5, yend = 1.6,
           colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 0.85, y = 1.1, label = "b^'*'", parse = TRUE,
           colour = oNG, size = 6, fontface = "italic") +
  box(2.2, "Books at\nhome") + box(0, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.9, 1.9), y = c(-0.8, 3)) +
  theme_void()
```
:::

::: {.column width="50%"}
::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Covariance matrix**

$$
\begin{array}{c|cc}
 & \text{Books} & \text{Reading} \\ \hline
\text{Books}   & \sigma^2_{B} &  \\
\text{Reading} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::
:::
:::

::::


# Degrees of Freedom

## DF in Regression

How do we calculate DF in Regression models? 

::: {.content-visible when-profile="teacher"}
::: {.fragment}
$$df = n - k - 1$$

- n: number of cases in the study
- k: number of predictors in the regression model
- 1: subtract one for the intercept

:::
:::

::: {.content-visible when-profile="teacher"}
::: {.fragment}

$$df = \underbrace{\text{information}}_{\text{n cases}} - \underbrace{\text{constraints}}_{\text{n parameters}}$$

:::
:::

## DF in SEM

$$df = \underbrace{\text{information}}_{\color{#C81E7A}{\text{n variances/covariances}}} - \underbrace{\text{constraints}}_{\text{n parameters}}$$

## SEM information available

:::: {.columns}

::: {.column width="50%"}

- n variances & covariances in the matrix
- Information $= \dfrac{p(p+1)}{2}$


::: {.content-visible when-profile="teacher"}

::: {.fragment}

- Information $= \dfrac{3(4)}{2} = 6$

:::

:::

:::

::: {.column width="50%"}

$$
\begin{array}{c|cc}
 & \text{Books} & \text{Reading} \\ \hline
\text{Books}   & \sigma^2_{B} &  \\
\text{Reading} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::

::::


## DF Parameters

:::: {.columns}

::: {.column width="50%"}

* path coefficients,
* variances of exogenous variables,
* disturbance of endogenous variables,
* Covariances among exogenous variables

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
  theme_void()
```

:::

::::

Parameters: a1, b1, var(parent ed), Disturbance of books, disturbance of reading score

## Putting it together

$$df = \underbrace{\text{information}}_{\text{n variances/covariances}} - \underbrace{\text{constraints}}_{\text{n parameters}}$$

$$df = \dfrac{p(p+1)}{2} - q$$

- p: n variables
- q: n parameters to be estimated

$$df = \left(\dfrac{3(4)}{2}\right) - 5 = 1$$

# Identification

Algebraic Illustration:

- Treat A & B as parameters to estimate. 
- Treat Outcome as a piece of information (cases)

## Under-Identification

:::: {.columns}

::: {.column width="50%"}
**Algebra**

Solve Example 1:

$$A + B = 6$$

::: {.content-visible when-profile="teacher"}
::: {.fragment}
More parameters than information

:::
:::
:::

::: {.column width="50%"}
::: {.content-visible when-profile="teacher"}
::: {.fragment}
**In SEM**

```{r}
#| echo: false
#| fig-width: 4.2
#| fig-height: 5
library(ggplot2)

BLU <- "#2C6FB5"; oNG <- "#E07B1A"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")

box <- function(y, label) list(
  annotate("rect", xmin = -1.15, xmax = 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = 0, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # b: Books -> Reading (left, pointing down)
  annotate("segment", x = -0.5, y = 1.6, xend = -0.5, yend = 0.6,
           colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = -0.85, y = 1.1, label = "b", parse = TRUE,
           colour = BLU, size = 6, fontface = "italic") +
  # b*: Reading -> Books (right, pointing up)
  annotate("segment", x = 0.5, y = 0.6, xend = 0.5, yend = 1.6,
           colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 0.85, y = 1.1, label = "b^'*'", parse = TRUE,
           colour = oNG, size = 6, fontface = "italic") +
  box(2.2, "Books at\nhome") + box(0, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.9, 1.9), y = c(-0.8, 3)) +
  theme_void()
```


:::
:::
:::

::::


## Under-identification Summary

$$df < 0$$

- Infinite correct solutions
- No way to select a single solution over others


## Over-identification

:::: {.columns}

::: {.column width="50%"}

Solve Example 2:

$$\begin{aligned}
A + B &= 6 \\
2A + B &= 10 \\
3A + B &= 12
\end{aligned}$$

:::

::: {.column width="50%"}

::: {.fragment}

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.6
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.7, 2.6)) +
  theme_void()
```

$$
\begin{array}{c|ccc}
 & \text{Parent Ed} & \text{Books} & \text{Reading} \\ \hline
\text{Parent Ed} & \sigma^2_{P} & & \\
\text{Books}     & \sigma_{PB} & \sigma^2_{B} & \\
\text{Reading}   & \sigma_{PR} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::

:::

::::

## Over-identification Summary

$$df > 0$$

More pieces of information than parameters

- no perfect solution that satisfies all equations
- Add a criterion: get as close as possible to solving all three (OLS/ML)


## Just-identification

:::: {.columns}

::: {.column width="50%"}

::: {.content-visible when-profile="teacher"}

Solve Example 3:

$$\begin{aligned}
A + B &= 6 \\
2A + B &= 10
\end{aligned}$$

:::

:::

::: {.column width="50%"}

::: {.fragment}

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; PUR <- "#7A3FA0"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # direct path c1: ParentEd -> Reading (straight, below the boxes, purple)
  annotate("segment", x = 0.2, xend = 7.8, y = -0.75, yend = -0.75, colour = PUR, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 4, y = -1.1, label = "c[1]", parse = TRUE, colour = PUR, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-1.5, 3)) +
  theme_void()
```

$$
\begin{array}{c|ccc}
 & \text{Parent Ed} & \text{Books} & \text{Reading} \\ \hline
\text{Parent Ed} & \sigma^2_{P} & & \\
\text{Books}     & \sigma_{PB} & \sigma^2_{B} & \\
\text{Reading}   & \sigma_{PR} & \sigma_{BR} & \sigma^2_{R}
\end{array}
$$

:::

:::

::::


## Just-identified Summary

$$df = 0$$

- Always fits the data perfectly, regardless of whether the causal structure you specified is actually correct
- Not falsifiable 

## Pause & Recap

- How do degrees of freedom differ between Regression & SEM? 
- How are degrees of freedom used to determine how a model is identified? 


# MVPA Assumptions

## Correct Specification
## Identification

## Multicollinearity

Solve Example 4: 

$$\begin{aligned}
A + B &= 6 \\
3A + 3B &= 18
\end{aligned}$$

::: {.content-visible when-profile="teacher"}
::: {.fragment}

Same as Example 1

No new information is provided by the second observation

:::
:::

## Sample Size

- Adequate sample size. ~20 cases per parameter.

## Additional Assumptions you can adjust for

- Multivariate normality > required for ML; other estimators can handle violations
- Continuous predictors > Ordinal can be set in Lavaan
- Linearity: Ways to handle non-linear relationships
- Uncorrelated disturbances: Applies only to recursive models

## Pause & Recap

- What assumptions should be met or adjusted for for MVPA?

# Model Fit

## Null Hypothesis in Regression

::: {.content-visible when-profile="teacher"}
::: {.fragment}

H0: There is no effect of the predictor on the outcome

H1: There is an effect of the predictor on the outcome

Rejecting the null means we have found an effect (evidence SUPPORTS our hypothesis)

::: 
:::

## Null hypothesis in SEM 🤸


H0: The model-implied covariance matrix equals the observed covariance matrix

H1: The model-implied covariance matrix does not equal the observed covariance matrix

Rejecting the null means the model we specified is not well-aligned to observed data (evidence AGAINST our hypothesis)

## Key fit indices: Model χ² 

Model χ² (chi-square): If model-implied cov matrix differs significantly from observed. 

- Significant: reject the null (hypothesis not supported)
- Non-significant: Fail to rejecct the null (hypothesis supported) 
- Sensitive to sample size. 

## Key fit indices: CFI

CFI (Comparative Fit Index): Uses a baseline model 

- Creates a baseline model in which nothing is related
- Determines How much better your model is than assuming nothing is related. 
- 0-1 range; higher is better; >= .95 indicates good fit; >= .9 is acceptable

## Key fit indices: SRMR

SRMR (Standardized Root Mean Square Residual)

- Average size of the differences between the observed and model-predicted correlations
- Standardized so it's comparable across variables. 
- Lower is better; ≤ .08 indicates good fit

## Key fit indices: RMSEA 

RMSEA (Root Mean Square Error of Approximation)

- Error per degree of freedom 
- "how far off is the model, adjusting for how simple or complex it is." 
- Lower is better ≤ .05 indicates good fit; .05–.08 is acceptable; ≥ .10 indicates poor fit

## Key Fit Indices: AIC/BIC

Useful for model comparison

- How well a model fits considering how complex it is using −2(log-likelihood). 

Penalizes for complexity:

- AIC: Fixed penalty based on number of parameters
- BIC: Penalty for extra parameters grows with sample size 

# Models 

Q0: Does parent education relate to reading through the number of books in the home.

# Model 0 Diagram

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
  theme_void()
```

- What theoretical claims does this model make?
- How many degrees of freedom?
- How to describe its identification?

## Model 0 Results

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(lavaan)

pisa <- read_csv("../rlabs/w04_data_pisa22_USA_cln.csv", show_col_types = FALSE)

fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")

# Focused fit table: the headline indices with conventional good-fit cutoffs.
# With MLR we report the robust/scaled statistics.
fit_table <- function(fit) {
  m <- fitMeasures(fit, c("chisq.scaled", "df", "pvalue.scaled", "cfi.robust",
                          "rmsea.robust", "rmsea.ci.lower.robust",
                          "rmsea.ci.upper.robust", "srmr"))
  p <- if (m[["pvalue.scaled"]] < 0.001) "< .001" else
    paste0("= ", fmt(m[["pvalue.scaled"]], 3))
  rmsea <- sprintf("%s [%s, %s]", fmt(m[["rmsea.robust"]], 3),
                   fmt(m[["rmsea.ci.lower.robust"]], 3),
                   fmt(m[["rmsea.ci.upper.robust"]], 3))
  tibble::tibble(
    Index      = c("χ² (df)", "CFI", "RMSEA [90% CI]", "SRMR"),
    Value      = c(sprintf("%s (%d), p %s", fmt(m[["chisq.scaled"]], 1), as.integer(m[["df"]]), p),
                   fmt(m[["cfi.robust"]], 3), rmsea, fmt(m[["srmr"]], 3)),
    `Good fit` = c("p > .05", "≥ .95", "≤ .06", "≤ .08")
  )
}

# Model 0: the original simple chain -- parent education -> books -> reading
# (no technology). A three-variable recursive mediation model.
model0 <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"
fit0 <- lavaan(model0, data = pisa, auto.var = TRUE, auto.cov.y = FALSE, estimator = "MLR")
```



$$\text{Parent Ed} \;\rightarrow\; \text{Books at home} \;\rightarrow\; \text{Reading}$$

```{r}
#| echo: false
knitr::kable(fit_table(fit0))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Over-identified (df = `r as.integer(fitMeasures(fit0, "df"))`), so fit is
testable. Robust $\chi^2$ = `r fmt(fitMeasures(fit0, "chisq.scaled"), 1)`, $p$ =
`r fmt(fitMeasures(fit0, "pvalue.scaled"), 3)` (CFI =
`r fmt(fitMeasures(fit0, "cfi.robust"), 3)`, RMSEA =
`r fmt(fitMeasures(fit0, "rmsea.robust"), 3)`, SRMR =
`r fmt(fitMeasures(fit0, "srmr"), 3)`).
:::
:::


# Model Comparison

Rules: 
- Same variables
- Same sample
- Same estimator
- Nested structure: One model's parameter space must be a subset of the other's, achieved by

    - Fixing one or more parameters to a specific value (usually zero)
    - Constraining two or more parameters to be equal to each other

# Model Comparison Example

RQ: Does parent education relate to students' reading achievement only through the number of books in the home or does access to technology serve as an additional mediating pathway? 

1. Develop a model with only books as a mediator
2. Develop a model with both books and technology as a mediator
3. Compare the two models.
 
## Model 1

```{r}
#| echo: false
#| message: false
#| warning: false

# Model 1: reading mediated by books only. Technology is in the model
# (predicted by parent education) but has no path to reading. This keeps
# both models on the same four variables so they are nested and comparable.
# auto.cov.y = FALSE keeps the disturbances uncorrelated (recursive model).
model1 <- "
  BOOKSHOME ~ PAREDINT
  TECH      ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"
fit1 <- lavaan(model1, data = pisa, auto.var = TRUE, auto.cov.y = FALSE, estimator = "MLR")

# Model 2: adds the TECH -> Reading path (books AND tech mediate).
# Model 1 is Model 2 with that one path fixed to 0 -> nested.
model2 <- "
  BOOKSHOME ~ PAREDINT
  TECH      ~ PAREDINT
  READ_AVG  ~ BOOKSHOME + TECH
"
fit2 <- lavaan(model2, data = pisa, auto.var = TRUE, auto.cov.y = FALSE, estimator = "MLR")
```

```{r}
#| echo: false
#| fig-width: 7.5
#| fig-height: 5
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"
PUR <- "#7A3FA0"; TEA <- "#0E8F8F"; GLD <- "#C89A2B"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

box <- function(x, y, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = x, y = y, label = label, colour = "white", size = 4.6, lineheight = 0.9)
)

ggplot() +
  # a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.0, y = 0.45, xend = 3.85, yend = 1.35, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2.0, y = 1.2, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # a2: ParentEd -> Tech (purple)
  annotate("segment", x = 1.0, y = -0.45, xend = 3.85, yend = -1.35, colour = PUR, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2.0, y = -1.2, label = "a[2]", parse = TRUE, colour = PUR, size = 6, fontface = "italic") +
  # b1: Books -> Reading (blue)
  annotate("segment", x = 6.15, y = 1.35, xend = 9.0, yend = 0.45, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 8.0, y = 1.2, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.5, label = "Var", colour = dk, size = 4.4, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 5, xend = 5, y = 3.05, yend = 2.2, colour = MAG, arrow = ar) +
  annotate("text", x = 5, y = 3.35, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.4) +
  # disturbance zeta_T on Tech (gold)
  annotate("segment", x = 5, xend = 5, y = -3.05, yend = -2.2, colour = GLD, arrow = ar) +
  annotate("text", x = 5, y = -3.35, label = "zeta[T]", parse = TRUE, colour = GLD, size = 4.4) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 11.7, xend = 11.15, y = 0.9, yend = 0.35, colour = LIM, arrow = ar) +
  annotate("text", x = 12.1, y = 1.2, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.4) +
  # measured variables
  box(0, 0, "Parent\neducation") + box(5, 1.6, "Books at\nhome") +
  box(5, -1.6, "Technology") + box(10, 0, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 12.6), y = c(-3.7, 3.7)) +
  theme_void()
```

- What theoretical claims does this model make?
- How many degrees of freedom?
- How to describe its identification?

## How well does the model fit the data?

```{r}
#| echo: false
knitr::kable(fit_table(fit1))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Over-identified (df = `r as.integer(fitMeasures(fit1, "df"))`), so fit is
testable. Robust $\chi^2$ = `r fmt(fitMeasures(fit1, "chisq.scaled"), 1)`, $p$ =
`r fmt(fitMeasures(fit1, "pvalue.scaled"), 3)` (CFI =
`r fmt(fitMeasures(fit1, "cfi.robust"), 3)`, RMSEA =
`r fmt(fitMeasures(fit1, "rmsea.robust"), 3)`, SRMR =
`r fmt(fitMeasures(fit1, "srmr"), 3)`) -- reading is modeled through books
only; technology is in the model but not yet allowed to predict reading.
:::
:::

## Model 2

```{r}
#| echo: false
#| fig-width: 7.5
#| fig-height: 5
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"
PUR <- "#7A3FA0"; TEA <- "#0E8F8F"; GLD <- "#C89A2B"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

box <- function(x, y, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = x, y = y, label = label, colour = "white", size = 4.6, lineheight = 0.9)
)

ggplot() +
  # a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.0, y = 0.45, xend = 3.85, yend = 1.35, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2.0, y = 1.2, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # a2: ParentEd -> Tech (purple)
  annotate("segment", x = 1.0, y = -0.45, xend = 3.85, yend = -1.35, colour = PUR, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2.0, y = -1.2, label = "a[2]", parse = TRUE, colour = PUR, size = 6, fontface = "italic") +
  # b1: Books -> Reading (blue)
  annotate("segment", x = 6.15, y = 1.35, xend = 9.0, yend = 0.45, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 8.0, y = 1.2, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # b2: Tech -> Reading (teal)
  annotate("segment", x = 6.15, y = -1.35, xend = 9.0, yend = -0.45, colour = TEA, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 8.0, y = -1.2, label = "b[2]", parse = TRUE, colour = TEA, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.5, label = "Var", colour = dk, size = 4.4, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 5, xend = 5, y = 3.05, yend = 2.2, colour = MAG, arrow = ar) +
  annotate("text", x = 5, y = 3.35, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.4) +
  # disturbance zeta_T on Tech (gold)
  annotate("segment", x = 5, xend = 5, y = -3.05, yend = -2.2, colour = GLD, arrow = ar) +
  annotate("text", x = 5, y = -3.35, label = "zeta[T]", parse = TRUE, colour = GLD, size = 4.4) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 11.7, xend = 11.15, y = 0.9, yend = 0.35, colour = LIM, arrow = ar) +
  annotate("text", x = 12.1, y = 1.2, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.4) +
  # measured variables
  box(0, 0, "Parent\neducation") + box(5, 1.6, "Books at\nhome") +
  box(5, -1.6, "Technology") + box(10, 0, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 12.6), y = c(-3.7, 3.7)) +
  theme_void()
```

- What theoretical claims does this make?
- How many degrees of freedom?
- How to describe its identification?

## How well does model 2 fit the data?

```{r}
#| echo: false
knitr::kable(fit_table(fit2))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Over-identified (df = `r as.integer(fitMeasures(fit2, "df"))`). Adding the
technology -> reading path barely moves fit (CFI =
`r fmt(fitMeasures(fit2, "cfi.robust"), 3)`, RMSEA =
`r fmt(fitMeasures(fit2, "rmsea.robust"), 3)`), and that path itself is not
significant. The question is whether it improves on Model 1 -- next slide.
:::
:::

## Model comparison:

Does adding the technology -> reading path actually improve the model?

```{r}
#| echo: false
knitr::kable(anova(fit1, fit2))
```

```{r}
#| echo: false
knitr::kable(tibble::tibble(
  Model = c("M1: books-only", "M2: books + tech"),
  df    = c(fitMeasures(fit1, "df"),  fitMeasures(fit2, "df")),
  AIC   = c(fitMeasures(fit1, "aic"), fitMeasures(fit2, "aic")),
  BIC   = c(fitMeasures(fit1, "bic"), fitMeasures(fit2, "bic"))
))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
Nested models (M1 = M2 with the tech -> reading path fixed to 0). The
$\chi^2$ difference is `r fmt(anova(fit1, fit2)[2, "Chisq diff"], 2)` on 1 df,
$p$ = `r fmt(anova(fit1, fit2)[2, "Pr(>Chisq)"], 3)` -- **not** significant.
AIC and BIC are both **lower for Model 1**. Adding the technology pathway does
*not* improve the model: keep the simpler Model 1.
:::
:::


# R Lab: 

1. File X is the code used to run today's models. 

- make sure you can run it.

2. Continue to Specify your model in the Lavaan (or other) GUI.

- Start your own qmd or rmd file for your mini project
- use the code provided & the code from Lavaan GUI to start building your model
- Run at least a draft version of your model today.   

Use code from this analysis > change 


# Up Next: 

Interpreting the models

Mini Project Checkin: 

- You should have preliminary ideas about how to specify a model using your data
- You should be able to run at least a draft version of your own model with your own data
- You should understand if your draft model is a good fit.

Prep: Lavaan package exploration

- What arguments are relevant to your project/data?
- Use AI to help interpret tech writing into less technical language.




