---
title: "W06 Practice: Interpreting an Ordinal Path Model"
date: today
format:
  html:
    toc: false
    number-sections: false
    embed-resources: true
description: '<a href="lectures/w06_mvpa_ordinal_practice.pdf" download="w06_mvpa_ordinal_practice.pdf">Download PDF</a>'
---

## Purpose

Practice reporting on an ordinal MVPA. 

::: {.callout-note appearance="simple"}
**This is all fake data**, generated specifically to illustrate an ordinal
measured-variable path analysis.
:::

## The scenario

**Research question.** Among undergraduates in online courses, do **instructor
responsiveness** and **peer interaction** relate to students' **intention to
persist** in the course *through* their **sense of belonging**?

**The data.** A survey of 450 students. Every variable is a Likert item.

| Variable  | Meaning                                         | Scale                           |
|-----------|-------------------------------------------------|---------------------------------|
| `respond` | "My instructor responds to questions promptly"  | 1–4 (strongly disagree → agree) |
| `peer`    | "I interact regularly with classmates"          | 1–4 (strongly disagree → agree) |
| `belong`  | sense of belonging in the course                | 1–5 (not at all → completely)   |
| `persist` | intention to continue / re-enroll               | 1–4 (very unlikely → very likely) |

**The hypothesized model.**

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 3.2
#| fig-align: center
library(lavaan)
library(semPlot)
hyp <- "belong  ~ respond + peer
        persist ~ belong"
spm <- semPlotModel(lavaanify(hyp))
pos <- rbind(respond = c(0, 1), peer = c(0, -1),
             belong  = c(1, 0), persist = c(2, 0))
ord <- semPaths(spm, DoNotPlot = TRUE, whatLabels = "no", nCharNodes = 0)$graphAttributes$Nodes$labels
semPaths(spm, whatLabels = "no", layout = pos[ord, ], style = "ram",
         residuals = FALSE, intercepts = FALSE, exoCov = FALSE,
         nCharNodes = 0, sizeMan = 10, edge.color = "black")
```

## The data

```{r}
#| message: false
#| warning: false
library(lavaan)

set.seed(625)
n <- 450
respond_s <- rnorm(n)
peer_s    <- 0.30 * respond_s + rnorm(n, 0, sqrt(1 - 0.30^2))
belong_s  <- 0.45 * respond_s + 0.35 * peer_s +
             rnorm(n, 0, sqrt(1 - 0.45^2 - 0.35^2))
persist_s <- 0.50 * belong_s + rnorm(n, 0, sqrt(1 - 0.50^2))

to4 <- function(z) as.integer(cut(z, c(-Inf, -0.8, 0, 0.8, Inf), labels = 1:4))
to5 <- function(z) as.integer(cut(z, c(-Inf, -1, -0.2, 0.5, 1.1, Inf), labels = 1:5))

belong_survey <- data.frame(
  respond = to4(respond_s),
  peer    = to4(peer_s),
  belong  = to5(belong_s),
  persist = to4(persist_s)
)

head(belong_survey)
sapply(belong_survey, table)   # a few discrete categories each
```

## The fitted model

Notice some additional arguments to last week's Rlab:

```{r}
#| message: false
#| warning: false
model <- "
  belong  ~ a1*respond + a2*peer
  persist ~ b*belong

  ind_respond := a1*b
  ind_peer    := a2*b
"

fit <- lavaan(model, data = belong_survey,
              ordered = c("respond", "peer", "belong", "persist"),
              estimator = "DWLS",
              test = "scaled.shifted", # adds the scaled (robust) test statistic, so fit indices report both standard & scaled columns
              auto.var = TRUE,   # disturbance variances of the endogenous variables
              auto.cov.y = FALSE,# keep endogenous disturbances uncorrelated (recursive)
              auto.th = TRUE,    # thresholds of the ordered variables as free parameters
              auto.delta = TRUE, # scaling parameters (delta) for ordered variables
              se = "bootstrap", bootstrap = 200
              )
```

## The output (path coefficients)

```{r}
#| message: false
#| warning: false
summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```{r}
#| message: false
#| warning: false
# Standardized estimates with standard errors, z, p, and CIs
standardizedSolution(fit)
```

```{r}
#| message: false
#| warning: false
# Unstandardized estimates with percentile bootstrap CIs
parameterEstimates(fit, boot.ci.type = "perc", standardized = TRUE)
```

## Your task

1. What info is needed to answer the question?  (curate) 
2. Write up a results section. Bullet points is fine. Include key information.
3. Answer the question based on the results. 
