---
title: "Week 7: CFA Specification"
date: today
description: '<a href="lectures/w07_cfa_specification.pdf" download="w07_cfa_specification.pdf">Download PDF</a>'
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    self-contained-math: true
---

# Agenda

- CFA Intro
- CFA Path Diagrams & Sysems of Equations
- CFA Model Implied Covariance 
- Implications of CFA
- Specifying CFA in Lavaan
- Rlab

# Overview

Fundamentally the same: 

- Theory remains the motivation for all specification decisions
- Specify a theoretically motivated model
- Estimate parameters so that model-implied covariance matrix is as close to observed as possible

Key difference: 

- Relationship between measured variables (indicators) and latent variables 

# Review

Definition of latent variables

::: {.content-visible when-profile="teacher"}
::: {.fragment}

- not observed directly
- inferred from patterns among measured variables
- typically not easily captured by a single variable
- Ex: academic performance, intelligence, problem-solving ability.
:::
:::


# Reframe 

:::: {.columns}
::: {.column width="50%"}
**MVPA RQ:**
What is the effect of parent education on reading score through books in the home?
:::

::: {.column width="50%"}
**Latent Variable RQ:**
To what extent do parent education, books in the home, and technology in the home reflect a single latent socioeconomic status variable?
:::
::::

## System of equations

:::: {.columns}
::: {.column width="46%" style="font-size: 0.75em;"}
**MVPA**

$$
\begin{aligned}
\text{Books} &= a_0 + a_1\,(\text{ParentEd}) + \zeta_B \\
\text{Reading} &= b_0 + b_1\,(\text{Books}) + \zeta_R
\end{aligned}
$$
:::

::: {.column width="54%" style="font-size: 0.75em;"}
**Latent Variable**

$$
\begin{aligned}
\text{ParentEd} &= \tau_1 + \lambda_1\,\text{SES} + \varepsilon_1 \\
\text{BooksHome} &= \tau_2 + \lambda_2\,\text{SES} + \varepsilon_2 \\
\text{TechHome} &= \tau_3 + \lambda_3\,\text{SES} + \varepsilon_3
\end{aligned}
$$

Where: 

- τ = intercept 
- λ = path coefficient
- ε = residual

:::
::::

# Systems of equations to path diagrams

## Path Diagrams Review 

What do the following represent? 

- squares/rectangles
- Single-headed straight arrow
- curved double headed arrow

::: {.fragment}

- Circles/ovals: latent variables

:::


## Path diagram MVPA

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = -0.55, ymax = 0.55, fill = dk),
  annotate("text", x = x, y = 0, label = label, colour = "white", size = 5, lineheight = 0.9)
)

ggplot() +
  # path a1: ParentEd -> Books (orange)
  annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0, colour = oNG, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE, colour = oNG, size = 6, fontface = "italic") +
  # path b1: Books -> Reading (blue)
  annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0, colour = BLU, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE, colour = BLU, size = 6, fontface = "italic") +
  # variance of exogenous Parent education (self-loop)
  annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 1.85, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # disturbance zeta_B on Books (magenta)
  annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6, colour = MAG, arrow = ar) +
  annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE, colour = MAG, size = 4.6) +
  # disturbance zeta_R on Reading (lime)
  annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6, colour = LIM, arrow = ar) +
  annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE, colour = LIM, size = 4.6) +
  # measured variables
  box(0, "Parent\neducation") + box(4, "Books at\nhome") + box(8, "Reading\nscore") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
  theme_void()
```

## Path diagram Latent Variable

```{r}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, y, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = x, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

r_ses <- 1.1
bx <- 6
ys <- c(2.4, 0, -2.4)
cols <- c(oNG, BLU, LIM)
lam <- c("lambda[1]", "lambda[2]", "lambda[3]")
eps <- c("epsilon[1]", "epsilon[2]", "epsilon[3]")
th <- atan2(ys, bx - 1.15)
x0 <- r_ses * cos(th)
y0 <- r_ses * sin(th)
x1 <- rep(bx - 1.15, 3)

ggplot() +
  # latent factor (circle)
  geom_polygon(data = circle(0, 0, r_ses), aes(x, y), fill = dk) +
  annotate("text", x = 0, y = 0, label = "SES", colour = "white", size = 6) +
  # variance of the latent factor (self-loop)
  annotate("curve", x = -0.5, y = 1.0, xend = 0.5, yend = 1.0,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2) +
  annotate("text", x = 0, y = 2.25, label = "Var", colour = dk, size = 4.6, fontface = "italic") +
  # loadings: SES -> indicators
  annotate("segment", x = x0, xend = x1, y = y0, yend = ys,
           colour = cols, linewidth = 1.3, arrow = ar) +
  annotate("text", x = 0.55 * x1 + 0.45 * x0, y = 0.55 * ys + 0.45 * y0 + 0.5,
           label = lam, parse = TRUE, colour = cols, size = 6, fontface = "italic") +
  # residuals on each indicator (magenta)
  annotate("segment", x = bx + 3, xend = bx + 1.15, y = ys, yend = ys,
           colour = MAG, arrow = ar) +
  annotate("text", x = bx + 3.5, y = ys, label = eps, parse = TRUE, colour = MAG, size = 4.6) +
  # measured variables
  box(bx, ys[1], "Parent\neducation") +
  box(bx, ys[2], "Books at\nhome") +
  box(bx, ys[3], "Technology\nat home") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.5, 10.5), y = c(-3.3, 3)) +
  theme_void()
```

# Model-implied covariance matrix

- Use path tracing rules & the path diagram


## Review path tracing rules

Take a few minutes to review the path tracing rules from [Lecture 3](https://srdevore.github.io/edep_625_sem/lectures/w03_mvpa_specification.html#/title-slide). 

Then you'll decide if some paths are allowed or not allowed. 

## Allowed or not allowed? 

For each example, the **magenta** path is the route to evaluate. Is it an allowed route?

```{r}
#| include: false
library(ggplot2)

MAG <- "#C81E7A"; GRY <- "#9AA0A6"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

pc <- function(id, hl) ifelse(id %in% hl, MAG, GRY)
pw <- function(id, hl) ifelse(id %in% hl, 1.8, 1.0)

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
qbox <- function(x, y, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = x, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

# MVPA diagram; hl = ids to highlight: a1, b1, zB, zR, varP
mvpa_plot <- function(hl = character()) {
  ggplot() +
    annotate("segment", x = 1.15, xend = 2.85, y = 0, yend = 0,
             colour = pc("a1", hl), linewidth = pw("a1", hl), arrow = ar) +
    annotate("text", x = 2, y = 0.45, label = "a[1]", parse = TRUE,
             colour = pc("a1", hl), size = 6, fontface = "italic") +
    annotate("segment", x = 5.15, xend = 6.85, y = 0, yend = 0,
             colour = pc("b1", hl), linewidth = pw("b1", hl), arrow = ar) +
    annotate("text", x = 6, y = 0.45, label = "b[1]", parse = TRUE,
             colour = pc("b1", hl), size = 6, fontface = "italic") +
    annotate("curve", x = -0.55, y = 0.6, xend = 0.55, yend = 0.6,
             colour = pc("varP", hl), linewidth = pw("varP", hl), curvature = -1.4, arrow = ar2) +
    annotate("text", x = 0, y = 1.85, label = "Var", colour = pc("varP", hl),
             size = 4.6, fontface = "italic") +
    annotate("segment", x = 4, xend = 4, y = 1.85, yend = 0.6,
             colour = pc("zB", hl), linewidth = pw("zB", hl), arrow = ar) +
    annotate("text", x = 4, y = 2.25, label = "zeta[B]", parse = TRUE,
             colour = pc("zB", hl), size = 4.6) +
    annotate("segment", x = 8, xend = 8, y = 1.85, yend = 0.6,
             colour = pc("zR", hl), linewidth = pw("zR", hl), arrow = ar) +
    annotate("text", x = 8, y = 2.25, label = "zeta[R]", parse = TRUE,
             colour = pc("zR", hl), size = 4.6) +
    qbox(0, 0, "Parent\neducation") + qbox(4, 0, "Books at\nhome") + qbox(8, 0, "Reading\nscore") +
    coord_equal(clip = "off") +
    expand_limits(x = c(-1.3, 9.3), y = c(-0.9, 3)) +
    theme_void()
}

# Latent diagram; hl = ids to highlight: l1, l2, l3, e1, e2, e3, varS
latent_plot <- function(hl = character()) {
  r_ses <- 1.1
  bx <- 6
  ys <- c(2.4, 0, -2.4)
  ids_l <- c("l1", "l2", "l3")
  ids_e <- c("e1", "e2", "e3")
  th <- atan2(ys, bx - 1.15)
  x0 <- r_ses * cos(th)
  y0 <- r_ses * sin(th)
  x1 <- rep(bx - 1.15, 3)
  ggplot() +
    geom_polygon(data = circle(0, 0, r_ses), aes(x, y), fill = dk) +
    annotate("text", x = 0, y = 0, label = "SES", colour = "white", size = 6) +
    annotate("curve", x = -0.5, y = 1.0, xend = 0.5, yend = 1.0,
             colour = pc("varS", hl), linewidth = pw("varS", hl), curvature = -1.4, arrow = ar2) +
    annotate("text", x = 0, y = 2.25, label = "Var", colour = pc("varS", hl),
             size = 4.6, fontface = "italic") +
    annotate("segment", x = x0, xend = x1, y = y0, yend = ys,
             colour = pc(ids_l, hl), linewidth = pw(ids_l, hl), arrow = ar) +
    annotate("text", x = 0.55 * x1 + 0.45 * x0, y = 0.55 * ys + 0.45 * y0 + 0.5,
             label = c("lambda[1]", "lambda[2]", "lambda[3]"), parse = TRUE,
             colour = pc(ids_l, hl), size = 6, fontface = "italic") +
    annotate("segment", x = bx + 3, xend = bx + 1.15, y = ys, yend = ys,
             colour = pc(ids_e, hl), linewidth = pw(ids_e, hl), arrow = ar) +
    annotate("text", x = bx + 3.5, y = ys, label = c("epsilon[1]", "epsilon[2]", "epsilon[3]"),
             parse = TRUE, colour = pc(ids_e, hl), size = 4.6) +
    qbox(bx, ys[1], "Parent\neducation") +
    qbox(bx, ys[2], "Books at\nhome") +
    qbox(bx, ys[3], "Technology\nat home") +
    coord_equal(clip = "off") +
    expand_limits(x = c(-1.5, 10.5), y = c(-3.3, 3)) +
    theme_void()
}
```

## Allowed or not allowed? 

**Route:** Parent education → Books at home → Reading score

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 2.6
mvpa_plot(c("a1", "b1"))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Allowed.** Every arrow is followed in the direction of its arrowhead, and no variable is visited more than once.
:::
:::

## Allowed or not allowed? 

**Route:** Parent education ← SES → Books at home

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 5.6
#| fig-height: 3.2
latent_plot(c("l1", "l2", "varS"))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Allowed.** The route goes backward against λ₁, then forward with λ₂ — backward-then-forward is fine. Each variable appears once on the path.
:::
:::

## Allowed or not allowed?

**Route:** Parent education ← SES ↺ SES ↺ SES → Technology at home

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 5.6
#| fig-height: 3.2
latent_plot(c("l1", "l3", "varS"))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Not allowed.** The route passes through the curved double-headed arrow (Var of SES) twice. A route may use at most one curved arrow, and only at its turning point.
:::
:::

## Allowed or not allowed?

**Route:** Parent education → Books at home ← ζ<sub>B</sub>

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 2.6
mvpa_plot(c("a1", "zB"))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Not allowed.** The route goes forward along a₁ and then backward against the arrow from ζ<sub>B</sub>. Forward-then-backward is never allowed.
:::
:::

## Allowed or not allowed?

**Route:** Parent education ← SES → Books at home ← SES → Technology at home

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 5.6
#| fig-height: 3.2
latent_plot(c("l1", "l2", "l3"))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Not allowed.** The route passes through SES twice. It also goes forward into Books at home and then backward out of it.
:::
:::

## Allowed or not allowed?

**Route:** Books at home ← Parent education ↺ Parent education → Books at home

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 2.6
mvpa_plot(c("a1", "varP"))
```

::: {.content-visible when-profile="teacher"}
::: {.fragment}
**Allowed.** The route goes backward along a₁, crosses one curved arrow (Var of Parent education) exactly at its turning point, then forward along a₁. Parent education is passed only once.
:::
:::

## Review: Getting Model-implied cov

First identify the paths using tracing rules: 

1. A route may go backward against an arrowhead, then forward with an arrowhead — but never forward-then-backward. 
4. Never pass through the same variable twice in one route (no loops).
3. A route may pass through at most one curved double-headed (covariance) arrow, and only at the route's turning point.

## Review: Getting model-implied covariance

Then get the covariance using the paths

1. Find every valid route between the two variables (following the rules above).
2. For each route, multiply together all the path coefficients (and the one correlation/covariance term, if a curved arrow is involved).
3. Multiply the whole product by the variance of the route's ultimate common source variable.
4. Sum the contributions across all valid routes — that sum equals the total covariance between the two variables.


## Latent Variable Model-implied Covariance

The path diagram is simply a representation of the system of equations

$$
\begin{aligned}
\text{ParentEd} &= \tau_1 + \lambda_1\,\text{SES} + \varepsilon_1 \\
\text{BooksHome} &= \tau_2 + \lambda_2\,\text{SES} + \varepsilon_2 \\
\text{TechHome} &= \tau_3 + \lambda_3\,\text{SES} + \varepsilon_3
\end{aligned}
$$

*PE = Parent ed · BH = Books at home · TH = Technology at home · S = SES*

```{r}
#| include: false
library(ggplot2)

## one colour per matrix cell (same palette as the MVPA example)
lv_c11 <- "#128A7D"; lv_c12 <- "#E07B1A"; lv_c22 <- "#C81E7A"
lv_c13 <- "#2C6FB5"; lv_c23 <- "#7A3FA0"; lv_c33 <- "#6FA80F"
lv_BASE <- "#BFC3C7"; lv_DK <- "#3A3A3A"

lv_AR <- grid::arrow(length = grid::unit(0.18, "cm"), type = "closed")
lv_AR2 <- grid::arrow(length = grid::unit(0.14, "cm"), ends = "both", type = "closed")

## geometry (same as the latent path diagram)
lv_r <- 1.1
lv_bx <- 6
lv_ys <- c(2.4, 0, -2.4)
lv_th <- atan2(lv_ys, lv_bx - 1.15)
lv_x0 <- lv_r * cos(lv_th)
lv_y0 <- lv_r * sin(lv_th)

lv_circle <- function(x, y, r) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
# Var(SES) self-loop, above the SES circle
lv_varloop <- function(col, lty, lwd) annotate("curve", x = -0.5, y = 1.0, xend = 0.5, yend = 1.0,
                                               curvature = -1.4, colour = col, linewidth = lwd,
                                               linetype = lty, arrow = lv_AR2)
lv_nodebox <- function(y, label) list(
  annotate("rect", xmin = lv_bx - 1.15, xmax = lv_bx + 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = lv_DK),
  annotate("text", x = lv_bx, y = y, label = label, colour = "white", size = 4.4, lineheight = 0.9)
)
lv_base_model <- function(col = lv_BASE,
                          box_labels = c("Parent\neducation", "Books at\nhome", "Technology\nat home"),
                          latent = "SES", latent_size = 5, varlab = "Var(S)") list(
  geom_polygon(data = lv_circle(0, 0, lv_r), aes(x, y), fill = lv_DK),
  annotate("text", x = 0, y = 0, label = latent, colour = "white", size = latent_size),
  annotate("segment", x = lv_x0, xend = rep(lv_bx - 1.15, 3), y = lv_y0, yend = lv_ys,
           colour = col, linewidth = 1, arrow = lv_AR),
  annotate("text", x = 0.55 * (lv_bx - 1.15) + 0.45 * lv_x0, y = 0.55 * lv_ys + 0.45 * lv_y0 + 0.5,
           label = c("lambda[1]", "lambda[2]", "lambda[3]"), parse = TRUE, colour = col, size = 4.2),
  annotate("segment", x = lv_bx + 3, xend = lv_bx + 1.15, y = lv_ys, yend = lv_ys,
           colour = col, arrow = lv_AR),
  annotate("text", x = lv_bx + 3.5, y = lv_ys, label = c("epsilon[1]", "epsilon[2]", "epsilon[3]"),
           parse = TRUE, colour = col, size = 3.6),
  lv_varloop(col, "solid", 0.9),
  annotate("text", x = 0, y = 2.25, label = varlab, colour = col, size = 3.4),
  lv_nodebox(lv_ys[1], box_labels[1]),
  lv_nodebox(lv_ys[2], box_labels[2]),
  lv_nodebox(lv_ys[3], box_labels[3])
)
lv_trace_edge <- function(e, col) annotate("segment", x = e[1], y = e[2], xend = e[3], yend = e[4],
                                           colour = col, linewidth = 1.7, linetype = "dotted", arrow = lv_AR)
lv_L <- lapply(1:3, function(i) c(lv_x0[i], lv_y0[i], lv_bx - 1.15, lv_ys[i]))
lv_E <- lapply(1:3, function(i) c(lv_bx + 3, lv_ys[i], lv_bx + 1.15, lv_ys[i]))

lv_STEPS <- list(
  list(cells = list(c(1, 1)), col = lv_c11,
       expr = "atop(lambda[1]^2 %.% Var(S), +~Var(epsilon[1]))", edges = list(lv_L[[1]], lv_E[[1]])),
  list(cells = list(c(1, 2), c(2, 1)), col = lv_c12,
       expr = "lambda[1] %.% lambda[2] %.% Var(S)", edges = list(lv_L[[1]], lv_L[[2]])),
  list(cells = list(c(2, 2)), col = lv_c22,
       expr = "atop(lambda[2]^2 %.% Var(S), +~Var(epsilon[2]))", edges = list(lv_L[[2]], lv_E[[2]])),
  list(cells = list(c(1, 3), c(3, 1)), col = lv_c13,
       expr = "lambda[1] %.% lambda[3] %.% Var(S)", edges = list(lv_L[[1]], lv_L[[3]])),
  list(cells = list(c(2, 3), c(3, 2)), col = lv_c23,
       expr = "lambda[2] %.% lambda[3] %.% Var(S)", edges = list(lv_L[[2]], lv_L[[3]])),
  list(cells = list(c(3, 3)), col = lv_c33,
       expr = "atop(lambda[3]^2 %.% Var(S), +~Var(epsilon[3]))", edges = list(lv_L[[3]], lv_E[[3]]))
)

lv_diagram <- function(i, steps = lv_STEPS, ...) {
  g <- ggplot() + lv_base_model(if (i == 0) lv_DK else lv_BASE, ...)
  if (i >= 1) {
    st <- steps[[i]]
    for (e in st$edges) g <- g + lv_trace_edge(e, st$col)
    g <- g + lv_varloop(st$col, "dotted", 1.7)   # every cell passes through Var(SES) at the top
  }
  g + coord_equal(clip = "off") + theme_void() +
    expand_limits(x = c(-1.5, 10.5), y = c(-3.3, 3))
}

lv_covmat <- function(i, steps = lv_STEPS, labs = c("PE", "BH", "TH")) {
  g <- ggplot()
  for (r in 1:3) for (cc in 1:3) {
    g <- g + annotate("rect", xmin = cc - 0.48, xmax = cc + 0.48,
                      ymin = (4 - r) - 0.48, ymax = (4 - r) + 0.48,
                      fill = "grey95", colour = "grey80")
  }
  for (s in seq_len(i)) {
    st <- steps[[s]]; emph <- (s == i)
    for (cell in st$cells) {
      rr <- cell[1]; ccx <- cell[2]
      g <- g +
        annotate("rect", xmin = ccx - 0.48, xmax = ccx + 0.48,
                 ymin = (4 - rr) - 0.48, ymax = (4 - rr) + 0.48,
                 fill = st$col, alpha = if (emph) 0.92 else 0.32,
                 colour = if (emph) "black" else NA, linewidth = if (emph) 1 else 0) +
        annotate("text", x = ccx, y = (4 - rr), label = st$expr, parse = TRUE,
                 colour = if (emph) "white" else "grey25", size = 3.1, lineheight = 0.9)
    }
  }
  for (k in 1:3) {
    g <- g + annotate("text", x = k, y = 3.72, label = labs[k], fontface = "bold", size = 4.2) +
      annotate("text", x = 0.32, y = 4 - k, label = labs[k], fontface = "bold", size = 4.2)
  }
  g + coord_equal(clip = "off") + theme_void() +
    expand_limits(x = c(0.1, 3.6), y = c(0.4, 4.0))
}
```

## Trace 1 — Var(Parent ed)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
lv_diagram(1)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
lv_covmat(1)
```
:::
::::

**Var(PE):** trace PE to itself. Up-and-back through the common cause SES → λ₁²·Var(S); plus its own residual → Var(ε₁). Sum the routes.

## Trace 2 — Cov(Parent ed, Books)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
lv_diagram(2)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
lv_covmat(2)
```
:::
::::

**Cov(PE, BH):** one route PE ← SES → BH (backward, forward). Multiply the loadings λ₁·λ₂, then × the variance at the top, Var(S). The indicators share only SES.

## Trace 3 — Var(Books)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
lv_diagram(3)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
lv_covmat(3)
```
:::
::::

**Var(BH):** up-and-back through SES → λ₂²·Var(S); plus its own residual → Var(ε₂). Sum.

## Trace 4 — Cov(Parent ed, Technology)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
lv_diagram(4)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
lv_covmat(4)
```
:::
::::

**Cov(PE, TH):** one route PE ← SES → TH. Multiply λ₁·λ₃, then × Var(S).

## Trace 5 — Cov(Books, Technology)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
lv_diagram(5)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
lv_covmat(5)
```
:::
::::

**Cov(BH, TH):** one route BH ← SES → TH. Multiply λ₂·λ₃, then × Var(S).

## Trace 6 — Var(Technology)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
lv_diagram(6)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
lv_covmat(6)
```
:::
::::

**Var(TH):** up-and-back through SES → λ₃²·Var(S); plus its own residual → Var(ε₃). Sum.

## Practice

Motivation is measured by **Effort1**, **StudyHmW**, and **MissSc**.

```{r}
#| include: false
## Practice model: Motivation -> Effort1, StudyHmW, MissSc (reuses the lv_ helpers above)
pm_boxes <- c("Effort1", "StudyHmW", "MissSc")
pm_STEPS <- lapply(lv_STEPS, function(st) {
  st$expr <- gsub("Var(S)", "Var(Mot)", st$expr, fixed = TRUE)
  st
})
pm_diagram <- function(i) lv_diagram(i, steps = pm_STEPS, box_labels = pm_boxes,
                                     latent = "Motivation", latent_size = 3.8, varlab = "Var(Mot)")
pm_covmat <- function(i) lv_covmat(i, steps = pm_STEPS, labs = c("E", "H", "M"))
```

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
pm_diagram(0)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
pm_covmat(0)
```
:::
::::

::::: {.content-visible when-profile="teacher"}

## Practice Trace 1 — Var(Effort1)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
pm_diagram(1)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
pm_covmat(1)
```
:::
::::

**Var(E):** up-and-back through Motivation → λ₁²·Var(Mot); plus its own residual → Var(ε₁). Sum.

## Practice Trace 2 — Cov(Effort1, StudyHmW)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
pm_diagram(2)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
pm_covmat(2)
```
:::
::::

**Cov(E, H):** one route E ← Motivation → H. Multiply λ₁·λ₂, then × Var(Mot).

## Practice Trace 3 — Var(StudyHmW)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
pm_diagram(3)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
pm_covmat(3)
```
:::
::::

**Var(H):** up-and-back through Motivation → λ₂²·Var(Mot); plus its own residual → Var(ε₂). Sum.

## Practice Trace 4 — Cov(Effort1, MissSc)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
pm_diagram(4)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
pm_covmat(4)
```
:::
::::

**Cov(E, M):** one route E ← Motivation → M. Multiply λ₁·λ₃, then × Var(Mot).

## Practice Trace 5 — Cov(StudyHmW, MissSc)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
pm_diagram(5)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
pm_covmat(5)
```
:::
::::

**Cov(H, M):** one route H ← Motivation → M. Multiply λ₂·λ₃, then × Var(Mot).

## Practice Trace 6 — Var(MissSc)

:::: {.columns}
::: {.column width="56%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
pm_diagram(6)
```
:::
::: {.column width="44%"}
```{r}
#| echo: false
#| fig-width: 4.4
#| fig-height: 4.4
pm_covmat(6)
```
:::
::::

**Var(M):** up-and-back through Motivation → λ₃²·Var(Mot); plus its own residual → Var(ε₃). Sum.

:::::

## Practice: system of equations

Write the system of equations for the Motivation model.

$$
\begin{aligned}
\text{Effort1} &= \underline{\phantom{\lambda_1}} + \underline{\phantom{\lambda_1}}\,\underline{\phantom{\text{Motivation}}} + \underline{\phantom{\lambda_1}} \\
\text{StudyHmW} &= \underline{\phantom{\lambda_1}} + \underline{\phantom{\lambda_1}}\,\underline{\phantom{\text{Motivation}}} + \underline{\phantom{\lambda_1}} \\
\text{MissSc} &= \underline{\phantom{\lambda_1}} + \underline{\phantom{\lambda_1}}\,\underline{\phantom{\text{Motivation}}} + \underline{\phantom{\lambda_1}}
\end{aligned}
$$

::: {.content-visible when-profile="teacher"}
## Practice: system of equations (answers)

$$
\begin{aligned}
\text{Effort1} &= \tau_1 + \lambda_1\,\text{Motivation} + \varepsilon_1 \\
\text{StudyHmW} &= \tau_2 + \lambda_2\,\text{Motivation} + \varepsilon_2 \\
\text{MissSc} &= \tau_3 + \lambda_3\,\text{Motivation} + \varepsilon_3
\end{aligned}
$$
:::

# Implications

## Residuals & Random Measurement Error 

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 7.5
#| fig-height: 4.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, y, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = x, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

r_ses <- 1.1
bx <- 6
ys <- c(2.4, 0, -2.4)
cols <- c(oNG, BLU, LIM)
lam <- c("lambda[1]", "lambda[2]", "lambda[3]")
eps <- c("epsilon[1]", "epsilon[2]", "epsilon[3]")
th <- atan2(ys, bx - 1.15)
x0 <- r_ses * cos(th)
y0 <- r_ses * sin(th)
x1 <- rep(bx - 1.15, 3)

ggplot() +
  geom_polygon(data = circle(0, 0, r_ses), aes(x, y), fill = dk) +
  annotate("text", x = 0, y = 0, label = "SES", colour = "white", size = 6) +
  # shared variance: loadings, one colour each
  annotate("segment", x = x0, xend = x1, y = y0, yend = ys,
           colour = cols, linewidth = 1.8, arrow = ar) +
  annotate("text", x = 0.55 * x1 + 0.45 * x0, y = 0.55 * ys + 0.45 * y0 + 0.5,
           label = lam, parse = TRUE, colour = cols, size = 6, fontface = "italic") +
  # unique variance: residuals (magenta), drawn outside the loadings so nothing overlaps
  annotate("segment", x = bx + 3.2, xend = bx + 1.15, y = ys, yend = ys,
           colour = MAG, linewidth = 1.8, arrow = ar) +
  annotate("text", x = bx + 3.8, y = ys, label = eps, parse = TRUE, colour = MAG, size = 5) +
  annotate("text", x = 2.2, y = 4.1, label = "Shared variance",
           colour = dk, size = 4.6, fontface = "bold") +
  annotate("text", x = bx + 2.2, y = 4.1, label = "Unique variance (residuals)",
           colour = MAG, size = 4.6, fontface = "bold") +
  box(bx, ys[1], "Parent\neducation") +
  box(bx, ys[2], "Books at\nhome") +
  box(bx, ys[3], "Technology\nat home") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.5, 11), y = c(-3.3, 4.6)) +
  theme_void()
```


## Shared Variance & Residuals

- Shared variance: predicted or caused by the latent variable
- Residuals: 
  - Random measurement error: specific to the indicator & not shared
  - True variance caused by other things

✅ Pros: Random measurement error is isolated in the indicator, not in the latent variable.

🤔 Caution: Latent variables are still succeptible to error (ex: confounding factors, reporting error)


::: {.content-visible when-profile="teacher"}

only shared variance among the indicators is captured by the latent variable. Random measurement error is specific to its indicator (not across indicators nor factors) and so not captured in the shared variance. Therefore, the random measurement error is isolated in the measured variable residuals. Which is not to say that the latent variable is perfect. there can be other effects related to poor research design. Additionally, not all the residuals of the measured variables is necessarily random measurement error.
::: 

## Scale 

- What is 100 SES? 
- What is 50 Motivation?

## Scale

Latent variables have no natural scale, so we need to give them one. 

- Option 1: set the latent variable's variance to 1
  - each indicator is interpreted in terms of 1-SD change in the latent variable
- Option 2: set the first loading to 1
  - each indicator is interpreted in terms of the first variable

## Degrees of Freedom

## Information

Remember from MVPA: 

::: {.content-visible when-profile="teacher"}
::: {.fragment}
p(p+1)/2 

- Where p = the number of variables
- Number of cells in one triangle of the var-cov matrix
:::
:::

## Parameters (q) to be estimated {.smaller}

| Path analysis (measured) | CFA | CFA Matrix | CFA n (guide)
|---|---|---|---|
| Path coefficients (→ between measured vars | Path Coefficients* (factor → indicator) | Λ lambda p x k | often p |
| Variance/cov of exo var | **Factor variances/cov** | Ψ psi k x k | max of k(k+1)/2; k = n latent var |
| Disturbance variances of endo var | **Residual variances** of indicators | Θ theta p x p | p + 1 for residual cov|

*Traditionally called loadings but not in the sense of EFA/PCA or eigen decomposition. 

## Parameters to be estimated

q = sum of free parameters in Λ, Ψ, Θ

- Don't count fixed paths/variances (i.e. when scaling)

## Example

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 7.5
#| fig-height: 3.8
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; LIM <- "#6FA80F"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.22, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

circle <- function(x, y, r = 0.5) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}
box <- function(x, y, label) list(
  annotate("rect", xmin = x - 1.15, xmax = x + 1.15, ymin = y - 0.55, ymax = y + 0.55, fill = dk),
  annotate("text", x = x, y = y, label = label, colour = "white", size = 5, lineheight = 0.9)
)

r_ses <- 1.1
bx <- 6
ys <- c(2.4, 0, -2.4)
cols <- c(oNG, BLU, LIM)
lam <- c("lambda[1]", "lambda[2]", "lambda[3]")
eps <- c("epsilon[1]", "epsilon[2]", "epsilon[3]")
th <- atan2(ys, bx - 1.15)
x0 <- r_ses * cos(th)
y0 <- r_ses * sin(th)
x1 <- rep(bx - 1.15, 3)

ggplot() +
  geom_polygon(data = circle(0, 0, r_ses), aes(x, y), fill = dk) +
  annotate("text", x = 0, y = 0, label = "SES", colour = "white", size = 6) +
  # variance of the latent factor (self-loop)
  annotate("curve", x = -0.5, y = 1.0, xend = 0.5, yend = 1.0,
           colour = dk, linewidth = 0.9, curvature = -1.4, arrow = ar2, linetype = "dashed") +
  annotate("text", x = 0, y = 2.25, label = "Var = 1", colour = dk, size = 4.6, fontface = "italic") +
  # loadings, one colour each
  annotate("segment", x = x0, xend = x1, y = y0, yend = ys,
           colour = cols, linewidth = 1.8, arrow = ar) +
  annotate("text", x = 0.55 * x1 + 0.45 * x0, y = 0.55 * ys + 0.45 * y0 + 0.5,
           label = lam, parse = TRUE, colour = cols, size = 6, fontface = "italic") +
  # unique variance: residuals (magenta), drawn outside the loadings so nothing overlaps
  annotate("segment", x = bx + 3.2, xend = bx + 1.15, y = ys, yend = ys,
           colour = MAG, linewidth = 1.8, arrow = ar) +
  annotate("text", x = bx + 3.8, y = ys, label = eps, parse = TRUE, colour = MAG, size = 5) +
  box(bx, ys[1], "Parent\neducation") +
  box(bx, ys[2], "Books at\nhome") +
  box(bx, ys[3], "Technology\nat home") +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1.5, 11), y = c(-3.3, 3)) +
  theme_void()
```

Information: 3(3+1) / 2 = 6 pieces of information

Parameters to estimate = 3 + 3 = 6

- path estimates: $\lambda_1, \lambda_2, \lambda_3$
- Residual variances: $\theta_{11}, \theta_{22}, \theta_{33}$
- Factor variance $\psi_{11}$ is fixed to 1 to scale the factor (not estimated)

## Practice

Calculate degrees of freedom for the SES model

## CFA Example 2 latent variables

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.6
#| fig-align: center
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; MAG <- "#C81E7A"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.2, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

ex_circle <- function(x, y, r) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}

ex_names <- c("PARENTED", "BOOKSHOME", "TECHHOME", "IMMIG", "WORKPAY", "WORKHOME",
              "EFFORT1", "STUDYHMW", "MISSSC")
ex_fac <- c(1, 1, 1, 1, 1, 1, 2, 2, 2)            # which factor each indicator loads on
ex_x <- 1.8 * (seq_along(ex_names) - 1)           # indicator positions
ex_fx <- c(mean(ex_x[1:6]), mean(ex_x[7:9]))      # factor positions
ex_fy <- 4.2
ex_r <- 1.0
ex_col <- c(oNG, BLU)[ex_fac]

# loadings: from the factor circle edge to the top of each indicator box
ex_dx <- ex_x - ex_fx[ex_fac]
ex_dy <- 0.55 - ex_fy
ex_len <- sqrt(ex_dx^2 + ex_dy^2)
ex_x0 <- ex_fx[ex_fac] + ex_r * ex_dx / ex_len
ex_y0 <- ex_fy + ex_r * ex_dy / ex_len

ggplot() +
  # factors
  geom_polygon(data = ex_circle(ex_fx[1], ex_fy, ex_r), aes(x, y), fill = dk) +
  geom_polygon(data = ex_circle(ex_fx[2], ex_fy, ex_r), aes(x, y), fill = dk) +
  annotate("text", x = ex_fx[1], y = ex_fy, label = "SES", colour = "white", size = 5) +
  annotate("text", x = ex_fx[2], y = ex_fy, label = "Student\nmotivation", colour = "white",
           size = 2.9, lineheight = 0.9) +
  # factor variances (self-loops)
  annotate("curve", x = ex_fx[1] - 0.45, y = ex_fy + 0.9, xend = ex_fx[1] + 0.45, yend = ex_fy + 0.9,
           colour = dk, linewidth = 0.8, curvature = -1.4, arrow = ar2) +
  annotate("curve", x = ex_fx[2] - 0.45, y = ex_fy + 0.9, xend = ex_fx[2] + 0.45, yend = ex_fy + 0.9,
           colour = dk, linewidth = 0.8, curvature = -1.4, arrow = ar2) +
  # factor covariance (curved double-headed arrow)
  annotate("curve", x = ex_fx[1] + ex_r + 0.1, y = ex_fy, xend = ex_fx[2] - ex_r - 0.1, yend = ex_fy,
           colour = dk, linewidth = 1, curvature = 0.25, arrow = ar2) +
  annotate("text", x = mean(ex_fx), y = ex_fy - 1.7, label = "Cov", colour = dk,
           size = 4.2, fontface = "italic") +
  # loadings
  annotate("segment", x = ex_x0, xend = ex_x, y = ex_y0, yend = 0.55,
           colour = ex_col, linewidth = 1, arrow = ar) +
  annotate("text", x = 0.45 * ex_x0 + 0.55 * ex_x + 0.35, y = 0.45 * ex_y0 + 0.55 * 0.55 + 0.1,
           label = paste0("lambda[", 1:9, "]"), parse = TRUE, colour = ex_col, size = 3.6) +
  # residuals
  annotate("segment", x = ex_x, xend = ex_x, y = -1.9, yend = -0.55, colour = MAG, arrow = ar) +
  annotate("text", x = ex_x, y = -2.3, label = paste0("epsilon[", 1:9, "]"), parse = TRUE,
           colour = MAG, size = 3.8) +
  # indicators
  annotate("rect", xmin = ex_x - 0.8, xmax = ex_x + 0.8, ymin = -0.55, ymax = 0.55, fill = dk) +
  annotate("text", x = ex_x, y = 0, label = ex_names, colour = "white", size = 2.9) +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1, 15.4), y = c(-2.7, 6)) +
  theme_void()
```

<!--
SES > PARENTED, BOOKSHOME, TECHHOME, IMMIG, WORKPAY, WORKHOME
STUDENT MOTIVATION: EFFORT1, STUDYHMW, MISSSC,
-->

## Answers

Information: 9(9+1) / 2 = 45 pieces of information

Parameters to estimate (factor variances fixed to 1):

- path estimates: $\lambda_1, \dots, \lambda_9$ = 9
- Residual variances: $\theta_{11}, \dots, \theta_{99}$ = 9
- Factor covariance (a correlation, since variances are 1): $\psi_{21}$ = 1
- Total: 9 + 9 + 1 = 19

Degrees of freedom: 45 - 19 = 26

## CFA Example 3 latent variables

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
#| fig-align: center
library(ggplot2)

oNG <- "#E07B1A"; BLU <- "#2C6FB5"; LIM <- "#6FA80F"; MAG <- "#C81E7A"; dk <- "#3A3A3A"
ar <- grid::arrow(length = grid::unit(0.2, "cm"), type = "closed")
ar2 <- grid::arrow(length = grid::unit(0.18, "cm"), ends = "both", type = "closed")

ex3_circle <- function(x, y, r) {
  t <- seq(0, 2 * pi, length.out = 60)
  data.frame(x = x + r * cos(t), y = y + r * sin(t))
}

ex3_names <- c("PARENTED", "BOOKSHOME", "TECHHOME", "IMMIG", "WORKPAY", "WORKHOME",
               "EFFORT1", "STUDYHMW", "MISSSC",
               "READING_AVE", "MATH_AVE", "SCIENCE_AVE")
ex3_fac <- c(rep(1, 6), rep(2, 3), rep(3, 3))      # which factor each indicator loads on
ex3_x <- 1.8 * (seq_along(ex3_names) - 1)          # indicator positions
ex3_fx <- c(mean(ex3_x[1:6]), mean(ex3_x[7:9]), mean(ex3_x[10:12]))
ex3_fy <- 4.2
ex3_r <- 1.0
ex3_cols <- c(oNG, BLU, LIM)
ex3_col <- ex3_cols[ex3_fac]
ex3_fcol <- c(dk, dk, dk)

# loadings: from the factor circle edge to the top of each indicator box
ex3_dx <- ex3_x - ex3_fx[ex3_fac]
ex3_dy <- 0.55 - ex3_fy
ex3_len <- sqrt(ex3_dx^2 + ex3_dy^2)
ex3_x0 <- ex3_fx[ex3_fac] + ex3_r * ex3_dx / ex3_len
ex3_y0 <- ex3_fy + ex3_r * ex3_dy / ex3_len

ggplot() +
  # factors
  geom_polygon(data = ex3_circle(ex3_fx[1], ex3_fy, ex3_r), aes(x, y), fill = dk) +
  geom_polygon(data = ex3_circle(ex3_fx[2], ex3_fy, ex3_r), aes(x, y), fill = dk) +
  geom_polygon(data = ex3_circle(ex3_fx[3], ex3_fy, ex3_r), aes(x, y), fill = dk) +
  annotate("text", x = ex3_fx[1], y = ex3_fy, label = "SES", colour = "white", size = 5) +
  annotate("text", x = ex3_fx[2], y = ex3_fy, label = "Student\nmotivation", colour = "white",
           size = 2.9, lineheight = 0.9) +
  annotate("text", x = ex3_fx[3], y = ex3_fy, label = "Student\noutcomes", colour = "white",
           size = 2.9, lineheight = 0.9) +
  # factor variances (self-loops)
  annotate("curve", x = ex3_fx - 0.45, y = ex3_fy + 0.9, xend = ex3_fx + 0.45, yend = ex3_fy + 0.9,
           colour = dk, linewidth = 0.8, curvature = -1.4, arrow = ar2) +
  # factor covariances (curved double-headed arrows)
  annotate("curve", x = ex3_fx[1] + ex3_r + 0.1, y = ex3_fy, xend = ex3_fx[2] - ex3_r - 0.1, yend = ex3_fy,
           colour = dk, linewidth = 1, curvature = 0.25, arrow = ar2) +
  annotate("curve", x = ex3_fx[2] + ex3_r + 0.1, y = ex3_fy, xend = ex3_fx[3] - ex3_r - 0.1, yend = ex3_fy,
           colour = dk, linewidth = 1, curvature = 0.25, arrow = ar2) +
  annotate("curve", x = ex3_fx[1] + 0.8, y = ex3_fy + 0.65, xend = ex3_fx[3] - 0.8, yend = ex3_fy + 0.65,
           colour = dk, linewidth = 1, curvature = -0.3, arrow = ar2) +
  annotate("text", x = c(mean(ex3_fx[1:2]), mean(ex3_fx[2:3])), y = c(ex3_fy - 1.25, ex3_fy - 0.95), label = "Cov",
           colour = dk, size = 4.2, fontface = "italic") +
  annotate("text", x = ex3_fx[2], y = ex3_fy + 2.85, label = "Cov", colour = dk,
           size = 4.2, fontface = "italic") +
  # loadings
  annotate("segment", x = ex3_x0, xend = ex3_x, y = ex3_y0, yend = 0.55,
           colour = ex3_col, linewidth = 1, arrow = ar) +
  annotate("text", x = 0.45 * ex3_x0 + 0.55 * ex3_x + 0.35, y = 0.45 * ex3_y0 + 0.55 * 0.55 + 0.1,
           label = paste0("lambda[", 1:12, "]"), parse = TRUE, colour = ex3_col, size = 3.4) +
  # residuals
  annotate("segment", x = ex3_x, xend = ex3_x, y = -1.9, yend = -0.55, colour = MAG, arrow = ar) +
  annotate("text", x = ex3_x, y = -2.3, label = paste0("epsilon[", 1:12, "]"), parse = TRUE,
           colour = MAG, size = 3.6) +
  # indicators
  annotate("rect", xmin = ex3_x - 0.85, xmax = ex3_x + 0.85, ymin = -0.55, ymax = 0.55, fill = dk) +
  annotate("text", x = ex3_x, y = 0, label = ex3_names, colour = "white", size = 2.2) +
  coord_equal(clip = "off") +
  expand_limits(x = c(-1, 20.8), y = c(-2.7, 7.4)) +
  theme_void()
```

<!--
SES > PARENTED, BOOKSHOME, TECHHOME, IMMIG, WORKPAY, WORKHOME
STUDENT MOTIVATION: EFFORT1, STUDYHMW, MISSSC,
STUDENT OUTCOMES: READING_AVE, MATH_AVE, SCIENCE_AVE
-->

## Answers

Information: 12(12+1) / 2 = 78 pieces of information

Parameters to estimate (factor variances fixed to 1):

- path estimates: $\lambda_1, \dots, \lambda_{12}$ = 12
- Residual variances: $\theta_{11}, \dots, \theta_{12,12}$ = 12
- Factor covariances (correlations): $\psi_{21}, \psi_{31}, \psi_{32}$ = 3
- Total: 12 + 12 + 3 = 27

Degrees of freedom: 78 - 27 = 51

# Specifying CFA models in Lavaan

## Lavaan model syntax MVPA {.smaller}

| Symbol | Name | Meaning | Example |
|:------:|:-----|:--------|:--------|
| `~` | regression | outcome ~ predictors | `READ ~ BOOKSHOME` |
| `:=` | define parameter | new parameter computed from labelled parameters (e.g., indirect effect) | `ind := a*b` |


## Lavaan model syntax: symbols {.smaller}

| Symbol | Name | Meaning | Example |
|:------:|:-----|:--------|:--------|
| `=~` | measured by | latent variable → indicators (path coefficients, $\lambda$) | `SES =~ PARENTED + BOOKSHOME` |
| `~~` | covariance | variance, residual variance, or covariance | `SES ~~ MOTIV` |
| `*` |  fix | fix a parameter value | `SES =~ 1*PARENTED`, `SES ~~ 1*SES` and `SES =~ NA*PARENTED`|
| `+` | plus | separates indicators or predictors | `a + b + c` |
| `#` | comment | ignored / doesn't run | `# SES factor`,  |

: {tbl-colwidths="[10,15,40,35]"}

## Lavaan model syntax: 1st path fixed to 1

```r
cfa_model <- '
  # measurement model (first loading fixed to 1 to set the scale)
  SES   =~ 1*PARENTED + BOOKSHOME + TECHHOME
  MOTIV =~ 1*EFFORT1  + STUDYHMW  + MISSSC

  # factor variances / covariance
  SES   ~~ SES
  MOTIV ~~ MOTIV
  SES   ~~ MOTIV

  # residual variances
  PARENTED ~~ PARENTED
  BOOKSHOME ~~ BOOKSHOME
  TECHHOME ~~ TECHHOME
  EFFORT1 ~~ EFFORT1
  STUDYHMW ~~ STUDYHMW
  MISSSC ~~ MISSSC
'

fit <- lavaan(cfa_model, data = dat)
summary(fit, fit.measures = TRUE)
```

- Anything not specified is fixed to 0 (e.g., no `SES ~~ MOTIV` line means the factors are uncorrelated)

## Lavaan model syntax: factor var fixed

```r
cfa_model2 <- '
  # measurement model (all loadings free; NA* frees the first one)
  SES   =~ NA*PARENTED + BOOKSHOME + TECHHOME
  MOTIV =~ NA*EFFORT1  + STUDYHMW  + MISSSC

  # factor variances fixed to 1 to set the scale
  SES   ~~ 1*SES
  MOTIV ~~ 1*MOTIV

  # factor covariance (now a correlation)
  SES   ~~ MOTIV

  # residual variances
  PARENTED ~~ PARENTED
  BOOKSHOME ~~ BOOKSHOME
  TECHHOME ~~ TECHHOME
  EFFORT1 ~~ EFFORT1
  STUDYHMW ~~ STUDYHMW
  MISSSC ~~ MISSSC
'

fit2 <- lavaan(cfa_model2, data = dat)
summary(fit2, fit.measures = TRUE)
```

- Scale is set by fixing the factor variance (`1*`), so every loading is free
- `SES ~~ MOTIV` is now a correlation because both variances are 1
- Same model fit and df as the marker-indicator version; only the scale of the estimates differs

## Path diagram from the model syntax

```{r}
#| echo: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 5
library(lavaan)
library(semPlot)

dat <- read.csv("../data/w07_data_pisa22_USA_cln_cfa.csv")

cfa_model <- '
  SES   =~ 1*PARENTED + BOOKSHOME + TECHHOME
  MOTIV =~ 1*EFFORT1  + STUDYHMW  + MISSSC
  SES   ~~ SES
  MOTIV ~~ MOTIV
  SES   ~~ MOTIV
  PARENTED ~~ PARENTED
  BOOKSHOME ~~ BOOKSHOME
  TECHHOME ~~ TECHHOME
  EFFORT1 ~~ EFFORT1
  STUDYHMW ~~ STUDYHMW
  MISSSC ~~ MISSSC
'

# parse the model and data, but do not estimate
fit0 <- lavaan(cfa_model, data = dat, do.fit = FALSE)

semPaths(fit0, what = "path", whatLabels = "name", style = "lisrel",
         layout = "tree", residuals = TRUE,
         nCharNodes = 0, sizeMan = 12, sizeMan2 = 5)
```

# Rlab

Go to the rlab for today. 

# Up Next: 

Reading on CFA Models 
