---
title: "Lab: Familiarizing with PISA Data"
draft: true
date: today
description: '<a href="rlab-downloads/w01_rlab_pisa_intro.qmd" download="w01_rlab_pisa_intro.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w01_data_pisa22_USA_cln.csv" download="w01_data_pisa22_USA_cln.csv">w01_data_pisa22_USA_cln.csv</a>'
format: html
---

## Purpose

Familiarize with Pisa 2022 data subset that we'll be working with over the next weeks by doing exploratory data analysis. 

You have taken some previous stats courses. What are some things that stand out to you as important from this exploration? Make some notes about things we should keep in mind as we work with the data. 

## About PISA



This file is a cleaned subset that includes three variables for U.S. students.

| Variable | What it is |
|---|---|
| `PAREDINT` | Highest parental education in years of schooling; takes the values 3, 6, 9, 12, 14.5, 16 |
| `READ_AVG` | Effectively the reading literacy score (technically the mean of the plausible values)|
| `BOOKSHOME` | Books in the home, an ordered 1–5 Likert-style response |


3: Completed ISCED 1 (Primary school education, usually equivalent to finishing roughly 3rd grade).6: Completed an intermediate level of primary education (typically finishing roughly 6th grade / elementary school).9: Completed ISCED 2 (Lower secondary education / Middle school, finishing roughly 9th grade).12: Completed ISCED 3A (Upper secondary education / High School Graduate with a standard diploma).14.5: Completed ISCED 4 or 5B (Non-university tertiary education / Associate Degree or vocational/technical credentials). The decimal accounts for a standardized international average weight assigned to these shorter degree tracks.16: Completed ISCED 5A or 6 (Tertiary higher education / Bachelor’s Degree or higher, e.g., Master's, Ph.D.).

## Setup

```{r}
#| message: false
library(tidyverse)

pisa <- read_csv("w01_data_pisa22_USA_cln.csv")
```

## Structure and data types

```{r}
glimpse(pisa)
```

```{r}
dim(pisa)
names(pisa)
```

```{r}
sapply(pisa, class)
```

## Summary statistics

```{r}
summary(pisa)
```

```{r}
pisa |>
  summarise(
    across(
      everything(),
      list(
        n_valid = ~ sum(!is.na(.x)),
        n_miss  = ~ sum(is.na(.x)),
        mean    = ~ mean(.x, na.rm = TRUE),
        sd      = ~ sd(.x, na.rm = TRUE),
        min     = ~ min(.x, na.rm = TRUE),
        max     = ~ max(.x, na.rm = TRUE)
      )
    )
  ) |>
  pivot_longer(everything(), names_to = c("variable", "stat"), names_pattern = "^([^_]+)_(.+)$") |>
  pivot_wider(names_from = stat, values_from = value)
```

## Missing data

```{r}
colSums(is.na(pisa))
```

```{r}
mean(complete.cases(pisa))
```

## How many distinct values does each variable take?

```{r}
sapply(pisa, function(x) length(unique(x)))
```

```{r}
count(pisa, PAREDINT)
count(pisa, BOOKSHOME)
```

## Distributions

```{r}
#| warning: false
ggplot(pisa, aes(x = READ_AVG)) +
  geom_histogram(bins = 40) +
  labs(x = "Reading (PV1)", y = "Students")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = factor(PAREDINT))) +
  geom_bar() +
  labs(x = "Parental education (years)", y = "Students")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = factor(BOOKSHOME))) +
  geom_bar() +
  labs(x = "Books at home (category)", y = "Students")
```

## Relationships among the variables

```{r}
cor(pisa, use = "pairwise.complete.obs")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.4, alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Parental education (years)", y = "Reading (PV1)")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = factor(BOOKSHOME), y = READ_AVG)) +
  geom_boxplot() +
  labs(x = "Books at home (category)", y = "Reading (PV1)")
```

```{r}
pisa |>
  filter(!is.na(BOOKSHOME)) |>
  group_by(BOOKSHOME) |>
  summarise(n = n(), mean_read = mean(READ_AVG), sd_read = sd(READ_AVG))
```

## Your observations

Having run the code above, answer the following in a few sentences each. Point to specific output when you can.

# Have some fun :)

If you have time, explore more with additional visualizations, etc. 

```{r}
# scratch space for anything else you want to check
```
