---
title: "W01: Familiarizing with PISA Data"
date: today
description: '<a href="rlab-downloads/w01_rlab_pisa_intro.qmd" download="w01_rlab_pisa_intro.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w01_data_pisa22_USA_cln.csv" download="w01_data_pisa22_USA_cln.csv">w01_data_pisa22_USA_cln.csv</a>'
format: html
---

## Purpose

Familiarize with Pisa 2022 data subset that we'll be working with over the next weeks by doing exploratory data analysis. 

You have taken some previous stats courses. What are some things that stand out to you as important from this exploration? Make some notes about things we should keep in mind as we work with the data. 

## About PISA



This file is a cleaned subset that includes three variables for U.S. students.

| Variable | What it is |
|---|---|
| `PAREDINT` | Highest parental education in years of schooling; takes the values 3, 6, 9, 12, 14.5, 16 |
| `READ_AVG` | Effectively the reading literacy score (technically the mean of the plausible values)|
| `BOOKSHOME` | Books in the home, an ordered 1–5 Likert-style response |

See [Pisa Codebook](https://www.oecd.org/en/data/datasets/pisa-2022-database.html#codebook) for details on what each variable 

## Setup

```{r}
#| message: false
library(tidyverse)

pisa <- read_csv("w01_data_pisa22_USA_cln.csv")
```

## Structure and data types

```{r}
glimpse(pisa)
```

```{r}
dim(pisa)
names(pisa)
```

```{r}
sapply(pisa, class)
```

## Summary statistics

```{r}
summary(pisa)
```

```{r}
pisa |>
  summarise(
    across(
      everything(),
      list(
        n_valid = ~ sum(!is.na(.x)),
        n_miss  = ~ sum(is.na(.x)),
        mean    = ~ mean(.x, na.rm = TRUE),
        sd      = ~ sd(.x, na.rm = TRUE),
        min     = ~ min(.x, na.rm = TRUE),
        max     = ~ max(.x, na.rm = TRUE)
      )
    )
  ) |>
  pivot_longer(everything(), names_to = c("variable", "stat"), names_pattern = "^([^_]+)_(.+)$") |>
  pivot_wider(names_from = stat, values_from = value)
```

## Missing data

```{r}
colSums(is.na(pisa))
```

```{r}
mean(complete.cases(pisa))
```

## How many distinct values does each variable take?

```{r}
sapply(pisa, function(x) length(unique(x)))
```

```{r}
count(pisa, PAREDINT)
count(pisa, BOOKSHOME)
```

## Distributions

```{r}
#| warning: false
ggplot(pisa, aes(x = READ_AVG)) +
  geom_histogram(bins = 40) +
  labs(x = "Reading (PV1)", y = "Students")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = factor(PAREDINT))) +
  geom_bar() +
  labs(x = "Parental education (years)", y = "Students")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = factor(BOOKSHOME))) +
  geom_bar() +
  labs(x = "Books at home (category)", y = "Students")
```

## Relationships among the variables

```{r}
cor(pisa, use = "pairwise.complete.obs")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = PAREDINT, y = READ_AVG)) +
  geom_jitter(width = 0.4, alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Parental education (years)", y = "Reading (PV1)")
```

```{r}
#| warning: false
ggplot(pisa, aes(x = factor(BOOKSHOME), y = READ_AVG)) +
  geom_boxplot() +
  labs(x = "Books at home (category)", y = "Reading (PV1)")
```

```{r}
pisa |>
  filter(!is.na(BOOKSHOME)) |>
  group_by(BOOKSHOME) |>
  summarise(n = n(), mean_read = mean(READ_AVG), sd_read = sd(READ_AVG))
```

## Your observations

Having run the code above, answer the following in a few sentences each. Point to specific output when you can.

# Have some fun :)

If you have time, explore more with additional visualizations, etc. 

```{r}
# scratch space for anything else you want to check
```
