---
title: "Lab: Regression Review"
date: today
description: '<a href="rlab-downloads/w02_rlab_review_SEM_FUNdamentals.qmd" download="w02_rlab_review_SEM_FUNdamentals.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w01_data_pisa22_USA_cln.csv" download="w01_data_pisa22_USA_cln.csv">w01_data_pisa22_USA_cln.csv</a>'
format: html
release: week 2
draft: true
---

## Purpose

Goal: Run & Interpret the regression models below. 

Purpose: 

* Refresher on foundational concepts in regression
* Compare/contrast these results to the results of an SEM model of the same data in future weeks. 

## Setup

```{r}
#| message: false
library(tidyverse)
library(lavaan)
library(lavaanPlot)

pisa <- read_csv("w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)
```


```{r}
pared_mean  <- mean(pisa$PAREDINT, na.rm = TRUE)
books_mean  <- mean(pisa$BOOKSHOME, na.rm = TRUE)

pisa <- pisa |>
  mutate(
    PAREDINT_c  = PAREDINT - pared_mean,
    BOOKSHOME_c = BOOKSHOME - books_mean
  )
```

```{r}
#| include: false
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

## Model 1: Simple regression

$$\widehat{\text{Reading}} = b_0 + b_1(\text{ParentEd})$$

```{r}
m1 <- lm(READ_AVG ~ PAREDINT_c, data = pisa)
summary(m1)
```

**Questions**

1. In your own words, what is this model doing?
2. $R^2$ indicates ________________________________________. What is its value here, and what does that value tell you about how well parent education predicts reading score?
3. The estimate for `PAREDINT_c` indicates ________________________________________. Is it statistically significant? How do you know?
4. What does the intercept represent here, now that `PAREDINT_c` is mean-centered?

::: {.content-visible when-profile="teacher"}
**Sample answer.** This model predicts reading score from a single variable,
parent education. $R^2$ indicates the proportion of variance in reading score
that is explained by (accounted for by) the model -- here
`r fmt(summary(m1)$r.squared, 3)`, so parent education alone explains only
about `r fmt(summary(m1)$r.squared * 100, 0)`% of the variation in reading
scores. The estimate for `PAREDINT_c`
(`r fmt(coef(m1)[["PAREDINT_c"]], 2)`) indicates that each additional year of
parental education is associated with about
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` more reading-score points, on average. It
is statistically significant ($p < .001$): the relationship is very unlikely
to be zero in the population. Because `PAREDINT_c` is centered at its mean
(`r fmt(pared_mean, 1)` years), the intercept
(`r fmt(coef(m1)[["(Intercept)"]], 1)`) is the predicted reading score for a
student at *average* parental education, not for a (nonexistent) student with
0 years of schooling. The slope is unchanged by centering -- only the
intercept's meaning changes.
:::

## Model 1.a Simple regression 2

```{r}
m1a <- lm(READ_AVG ~ BOOKSHOME_c, data = pisa)
summary(m1a)
```

## Model 2: Multiple regression

$$\widehat{\text{Reading}} = b_0 + b_1(\text{ParentEd}) + b_2(\text{Books})$$

```{r}
m2 <- lm(READ_AVG ~ PAREDINT_c + BOOKSHOME_c, data = pisa)
summary(m2)
```


::: {.content-visible when-profile="teacher"}
**Sample answer.** This model predicts reading score from parent education
*and* books at home simultaneously, so each estimate is the effect of one
predictor holding the other constant ("controlling for"). $R^2$ rises to
`r fmt(summary(m2)$r.squared, 3)`, meaning the two predictors together explain
about `r fmt(summary(m2)$r.squared * 100, 0)`% of the variance in reading
score -- notably more than parent education alone. Controlling for books at
home, each additional year of parent education predicts
`r fmt(coef(m2)[["PAREDINT_c"]], 2)` more reading points; controlling for
parent education, each additional category of books at home predicts
`r fmt(coef(m2)[["BOOKSHOME_c"]], 2)` more reading points. Both are still
significant. The `PAREDINT_c` estimate dropped from
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` to `r fmt(coef(m2)[["PAREDINT_c"]], 2)`
because parent education and books at home are correlated
(`r fmt(cor(pisa$PAREDINT, pisa$BOOKSHOME, use = "pairwise.complete.obs"), 2)`).
Some of what looked like a parent-education effect in Model 1 was really
variance shared with books at home; once books at home is in the model, that
shared part gets credited to books at home instead. The intercept
(`r fmt(coef(m2)[["(Intercept)"]], 1)`) is now the predicted reading score for
a student at average parent education *and* average books at home.
:::

## Model 3: Multiple regression with an interaction

$$\widehat{\text{Reading}} = b_0 + b_1(\text{ParentEd}) + b_2(\text{Books}) + b_3(\text{ParentEd} \times \text{Books})$$

```{r}
m3 <- lm(READ_AVG ~ PAREDINT_c * BOOKSHOME_c, data = pisa)
summary(m3)
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** The interaction lets the slope of one predictor depend on
the level of the other -- e.g., the effect of parent education on reading
could be stronger or weaker depending on how many books are in the home. Here
$R^2$ barely moves, from `r fmt(summary(m2)$r.squared, 3)` to
`r fmt(summary(m3)$r.squared, 3)`. Because both predictors are mean-centered,
the main effects still read as average slopes (the slope of one predictor at
the *average* level of the other), rather than slopes at "the other predictor
= 0" -- a point off the scale that can otherwise flip a main effect's sign for
purely arithmetic reasons. The interaction term itself
(`r fmt(coef(m3)[["PAREDINT_c:BOOKSHOME_c"]], 3)`) has
$p$ = `r fmt(summary(m3)$coefficients["PAREDINT_c:BOOKSHOME_c", "Pr(>|t|)"], 3)`,
which is not conventionally significant at $\alpha = .05$ (it's marginal). If
it *were* significant, it would mean parent education and books at home don't
just add up -- they combine to predict reading jointly, with one predictor's
effect changing across levels of the other.
:::

## Model Comparison

```{r}
tibble(
  Model     = c("M1: PAREDINT", "M1a: BOOKSHOME", "M2: PAREDINT + BOOKSHOME", "M3: PAREDINT x BOOKSHOME"),
  R2        = c(summary(m1)$r.squared, summary(m1a)$r.squared, summary(m2)$r.squared, summary(m3)$r.squared),
  `Adj. R2` = c(summary(m1)$adj.r.squared, summary(m1a)$adj.r.squared, summary(m2)$adj.r.squared, summary(m3)$adj.r.squared),
  AIC       = c(AIC(m1), AIC(m1a), AIC(m2), AIC(m3))
)
```


::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 2 (or 3, which is barely different) fits best: higher
$R^2$ and adjusted $R^2$, lower AIC than Model 1. Whether it's "good" depends
on the standard. It's statistically solid -- both predictors are significant
with a huge sample size -- but it still leaves about
`r fmt((1 - summary(m2)$r.squared) * 100, 0)`% of the variance in reading
score unexplained, and the residual standard error is about
`r fmt(sigma(m2), 0)` points, which is large relative to the outcome. So it's
a real, significant relationship but a weak-to-moderate predictive model.
Statistical significance is not the same as a model being complete or highly
predictive. It's also worth noting all three models treat parent education
and books at home as two competing predictors of the same outcome -- none of
them can say whether parent education might work *through* books at home
rather than alongside it. That's the question the SEM model below is built
to ask.
:::

