---
title: "Lab: Regression Review & SEM FUNdamentals"
date: today
description: '<a href="rlab-downloads/w02_rlab_review_SEM_FUNdamentals.qmd" download="w02_rlab_review_SEM_FUNdamentals.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w01_data_pisa22_USA_cln.csv" download="w01_data_pisa22_USA_cln.csv">w01_data_pisa22_USA_cln.csv</a>'
format: html
release: week 2
draft: true
---

## Purpose

In lecture we ran three familiar models predicting reading score: a simple
regression, a multiple regression, and a multiple regression with an
interaction. In this lab you'll run those same three models yourself and
practice reading and interpreting the output. Then we'll specify the same
variables as a very simple SEM path model in `lavaan` and see what changes.

The outcome is always reading score (`READ_AVG`). The predictors available are
parent education (`PAREDINT`) and books at home (`BOOKSHOME`).

## Setup

```{r}
#| message: false
library(tidyverse)
library(lavaan)
library(lavaanPlot)

pisa <- read_csv("w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)
```

Quick reminder of the variables (see the Week 1 lab for more detail):

| Variable | What it is |
|---|---|
| `PAREDINT` | Highest parental education in years of schooling |
| `BOOKSHOME` | Books in the home, an ordered 1-5 Likert-style response |
| `READ_AVG` | Reading literacy score |

Neither predictor has a real 0 in this data (no student has 0 years of
parental education), so an intercept from the raw variables would describe a
student who doesn't exist. We'll mean-center both predictors so the intercept
describes a student at the *average* level of each -- a point that's actually
in the data. Centering only shifts the intercept; it does not change the
slopes, $R^2$, or significance tests.

```{r}
pared_mean  <- mean(pisa$PAREDINT, na.rm = TRUE)
books_mean  <- mean(pisa$BOOKSHOME, na.rm = TRUE)

pisa <- pisa |>
  mutate(
    PAREDINT_c  = PAREDINT - pared_mean,
    BOOKSHOME_c = BOOKSHOME - books_mean
  )
```

```{r}
#| include: false
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

## Model 1: Simple regression

$$\widehat{\text{Reading}} = b_0 + b_1(\text{ParentEd})$$

```{r}
m1 <- lm(READ_AVG ~ PAREDINT_c, data = pisa)
summary(m1)
```

**Questions**

1. In your own words, what is this model doing?
2. $R^2$ indicates ________________________________________. What is its value here, and what does that value tell you about how well parent education predicts reading score?
3. The estimate for `PAREDINT_c` indicates ________________________________________. Is it statistically significant? How do you know?
4. What does the intercept represent here, now that `PAREDINT_c` is mean-centered?

::: {.content-visible when-profile="teacher"}
**Sample answer.** This model predicts reading score from a single variable,
parent education. $R^2$ indicates the proportion of variance in reading score
that is explained by (accounted for by) the model -- here
`r fmt(summary(m1)$r.squared, 3)`, so parent education alone explains only
about `r fmt(summary(m1)$r.squared * 100, 0)`% of the variation in reading
scores. The estimate for `PAREDINT_c`
(`r fmt(coef(m1)[["PAREDINT_c"]], 2)`) indicates that each additional year of
parental education is associated with about
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` more reading-score points, on average. It
is statistically significant ($p < .001$): the relationship is very unlikely
to be zero in the population. Because `PAREDINT_c` is centered at its mean
(`r fmt(pared_mean, 1)` years), the intercept
(`r fmt(coef(m1)[["(Intercept)"]], 1)`) is the predicted reading score for a
student at *average* parental education, not for a (nonexistent) student with
0 years of schooling. The slope is unchanged by centering -- only the
intercept's meaning changes.
:::

## Model 2: Multiple regression

$$\widehat{\text{Reading}} = b_0 + b_1(\text{ParentEd}) + b_2(\text{Books})$$

```{r}
m2 <- lm(READ_AVG ~ PAREDINT_c + BOOKSHOME_c, data = pisa)
summary(m2)
```

**Questions**

1. What is this model doing differently from Model 1?
2. $R^2$ indicates ________________________________________. How does its value compare to Model 1's?
3. The estimates now indicate ________________________________________. Interpret each one.
4. The estimate for `PAREDINT_c` changed from Model 1 to Model 2. What happened, and why might that be?

::: {.content-visible when-profile="teacher"}
**Sample answer.** This model predicts reading score from parent education
*and* books at home simultaneously, so each estimate is the effect of one
predictor holding the other constant ("controlling for"). $R^2$ rises to
`r fmt(summary(m2)$r.squared, 3)`, meaning the two predictors together explain
about `r fmt(summary(m2)$r.squared * 100, 0)`% of the variance in reading
score -- notably more than parent education alone. Controlling for books at
home, each additional year of parent education predicts
`r fmt(coef(m2)[["PAREDINT_c"]], 2)` more reading points; controlling for
parent education, each additional category of books at home predicts
`r fmt(coef(m2)[["BOOKSHOME_c"]], 2)` more reading points. Both are still
significant. The `PAREDINT_c` estimate dropped from
`r fmt(coef(m1)[["PAREDINT_c"]], 2)` to `r fmt(coef(m2)[["PAREDINT_c"]], 2)`
because parent education and books at home are correlated
(`r fmt(cor(pisa$PAREDINT, pisa$BOOKSHOME, use = "pairwise.complete.obs"), 2)`).
Some of what looked like a parent-education effect in Model 1 was really
variance shared with books at home; once books at home is in the model, that
shared part gets credited to books at home instead. The intercept
(`r fmt(coef(m2)[["(Intercept)"]], 1)`) is now the predicted reading score for
a student at average parent education *and* average books at home.
:::

## Model 3: Multiple regression with an interaction

$$\widehat{\text{Reading}} = b_0 + b_1(\text{ParentEd}) + b_2(\text{Books}) + b_3(\text{ParentEd} \times \text{Books})$$

```{r}
m3 <- lm(READ_AVG ~ PAREDINT_c * BOOKSHOME_c, data = pisa)
summary(m3)
```

**Questions**

1. What does adding the interaction term let this model do that Model 2 could not?
2. $R^2$ indicates ________________________________________. How much did it change from Model 2?
3. The estimates now indicate ________________________________________, including the interaction term specifically.
4. Is the interaction statistically significant? What would it mean if it were?

::: {.content-visible when-profile="teacher"}
**Sample answer.** The interaction lets the slope of one predictor depend on
the level of the other -- e.g., the effect of parent education on reading
could be stronger or weaker depending on how many books are in the home. Here
$R^2$ barely moves, from `r fmt(summary(m2)$r.squared, 3)` to
`r fmt(summary(m3)$r.squared, 3)`. Because both predictors are mean-centered,
the main effects still read as average slopes (the slope of one predictor at
the *average* level of the other), rather than slopes at "the other predictor
= 0" -- a point off the scale that can otherwise flip a main effect's sign for
purely arithmetic reasons. The interaction term itself
(`r fmt(coef(m3)[["PAREDINT_c:BOOKSHOME_c"]], 3)`) has
$p$ = `r fmt(summary(m3)$coefficients["PAREDINT_c:BOOKSHOME_c", "Pr(>|t|)"], 3)`,
which is not conventionally significant at $\alpha = .05$ (it's marginal). If
it *were* significant, it would mean parent education and books at home don't
just add up -- they combine to predict reading jointly, with one predictor's
effect changing across levels of the other.
:::

## Which model is best? Is it a good model?

```{r}
tibble(
  Model     = c("M1: PAREDINT", "M2: PAREDINT + BOOKSHOME", "M3: PAREDINT x BOOKSHOME"),
  R2        = c(summary(m1)$r.squared, summary(m2)$r.squared, summary(m3)$r.squared),
  `Adj. R2` = c(summary(m1)$adj.r.squared, summary(m2)$adj.r.squared, summary(m3)$adj.r.squared),
  AIC       = c(AIC(m1), AIC(m2), AIC(m3))
)
```

**Questions**

1. Based on $R^2$, adjusted $R^2$, and AIC, which of the three models fits best?
2. Is that model a *good* model? What do you mean by "good" -- statistically significant, practically useful, both, neither?
3. What is this model still missing? What would you want to know about reading score that it can't tell you?

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 2 (or 3, which is barely different) fits best: higher
$R^2$ and adjusted $R^2$, lower AIC than Model 1. Whether it's "good" depends
on the standard. It's statistically solid -- both predictors are significant
with a huge sample size -- but it still leaves about
`r fmt((1 - summary(m2)$r.squared) * 100, 0)`% of the variance in reading
score unexplained, and the residual standard error is about
`r fmt(sigma(m2), 0)` points, which is large relative to the outcome. So it's
a real, significant relationship but a weak-to-moderate predictive model.
Statistical significance is not the same as a model being complete or highly
predictive. It's also worth noting all three models treat parent education
and books at home as two competing predictors of the same outcome -- none of
them can say whether parent education might work *through* books at home
rather than alongside it. That's the question the SEM model below is built
to ask.
:::

## SEM FUNdamentals: a simple path model

Instead of putting `PAREDINT` and `BOOKSHOME` into one equation as competing
predictors, we can specify a small *system* of equations: parent education
predicts books at home, and books at home predicts reading score.

$$\widehat{\text{Books}} = a_0 + a_1(\text{ParentEd}) \qquad
\widehat{\text{Reading}} = b_0 + b_1(\text{Books})$$

In `lavaan`, each regression is written with `~` ("is predicted by") and the
whole system is estimated at once:

```{r}
#| warning: false
path_model <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"

fit <- sem(path_model, data = pisa)
summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

`lavaan` itself has no built-in plotting -- diagrams come from a companion
package. `lavaanPlot` draws the model straight from the fitted object, so the
picture is always in sync with the numbers above:

```{r}
lavaanPlot(model = fit, coefs = TRUE, stand = TRUE, stars = "regress")
```

**Questions**

1. `BOOKSHOME` is on the left of one equation and the right of another. Could you have specified that in a single `lm()` call? Why or why not?
2. What does the $R^2$ for the `BOOKSHOME` equation mean? What does the $R^2$ for the `READ_AVG` equation mean? How are these different from a single $R^2$ for the whole model?
3. Compare the `BOOKSHOME ~ PAREDINT` estimate here to Model 2's `PAREDINT` estimate. Why are they different?
4. This is the one model in this lab where you're given a Chi-square test of *model fit* ("Model Test User Model"). None of the regression models could give you that. What do you think that test is checking?

::: {.content-visible when-profile="teacher"}
**Sample answer.** No single `lm()` call can do this: `lm()` estimates one
equation with one outcome, and here `BOOKSHOME` needs to be an outcome in one
equation and a predictor in another. That is exactly what regression, multiple
regression, and interaction models cannot represent -- they force every
variable to be either a predictor or an outcome, never both at once.

There are two separate $R^2$ values because there are two separate equations:
$R^2$ = `r fmt(lavInspect(fit, "rsquare")[["BOOKSHOME"]], 3)` for how well
parent education predicts books at home, and $R^2$ =
`r fmt(lavInspect(fit, "rsquare")[["READ_AVG"]], 3)` for how well books at
home predicts reading. A single regression only ever gives you one $R^2$ for
one outcome; this model is explaining variance in *two* variables at once.

The `BOOKSHOME ~ PAREDINT` estimate here
(`r fmt(coef(fit)[["BOOKSHOME~PAREDINT"]], 3)`) is a *simple* regression of
books on parent education -- it is not adjusted for anything else, so it is
not directly comparable to Model 2's `PAREDINT` coefficient, which was
adjusted for books at home. This is the same "simple vs. controlled" logic
from Models 1 and 2, just applied to a different pair of variables.

The chi-square test compares the correlations/covariances this model implies
among `PAREDINT`, `BOOKSHOME`, and `READ_AVG` to the correlations/covariances
actually observed in the data. A regression is always *saturated* -- it has
exactly as many parameters as there are pieces of information to fit, so it
always fits perfectly and there is nothing to test. This path model leaves out
a direct arrow from `PAREDINT` to `READ_AVG`, which is a testable claim (that
parent education matters for reading *only* through books at home). With df =
`r as.integer(fitMeasures(fit, "df"))` and $p$ =
`r fmt(fitMeasures(fit, "pvalue"), 3)`, that claim doesn't perfectly hold in
this sample -- a hint that a direct path might belong in the model too, which
is exactly where we're headed next week.
:::

### Going further: the indirect effect (optional)

We can label the two paths and ask `lavaan` to compute their product directly
-- the *indirect effect* of parent education on reading, transmitted through
books at home.

```{r}
#| warning: false
path_model_labeled <- "
  BOOKSHOME ~ a*PAREDINT
  READ_AVG  ~ b*BOOKSHOME

  indirect := a*b
"

fit_labeled <- sem(path_model_labeled, data = pisa)
parameterEstimates(fit_labeled)
```

```{r}
lavaanPlot(model = fit_labeled, coefs = TRUE, stand = TRUE, stars = "regress")
```

The `indirect` row is $a \times b$: the predicted change in reading score for
a one-year increase in parent education, acting *through* books at home.
Notice the diagram looks identical to the one above -- `lavaanPlot` only draws
the actual `~` paths in the model (the arrows we specified), and `indirect`
isn't one of them. It's a *defined parameter*, a quantity computed from `a`
and `b` after the fact, not a separate arrow lavaan estimated. Same model,
same picture; the indirect effect is a summary of two of its paths, not a
third path.

::: {.content-visible when-profile="teacher"}
**Sample answer / notes.** Indirect effect =
`r fmt(coef(fit_labeled)[["a"]], 3)` x `r fmt(coef(fit_labeled)[["b"]], 2)` =
`r fmt(coef(fit_labeled)[["a"]] * coef(fit_labeled)[["b"]], 2)` reading points
per year of parent education, and it's significant. This is a preview of
mediation -- formal treatment (and the direct path suggested by the
significant chi-square above) is coming in the next couple of weeks.
:::

## Wrap-up

1. Of the three regression models and the SEM path model, which one gives you the most complete story about how parent education, books at home, and reading score relate to each other? Why?
2. What is fundamentally different about how the SEM model is *specified*, regardless of which model has the higher $R^2$?

::: {.content-visible when-profile="teacher"}
**Sample answer.** "Best $R^2$" and "most complete story" are different
questions. Model 2/3 fit the data about as well in the narrow sense of
variance explained, but the SEM model is the only one that lets `BOOKSHOME`
play both roles (outcome and predictor) and gives a directional,
mediation-style account: parent education -> books at home -> reading. The
regression models can only ever ask "how much does each predictor add,"
never "does one predictor work through the other." That structural
difference -- not a bigger $R^2$ -- is the point of this lab.
:::
