---
title: "Lab: SEM FUNdamentals"
date: today
description: '<a href="rlab-downloads/rlabs/w03_sem.qmd" download="rlabs/w03_sem.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w01_data_pisa22_USA_cln.csv" download="w01_data_pisa22_USA_cln.csv">w01_data_pisa22_USA_cln.csv</a>'
format: html
release: week 3
draft: true
---


## SEM FUNdamentals: a simple path model

Instead of putting `PAREDINT` and `BOOKSHOME` into one equation as competing
predictors, we can specify a small *system* of equations: parent education
predicts books at home, and books at home predicts reading score.

$$\widehat{\text{Books}} = a_0 + a_1(\text{ParentEd}) \qquad
\widehat{\text{Reading}} = b_0 + b_1(\text{Books})$$

In `lavaan`, each regression is written with `~` ("is predicted by") and the
whole system is estimated at once:

```{r}
#| warning: false
path_model <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"

fit <- sem(path_model, data = pisa)
summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

`lavaan` itself has no built-in plotting -- diagrams come from a companion
package. `lavaanPlot` draws the model straight from the fitted object, so the
picture is always in sync with the numbers above:

```{r}
lavaanPlot(model = fit, coefs = TRUE, stand = TRUE, stars = "regress")
```

**Questions**

1. `BOOKSHOME` is on the left of one equation and the right of another. Could you have specified that in a single `lm()` call? Why or why not?
2. What does the $R^2$ for the `BOOKSHOME` equation mean? What does the $R^2$ for the `READ_AVG` equation mean? How are these different from a single $R^2$ for the whole model?
3. Compare the `BOOKSHOME ~ PAREDINT` estimate here to Model 2's `PAREDINT` estimate. Why are they different?
4. This is the one model in this lab where you're given a Chi-square test of *model fit* ("Model Test User Model"). None of the regression models could give you that. What do you think that test is checking?

::: {.content-visible when-profile="teacher"}
**Sample answer.** No single `lm()` call can do this: `lm()` estimates one
equation with one outcome, and here `BOOKSHOME` needs to be an outcome in one
equation and a predictor in another. That is exactly what regression, multiple
regression, and interaction models cannot represent -- they force every
variable to be either a predictor or an outcome, never both at once.

There are two separate $R^2$ values because there are two separate equations:
$R^2$ = `r fmt(lavInspect(fit, "rsquare")[["BOOKSHOME"]], 3)` for how well
parent education predicts books at home, and $R^2$ =
`r fmt(lavInspect(fit, "rsquare")[["READ_AVG"]], 3)` for how well books at
home predicts reading. A single regression only ever gives you one $R^2$ for
one outcome; this model is explaining variance in *two* variables at once.

The `BOOKSHOME ~ PAREDINT` estimate here
(`r fmt(coef(fit)[["BOOKSHOME~PAREDINT"]], 3)`) is a *simple* regression of
books on parent education -- it is not adjusted for anything else, so it is
not directly comparable to Model 2's `PAREDINT` coefficient, which was
adjusted for books at home. This is the same "simple vs. controlled" logic
from Models 1 and 2, just applied to a different pair of variables.

The chi-square test compares the correlations/covariances this model implies
among `PAREDINT`, `BOOKSHOME`, and `READ_AVG` to the correlations/covariances
actually observed in the data. A regression is always *saturated* -- it has
exactly as many parameters as there are pieces of information to fit, so it
always fits perfectly and there is nothing to test. This path model leaves out
a direct arrow from `PAREDINT` to `READ_AVG`, which is a testable claim (that
parent education matters for reading *only* through books at home). With df =
`r as.integer(fitMeasures(fit, "df"))` and $p$ =
`r fmt(fitMeasures(fit, "pvalue"), 3)`, that claim doesn't perfectly hold in
this sample -- a hint that a direct path might belong in the model too, which
is exactly where we're headed next week.
:::

### Going further: the indirect effect (optional)

We can label the two paths and ask `lavaan` to compute their product directly
-- the *indirect effect* of parent education on reading, transmitted through
books at home.

```{r}
#| warning: false
path_model_labeled <- "
  BOOKSHOME ~ a*PAREDINT
  READ_AVG  ~ b*BOOKSHOME

  indirect := a*b
"

fit_labeled <- sem(path_model_labeled, data = pisa)
parameterEstimates(fit_labeled)
```

```{r}
lavaanPlot(model = fit_labeled, coefs = TRUE, stand = TRUE, stars = "regress")
```

The `indirect` row is $a \times b$: the predicted change in reading score for
a one-year increase in parent education, acting *through* books at home.
Notice the diagram looks identical to the one above -- `lavaanPlot` only draws
the actual `~` paths in the model (the arrows we specified), and `indirect`
isn't one of them. It's a *defined parameter*, a quantity computed from `a`
and `b` after the fact, not a separate arrow lavaan estimated. Same model,
same picture; the indirect effect is a summary of two of its paths, not a
third path.

::: {.content-visible when-profile="teacher"}
**Sample answer / notes.** Indirect effect =
`r fmt(coef(fit_labeled)[["a"]], 3)` x `r fmt(coef(fit_labeled)[["b"]], 2)` =
`r fmt(coef(fit_labeled)[["a"]] * coef(fit_labeled)[["b"]], 2)` reading points
per year of parent education, and it's significant. This is a preview of
mediation -- formal treatment (and the direct path suggested by the
significant chi-square above) is coming in the next couple of weeks.
:::

## Wrap-up

1. Of the three regression models and the SEM path model, which one gives you the most complete story about how parent education, books at home, and reading score relate to each other? Why?
2. What is fundamentally different about how the SEM model is *specified*, regardless of which model has the higher $R^2$?

::: {.content-visible when-profile="teacher"}
**Sample answer.** "Best $R^2$" and "most complete story" are different
questions. Model 2/3 fit the data about as well in the narrow sense of
variance explained, but the SEM model is the only one that lets `BOOKSHOME`
play both roles (outcome and predictor) and gives a directional,
mediation-style account: parent education -> books at home -> reading. The
regression models can only ever ask "how much does each predictor add,"
never "does one predictor work through the other." That structural
difference -- not a bigger $R^2$ -- is the point of this lab.
:::
