---
title: "Lab: Model Fit & Comparison"
date: today
description: '<a href="rlab-downloads/w04_rlab_model_fit.qmd" download="w04_rlab_model_fit.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w01_data_pisa22_USA_cln.csv" download="w01_data_pisa22_USA_cln.csv">w01_data_pisa22_USA_cln.csv</a>'
format: html
release: week 4
---

## Purpose

Goal: Fit two competing path models, check the assumptions behind them, judge
how well each fits, and compare them.

Question: Is the relationship between parental education (`PAREDINT`) and
children's reading scores (`READ_AVG`) *fully* mediated by books at home
(`BOOKSHOME`), or does parent education also have a *direct* effect on reading?

* **Model 1 (fully mediated):** parent education -> books -> reading, with no
  direct path. This leaves one covariance unaccounted for, so it is
  *over-identified* (df = 1) and its fit is testable.
* **Model 2 (partially mediated):** adds the direct parent education -> reading
  path. Every covariance now has a parameter, so it is *just-identified*
  (df = 0) and fits perfectly by construction.

## Setup

```{r}
#| message: false
library(tidyverse)
library(lavaan)
library(lavaanPlot)

pisa <- read_csv("w01_data_pisa22_USA_cln.csv", show_col_types = FALSE)
```

```{r}
#| include: false
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

## Assumptions

Path models estimated with maximum likelihood carry a handful of assumptions.
We check them *before* trusting the fit statistics.

### Multivariate normality

ML assumes the endogenous variables are multivariate normal. A quick screen is
the univariate skewness and kurtosis of each variable -- values near 0
(skew) and 3 (kurtosis) are consistent with normality.

```{r}
skew <- function(x) {
  x <- x[!is.na(x)]; n <- length(x)
  mean((x - mean(x))^3) / (mean((x - mean(x))^2))^(3 / 2)
}
kurt <- function(x) {
  x <- x[!is.na(x)]; n <- length(x)
  mean((x - mean(x))^4) / (mean((x - mean(x))^2))^2
}

tibble(
  Variable = c("PAREDINT", "BOOKSHOME", "READ_AVG"),
  Skewness = c(skew(pisa$PAREDINT), skew(pisa$BOOKSHOME), skew(pisa$READ_AVG)),
  Kurtosis = c(kurt(pisa$PAREDINT), kurt(pisa$BOOKSHOME), kurt(pisa$READ_AVG))
)
```

If normality is in doubt, refit with a robust estimator, `estimator = "MLR"`,
which corrects the standard errors and the chi-square for non-normality.

::: {.content-visible when-profile="teacher"}
**Sample answer.** With a large PISA sample, mild departures from normality are
common and rarely fatal. If skew/kurtosis are moderate, ML point estimates stay
trustworthy but the chi-square and standard errors can be off -- that is exactly
what `MLR` fixes, so reporting the robust (scaled) statistics is the safe habit.
:::

### Linearity

The `~` paths assume each relationship is linear. Scatterplots with a straight
(`lm`) fit overlaid let us eyeball that.

```{r}
#| warning: false
#| fig-width: 8
#| fig-height: 3
pisa |>
  select(PAREDINT, BOOKSHOME, READ_AVG) |>
  pivot_longer(-READ_AVG, names_to = "predictor", values_to = "x") |>
  ggplot(aes(x, READ_AVG)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C81E7A") +
  facet_wrap(~predictor, scales = "free_x") +
  theme_minimal()
```

### Multicollinearity

Exogenous predictors should not be near-perfectly correlated. Here the only
predictor pair that appears together in an equation is parent education and
books at home.

```{r}
cor(pisa[, c("PAREDINT", "BOOKSHOME", "READ_AVG")], use = "pairwise.complete.obs")

# Variance inflation factor for the PAREDINT / BOOKSHOME pair
r2 <- summary(lm(BOOKSHOME ~ PAREDINT, data = pisa))$r.squared
vif <- 1 / (1 - r2)
vif
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Correlation between parent education and books at home is
`r fmt(cor(pisa$PAREDINT, pisa$BOOKSHOME, use = "pairwise.complete.obs"), 2)`,
well below any collinearity threshold, and VIF = `r fmt(vif, 2)` (rule of thumb:
concern above ~5-10). No collinearity problem.
:::

### Uncorrelated disturbances

Recursive path models assume the disturbances (residual errors) of the
endogenous variables are uncorrelated. After fitting, we check the residual
covariances and modification indices for a hint that a disturbance covariance
was wrongly omitted. (We run this check on the fitted models below.)

## Model 1: Fully mediated

```{r}
#| warning: false
model1 <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"

fit1 <- sem(model1, data = pisa)
summary(fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```{r}
lavaanPlot(model = fit1, coefs = TRUE, stand = TRUE, stars = "regress")
```

Residual check for the uncorrelated-disturbance assumption:

```{r}
residuals(fit1, type = "cor")
modificationIndices(fit1, sort. = TRUE)
```

### How well does Model 1 fit?

```{r}
fitMeasures(fit1, c("chisq", "df", "pvalue", "cfi", "tli", "rmsea",
                    "srmr", "aic", "bic"))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 1 is over-identified (df =
`r as.integer(fitMeasures(fit1, "df"))`), so its fit is testable. The
chi-square is `r fmt(fitMeasures(fit1, "chisq"), 1)` with $p$ =
`r fmt(fitMeasures(fit1, "pvalue"), 3)`. A significant chi-square says the
model-implied covariances differ from the observed ones -- the omitted direct
path is missing something. The approximate-fit indices (CFI =
`r fmt(fitMeasures(fit1, "cfi"), 3)`, RMSEA =
`r fmt(fitMeasures(fit1, "rmsea"), 3)`, SRMR =
`r fmt(fitMeasures(fit1, "srmr"), 3)`) tell the same story. The largest
modification index points straight at the `READ_AVG ~ PAREDINT` direct path --
i.e., Model 2.
:::

## Model 2: Partially mediated

```{r}
#| warning: false
model2 <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME + PAREDINT
"

fit2 <- sem(model2, data = pisa)
summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```{r}
lavaanPlot(model = fit2, coefs = TRUE, stand = TRUE, stars = "regress")
```

### How well does Model 2 fit?

```{r}
fitMeasures(fit2, c("chisq", "df", "pvalue", "cfi", "tli", "rmsea",
                    "srmr", "aic", "bic"))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 2 is just-identified (df =
`r as.integer(fitMeasures(fit2, "df"))`): it has exactly as many parameters as
pieces of information, so it reproduces the observed covariances perfectly.
Chi-square is ~0, CFI = 1, RMSEA = 0 -- these are guaranteed, not evidence of a
good model, so they are *not* meaningful here. What *is* meaningful is whether
the added direct path is significant and whether AIC/BIC improve over Model 1.
:::

## Is Model 1 better than Model 2?

Because Model 1 is nested in Model 2 (Model 1 = Model 2 with the direct path
fixed to 0), we can compare them with a chi-square difference test, and also
weigh parsimony with AIC/BIC.

```{r}
anova(fit1, fit2)
```

```{r}
tibble(
  Model  = c("Model 1: fully mediated", "Model 2: partially mediated"),
  df     = c(fitMeasures(fit1, "df"),    fitMeasures(fit2, "df")),
  chisq  = c(fitMeasures(fit1, "chisq"), fitMeasures(fit2, "chisq")),
  CFI    = c(fitMeasures(fit1, "cfi"),   fitMeasures(fit2, "cfi")),
  RMSEA  = c(fitMeasures(fit1, "rmsea"), fitMeasures(fit2, "rmsea")),
  AIC    = c(fitMeasures(fit1, "aic"),   fitMeasures(fit2, "aic")),
  BIC    = c(fitMeasures(fit1, "bic"),   fitMeasures(fit2, "bic"))
)
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** The chi-square difference test is significant, so freeing the
direct path improves fit beyond chance: Model 2 is preferred. AIC and BIC are
both lower for Model 2, agreeing even after penalizing the extra parameter.
Substantively, parent education has a direct effect on reading over and above
what runs through books at home -- the relationship is only *partially*
mediated. (Model 2 being just-identified means it can't be rejected on its own;
the comparison to the nested Model 1 is what makes the direct path defensible.)
:::

## Wrap-up

1. Why can we test the fit of Model 1 but not Model 2?
2. Which single statistic in the comparison most directly answers the research
   question, and what does its value tell you?
3. If multivariate normality had been badly violated, what would you change
   about how you fit and compared these models?
