---
title: "Lab: Model Fit & Comparison"
date: today
description: '<a href="rlab-downloads/w04_rlab_model_fit.qmd" download="w04_rlab_model_fit.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w04_data_pisa22_USA_cln.csv" download="w04_data_pisa22_USA_cln.csv">w04_data_pisa22_USA_cln.csv</a>'
format: html
release: week 4
---

## Purpose

Goal: Fit two *nested* path models, check the assumptions behind them, judge how
well each fits, and compare them.

Question: Is the relationship between parental education (`PAREDINT`) and
children's reading scores (`READ_AVG`) mediated by books at home (`BOOKSHOME`)
alone, or does technology (`TECH`) *also* carry part of the effect?

* **Model 1 (books-only mediation):** parent education -> books -> reading.
  Technology is in the model (predicted by parent education) but has **no** path
  to reading. Keeping `TECH` in the model is what puts both models on the same
  four variables so they are comparable.
* **Model 2 (books + tech mediation):** adds the `TECH -> READ_AVG` path.

Model 1 is Model 2 with that one path fixed to 0, so the two are **nested** and
can be compared with a chi-square difference test and with AIC/BIC.

## Setup

```{r}
#| message: false
library(tidyverse)
library(lavaan)
library(lavaanPlot)

pisa <- read_csv("w04_data_pisa22_USA_cln.csv", show_col_types = FALSE)
```

```{r}
#| include: false
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")

# Focused fit table: the headline indices with conventional good-fit cutoffs.
fit_table <- function(fit) {
  m <- fitMeasures(fit, c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr"))
  p <- if (m[["pvalue"]] < 0.001) "< .001" else paste0("= ", fmt(m[["pvalue"]], 3))
  tibble(
    Index      = c("χ² (df)", "CFI", "RMSEA", "SRMR"),
    Value      = c(sprintf("%s (%d), p %s", fmt(m[["chisq"]], 1), as.integer(m[["df"]]), p),
                   fmt(m[["cfi"]], 3), fmt(m[["rmsea"]], 3), fmt(m[["srmr"]], 3)),
    `Good fit` = c("p > .05", "≥ .95", "≤ .06", "≤ .08")
  )
}
```

## Assumptions

Path models estimated with maximum likelihood carry a handful of assumptions.
We check them *before* trusting the fit statistics.

### Multivariate normality

ML assumes the endogenous variables are multivariate normal. A quick screen is
the univariate skewness and kurtosis of each variable -- values near 0
(skew) and 3 (kurtosis) are consistent with normality.

```{r}
skew <- function(x) {
  x <- x[!is.na(x)]
  mean((x - mean(x))^3) / (mean((x - mean(x))^2))^(3 / 2)
}
kurt <- function(x) {
  x <- x[!is.na(x)]
  mean((x - mean(x))^4) / (mean((x - mean(x))^2))^2
}

vars <- c("PAREDINT", "BOOKSHOME", "TECH", "READ_AVG")
tibble(
  Variable = vars,
  Skewness = map_dbl(vars, ~ skew(pisa[[.x]])),
  Kurtosis = map_dbl(vars, ~ kurt(pisa[[.x]]))
)
```

If normality is in doubt, refit with a robust estimator, `estimator = "MLR"`,
which corrects the standard errors and the chi-square for non-normality.

::: {.content-visible when-profile="teacher"}
**Sample answer.** With a large PISA sample, mild departures from normality are
common and rarely fatal. If skew/kurtosis are moderate, ML point estimates stay
trustworthy but the chi-square and standard errors can be off -- that is exactly
what `MLR` fixes, so reporting the robust (scaled) statistics is the safe habit.
:::

### Linearity

The `~` paths assume each relationship is linear. Scatterplots with a straight
(`lm`) fit overlaid let us eyeball the two paths into reading.

```{r}
#| warning: false
#| fig-width: 8
#| fig-height: 3
pisa |>
  select(BOOKSHOME, TECH, READ_AVG) |>
  pivot_longer(-READ_AVG, names_to = "predictor", values_to = "x") |>
  ggplot(aes(x, READ_AVG)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C81E7A") +
  facet_wrap(~predictor, scales = "free_x") +
  theme_minimal()
```

### Multicollinearity

The two predictors that appear together in the reading equation are books at
home and technology. They should not be near-perfectly correlated.

```{r}
cor(pisa[, vars], use = "pairwise.complete.obs")

# Variance inflation factor for the BOOKSHOME / TECH pair
r2  <- summary(lm(BOOKSHOME ~ TECH, data = pisa))$r.squared
vif <- 1 / (1 - r2)
vif
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Correlation between books at home and technology is
`r fmt(cor(pisa$BOOKSHOME, pisa$TECH, use = "pairwise.complete.obs"), 2)`, and
VIF = `r fmt(vif, 2)` (rule of thumb: concern above ~5-10). No collinearity
problem -- the two mediators carry distinct information.
:::

### Uncorrelated disturbances

Recursive path models assume the disturbances (residual errors) of the
endogenous variables are uncorrelated. We enforce this when fitting with
`auto.cov.y = FALSE` (otherwise lavaan would freely correlate the residuals of
`BOOKSHOME`, `TECH`, and `READ_AVG`, which would also make the two models
statistically equivalent instead of nested). After fitting we check the residual
covariances and modification indices for any hint that a disturbance covariance
was wrongly omitted.

## Model 0: The original chain

Before adding technology, start with the original model: parent education relates
to reading achievement *only* through books at home.

$$\text{PAREDINT} \;\rightarrow\; \text{BOOKSHOME} \;\rightarrow\; \text{READ\_AVG}$$

```{r}
#| warning: false
model0 <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"

fit0 <- sem(model0, data = pisa, auto.cov.y = FALSE)
summary(fit0, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```{r}
lavaanPlot(model = fit0, coefs = TRUE, stand = TRUE, stars = "regress")
```

### How well does Model 0 fit?

```{r}
knitr::kable(fit_table(fit0))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 0 is over-identified (df =
`r as.integer(fitMeasures(fit0, "df"))`), so its fit is testable. The chi-square
is `r fmt(fitMeasures(fit0, "chisq"), 1)` with $p$ =
`r fmt(fitMeasures(fit0, "pvalue"), 3)` (CFI =
`r fmt(fitMeasures(fit0, "cfi"), 3)`, RMSEA =
`r fmt(fitMeasures(fit0, "rmsea"), 3)`, SRMR =
`r fmt(fitMeasures(fit0, "srmr"), 3)`). This three-variable chain is the baseline
we build on -- next we bring technology into the model so the richer models stay
comparable on the same variables.
:::

## Model 1: Books-only mediation

```{r}
#| warning: false
model1 <- "
  BOOKSHOME ~ PAREDINT
  TECH      ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"

fit1 <- sem(model1, data = pisa, auto.cov.y = FALSE)
summary(fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```{r}
lavaanPlot(model = fit1, coefs = TRUE, stand = TRUE, stars = "regress")
```

Residual check for the uncorrelated-disturbance assumption:

```{r}
residuals(fit1, type = "cor")
modificationIndices(fit1, sort. = TRUE)
```

### How well does Model 1 fit?

```{r}
knitr::kable(fit_table(fit1))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 1 is over-identified (df =
`r as.integer(fitMeasures(fit1, "df"))`), so its fit is testable. The
chi-square is `r fmt(fitMeasures(fit1, "chisq"), 1)` with $p$ =
`r fmt(fitMeasures(fit1, "pvalue"), 3)`, and the approximate-fit indices (CFI =
`r fmt(fitMeasures(fit1, "cfi"), 3)`, RMSEA =
`r fmt(fitMeasures(fit1, "rmsea"), 3)`, SRMR =
`r fmt(fitMeasures(fit1, "srmr"), 3)`) are mediocre -- this simple structure
does not fully reproduce the four-variable covariance matrix. The question is
whether *adding the technology path* fixes that.
:::

## Model 2: Books + tech mediation

```{r}
#| warning: false
model2 <- "
  BOOKSHOME ~ PAREDINT
  TECH      ~ PAREDINT
  READ_AVG  ~ BOOKSHOME + TECH
"

fit2 <- sem(model2, data = pisa, auto.cov.y = FALSE)
summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```{r}
lavaanPlot(model = fit2, coefs = TRUE, stand = TRUE, stars = "regress")
```

### How well does Model 2 fit?

```{r}
knitr::kable(fit_table(fit2))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 2 frees one more path (`READ_AVG ~ TECH`), so df =
`r as.integer(fitMeasures(fit2, "df"))`. Its fit is essentially unchanged from
Model 1 (CFI = `r fmt(fitMeasures(fit2, "cfi"), 3)`, RMSEA =
`r fmt(fitMeasures(fit2, "rmsea"), 3)`), and the `TECH -> READ_AVG` coefficient
itself is not significant. That is the first sign the extra path buys us little.
:::

## Is Model 2 better than Model 1?

Because Model 1 is nested in Model 2 (Model 1 = Model 2 with the
`TECH -> READ_AVG` path fixed to 0), we can compare them with a chi-square
difference test, and also weigh parsimony with AIC/BIC.

```{r}
anova(fit1, fit2)
```

```{r}
tibble(
  Model  = c("Model 1: books-only", "Model 2: books + tech"),
  df     = c(fitMeasures(fit1, "df"),    fitMeasures(fit2, "df")),
  chisq  = c(fitMeasures(fit1, "chisq"), fitMeasures(fit2, "chisq")),
  CFI    = c(fitMeasures(fit1, "cfi"),   fitMeasures(fit2, "cfi")),
  RMSEA  = c(fitMeasures(fit1, "rmsea"), fitMeasures(fit2, "rmsea")),
  AIC    = c(fitMeasures(fit1, "aic"),   fitMeasures(fit2, "aic")),
  BIC    = c(fitMeasures(fit1, "bic"),   fitMeasures(fit2, "bic"))
)
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** The chi-square difference is
`r fmt(anova(fit1, fit2)[2, "Chisq diff"], 2)` on 1 df, $p$ =
`r fmt(anova(fit1, fit2)[2, "Pr(>Chisq)"], 3)` -- **not** significant. AIC and
BIC are both **lower for Model 1**, agreeing that the extra parameter is not
worth it. So freeing the `TECH -> READ_AVG` path does *not* improve the model:
we keep the simpler Model 1. Substantively, once books at home is accounted for,
technology does not add explanatory power for reading in these data. (Note both
models fit the covariance matrix only moderately -- the comparison tells us the
tech path isn't the missing piece, not that Model 1 is a great model.)
:::

## Wrap-up

1. Why did we have to keep `TECH` in Model 1 instead of just dropping it?
2. Which single statistic in the comparison most directly answers the research
   question, and what does its value tell you?
3. If multivariate normality had been badly violated, what would you change
   about how you fit and compared these models?
