---
title: "W04 Lab: Model Fit & Comparison"
date: today
description: '<a href="rlab-downloads/w04_rlab_model_fit.qmd" download="w04_rlab_model_fit.qmd">Download .qmd</a>'
data: '<a href="rlab-downloads/w04_data_pisa22_USA_cln.csv" download="w04_data_pisa22_USA_cln.csv">w04_data_pisa22_USA_cln.csv</a>'
format: html
release: week 4
---

## Purpose

Ensure that you can independently run the MVPA models. 

You can adapt these for your own mini-project.

## Setup

Install packages if needed: 

```{r}

#install.packages("tidyverse")
#install.packages("lavaan")
#install.packages("lavaanPlot")
#install.packages("psych")

```

```{r}
#| message: false
library(tidyverse)
library(lavaan)
library(lavaanPlot)
library(psych)

pisa <- read_csv("w04_data_pisa22_USA_cln.csv", show_col_types = FALSE)
```

```{r}
#| include: false
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")

# Focused fit table: the headline indices with conventional good-fit cutoffs.
# With MLR we report the robust/scaled statistics.
fit_table <- function(fit) {
  m <- fitMeasures(fit, c("chisq.scaled", "df", "pvalue.scaled", "cfi.robust",
                          "rmsea.robust", "rmsea.ci.lower.robust",
                          "rmsea.ci.upper.robust", "srmr"))
  p <- if (m[["pvalue.scaled"]] < 0.001) "< .001" else
    paste0("= ", fmt(m[["pvalue.scaled"]], 3))
  rmsea <- sprintf("%s [%s, %s]", fmt(m[["rmsea.robust"]], 3),
                   fmt(m[["rmsea.ci.lower.robust"]], 3),
                   fmt(m[["rmsea.ci.upper.robust"]], 3))
  tibble(
    Index      = c("χ² (df)", "CFI", "RMSEA [90% CI]", "SRMR"),
    Value      = c(sprintf("%s (%d), p %s", fmt(m[["chisq.scaled"]], 1), as.integer(m[["df"]]), p),
                   fmt(m[["cfi.robust"]], 3), rmsea, fmt(m[["srmr"]], 3)),
    `Good fit` = c("p > .05", "≥ .95", "≤ .06", "≤ .08")
  )
}
```

## Assumptions

MVPA models estimated with maximum likelihood carry a handful of assumptions.

### Multivariate normality

ML assumes the endogenous variables are multivariate normal.

```{r}

mardia(pisa)

```

Interpretation: Significant results indicate evidence against multivariate normality, so we fit with a robust estimator, `estimator = "MLR"`,
which corrects the standard errors and the chi-square for non-normality.

::: {.content-visible when-profile="teacher"}
**Sample answer.** With a large PISA sample, mild departures from normality are
common and rarely fatal. If skew/kurtosis are moderate, ML point estimates stay
trustworthy but the chi-square and standard errors can be off -- that is exactly
what `MLR` fixes, so reporting the robust (scaled) statistics is the safe habit.
:::

### Linearity

The `~` paths assume each relationship is linear. Scatterplots with a straight
(`lm`) fit overlaid let us eyeball the two paths into reading.

```{r}
#| warning: false
#| fig-width: 8
#| fig-height: 3
pisa |>
  select(BOOKSHOME, TECH, READ_AVG) |>
  pivot_longer(-READ_AVG, names_to = "predictor", values_to = "x") |>
  ggplot(aes(x, READ_AVG)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, colour = "#C81E7A") +
  facet_wrap(~predictor, scales = "free_x") +
  theme_minimal()
```

### Multicollinearity

The two predictors that appear together in the reading equation are books at
home and technology. They should not be near-perfectly correlated.

```{r}
vars <- c("PAREDINT", "BOOKSHOME", "TECH", "READ_AVG")
cor(pisa[, vars], use = "pairwise.complete.obs")

# Variance inflation factor for the BOOKSHOME / TECH pair
r2  <- summary(lm(BOOKSHOME ~ TECH, data = pisa))$r.squared
vif <- 1 / (1 - r2)
vif
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Correlation between books at home and technology is
`r fmt(cor(pisa$BOOKSHOME, pisa$TECH, use = "pairwise.complete.obs"), 2)`, and
VIF = `r fmt(vif, 2)` (rule of thumb: concern above ~5-10). No collinearity
problem -- the two mediators carry distinct information.
:::

### Uncorrelated disturbances

Recursive path models assume the disturbances of the
endogenous variables are uncorrelated. We enforce this when fitting with
`auto.cov.y = FALSE`. 

After fitting we check the residual covariances.

# MVPA Models

One research decision note: 

We have two ordinal variables, but each has more than 5 levels, so we follow convention that 5 or more levels may be treated as continuous. 

## Model 0: Original Example

This is the model we've been looking at in class. 

Does parent education relate to reading achievement through books at home?

This question does not imply any model comparison. 

$$\text{PAREDINT} \;\rightarrow\; \text{BOOKSHOME} \;\rightarrow\; \text{READ\_AVG}$$

```{r}
#| warning: false
model0 <- "
  BOOKSHOME ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"

fit0 <- lavaan(model0, data = pisa, auto.var = TRUE, auto.cov.y = FALSE, estimator = "MLR")
summary(fit0, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

Residual check for the uncorrelated-disturbance assumption:

This shows standardized residual correlations. Close to 0 is evidence that the disturbances are uncorrelated.

```{r}
residuals(fit0, type = "cor")

```

### How well does Model 0 fit?

```{r}
knitr::kable(fit_table(fit0))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 0 is over-identified (df =
`r as.integer(fitMeasures(fit0, "df"))`), so its fit is testable. The robust
chi-square is `r fmt(fitMeasures(fit0, "chisq.scaled"), 1)` with $p$ =
`r fmt(fitMeasures(fit0, "pvalue.scaled"), 3)` (CFI =
`r fmt(fitMeasures(fit0, "cfi.robust"), 3)`, RMSEA =
`r fmt(fitMeasures(fit0, "rmsea.robust"), 3)`, SRMR =
`r fmt(fitMeasures(fit0, "srmr"), 3)`). This three-variable chain is the baseline
we build on -- next we bring technology into the model so the richer models stay
comparable on the same variables.
:::


## Question 2: 

Question: 

Is the relationship between parental education (`PAREDINT`) and
children's reading scores (`READ_AVG`) mediated by books at home (`BOOKSHOME`)
alone, or does technology (`TECH`) *also* carry part of the effect?

This question does imply model comparison: One where tech is a predictor of reading scores and one where it is not. 

## Model 1: Books-only mediation

```{r}
#| warning: false
model1 <- "
  BOOKSHOME ~ PAREDINT
  TECH      ~ PAREDINT
  READ_AVG  ~ BOOKSHOME
"

fit1 <- lavaan(model1, data = pisa, auto.var = TRUE, auto.cov.y = FALSE, estimator = "MLR")
summary(fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```



Residual check for the uncorrelated-disturbance assumption:

```{r}
residuals(fit1, type = "cor")
```

### How well does Model 1 fit?

```{r}
knitr::kable(fit_table(fit1))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 1 is over-identified (df =
`r as.integer(fitMeasures(fit1, "df"))`), so its fit is testable. The
robust chi-square is `r fmt(fitMeasures(fit1, "chisq.scaled"), 1)` with $p$ =
`r fmt(fitMeasures(fit1, "pvalue.scaled"), 3)`, and the approximate-fit indices (CFI =
`r fmt(fitMeasures(fit1, "cfi.robust"), 3)`, RMSEA =
`r fmt(fitMeasures(fit1, "rmsea.robust"), 3)`, SRMR =
`r fmt(fitMeasures(fit1, "srmr"), 3)`) are mediocre -- this simple structure
does not fully reproduce the four-variable covariance matrix. The question is
whether *adding the technology path* fixes that.
:::


## Model 2: Books + tech mediation

```{r}
#| warning: false
model2 <- "
  BOOKSHOME ~ PAREDINT
  TECH      ~ PAREDINT
  READ_AVG  ~ BOOKSHOME + TECH
"

fit2 <- lavaan(model2, data = pisa, auto.var = TRUE, auto.cov.y = FALSE, estimator = "MLR")
summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```



Residual check for the uncorrelated-disturbance assumption:

```{r}
residuals(fit2, type = "cor")
```

### How well does Model 2 fit?

```{r}
knitr::kable(fit_table(fit2))
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** Model 2 frees one more path (`READ_AVG ~ TECH`), so df =
`r as.integer(fitMeasures(fit2, "df"))`. Its fit is essentially unchanged from
Model 1 (CFI = `r fmt(fitMeasures(fit2, "cfi.robust"), 3)`, RMSEA =
`r fmt(fitMeasures(fit2, "rmsea.robust"), 3)`), and the `TECH -> READ_AVG` coefficient
itself is not significant. That is the first sign the extra path buys us little.
:::

## Is Model 2 better than Model 1?

Because Model 1 is nested in Model 2 (Model 1 = Model 2 with the
`TECH -> READ_AVG` path fixed to 0), we can compare them with a chi-square
difference test, and also weigh parsimony with AIC/BIC.

```{r}
anova(fit1, fit2)
```

```{r}
tibble(
  Model  = c("Model 1: books-only", "Model 2: books + tech"),
  df     = c(fitMeasures(fit1, "df"),    fitMeasures(fit2, "df")),
  chisq  = c(fitMeasures(fit1, "chisq.scaled"), fitMeasures(fit2, "chisq.scaled")),
  CFI    = c(fitMeasures(fit1, "cfi.robust"),   fitMeasures(fit2, "cfi.robust")),
  RMSEA  = c(fitMeasures(fit1, "rmsea.robust"), fitMeasures(fit2, "rmsea.robust")),
  AIC    = c(fitMeasures(fit1, "aic"),   fitMeasures(fit2, "aic")),
  BIC    = c(fitMeasures(fit1, "bic"),   fitMeasures(fit2, "bic"))
)
```

::: {.content-visible when-profile="teacher"}
**Sample answer.** The chi-square difference is
`r fmt(anova(fit1, fit2)[2, "Chisq diff"], 2)` on 1 df, $p$ =
`r fmt(anova(fit1, fit2)[2, "Pr(>Chisq)"], 3)` -- **not** significant. AIC and
BIC are both **lower for Model 1**, agreeing that the extra parameter is not
worth it. So freeing the `TECH -> READ_AVG` path does *not* improve the model:
we keep the simpler Model 1. Substantively, once books at home is accounted for,
technology does not add explanatory power for reading in these data. (Note both
models fit the covariance matrix only moderately -- the comparison tells us the
tech path isn't the missing piece, not that Model 1 is a great model.)
:::


