---
title: "W05 Lab: Working with Lavaan"
date: today
description: '<a href="rlab-downloads/w05_rlab_interpretation.qmd" download="w05_rlab_interpretation.qmd">Download .qmd</a>'
format: html
code-fold: true
code-summary: "Show code"
---

## Purpose

Practice building models in Lavaan & applying basic Lavaan arguments. 

[**NOTE:** This is all fake data, specifically generated to highlight specific aspects of the Lavaan model]{style="color:red;"}

Each scenario below gives you a research question, a description of the data,
and the necessary inputs to the model. 

In this rlab: 

1. Develop the appropriate model for the RQ, using Lavaan syntax.
2. Determine the correct arguments to set in the lavaan function.
3. Compare your results to the solutions
4. Iterate until you get the same answers (or can make an argument for why yours are justified)

## Setup

```{r}
#| message: false
#| code-fold: false
library(tidyverse)
library(lavaan)
```

::: {.content-visible when-profile="teacher"}
```{r}
#| include: false
# Teacher-only: path-diagram helper (coefficients + variances + disturbances).
# Lives in a teacher block so it is stripped from the student .qmd download.
library(semPlot)
path_diagram <- function(fit, layout = "tree2", cols = NULL, seed = 3) {
  # cols: a named matrix of (x, y) node positions -- e.g. exogenous in a left
  # column, mediators in the middle, outcomes on the right. Rows are reordered
  # to lavaan's node order automatically.
  opts <- list(whatLabels = "est", style = "ram",
               residuals = TRUE, intercepts = FALSE, thresholds = FALSE,
               exoCov = FALSE, nCharNodes = 0, sizeMan = 9,
               edge.label.cex = 0.95, edge.color = "black")
  set.seed(seed)
  if (!is.null(cols)) {
    ord <- do.call(semPaths,
                   c(list(fit, DoNotPlot = TRUE), opts))$graphAttributes$Nodes$labels
    layout <- cols[ord, , drop = FALSE]
  }
  do.call(semPaths, c(list(fit, layout = layout), opts))
}
```
:::

## Default model

Here is the `lavaan()` function with its defaults. The argument(s) you need to
adjust yourself are limited to those in this list. 

For all three scenarios, you will also need to set these. It is already set up in the templates.

- `auto.var = TRUE` -- freely estimate the variances/disturbances.
- `auto.cov.y = FALSE` -- keep endogenous disturbances uncorrelated (recursive model).

```r
lavaan(model = NULL, data = NULL, ordered = NULL, aux = NULL,
       sampling_weights = NULL,
       sample_cov = NULL, sample_mean = NULL, sample_th = NULL,
       sample_nobs = NULL,
       group = NULL, cluster = NULL, constraints = "",
       wls_v = NULL, nacov = NULL, ov_order = "model",
       slot_options = NULL, slot_par_table = NULL, slot_sample_stats = NULL,
       slot_data = NULL, slot_model = NULL, slot_cache = NULL,
       sloth1 = NULL,
       ...)
```

---

# Scenario 1

**Research question.** Among high-school students, does **test anxiety**
relate to **self-rated exam performance** *through* **test-avoidance
behavior** (procrastinating, skipping practice)?

**The data.** A survey of 500 students. Every variable is a Likert item.

| Variable  | Meaning                        | Scale                       |
|-----------|--------------------------------|-----------------------------|
| `anxiety` | "I feel anxious before tests"  | 1-4 (strongly disagree→agree) |
| `avoid`   | test-avoidance behavior        | 1-4 (never→often)           |
| `perf`    | self-rated exam performance    | 1-5 (very poor→excellent)   |

```{r}
#| code-fold: false
set.seed(625)
n <- 500
anx_s   <- rnorm(n)
avoid_s <- 0.55 * anx_s   + rnorm(n, 0, sqrt(1 - 0.55^2))
perf_s  <- -0.50 * avoid_s + 0.30 * anx_s + rnorm(n, 0, sqrt(1 - 0.50^2))

to4 <- function(z) {
  as.integer(cut(z, c(-Inf, -0.8, 0, 0.8, Inf), labels = 1:4))
}
to5 <- function(z) {
  as.integer(cut(z, c(-Inf, -1, -0.2, 0.5, 1.1, Inf), labels = 1:5))
}

anxiety_survey <- data.frame(
  anxiety = to4(anx_s),
  avoid   = to4(avoid_s),
  perf    = to5(perf_s)
)

head(anxiety_survey)
sapply(anxiety_survey, table)   # a few discrete categories each
```


```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
# Step 1 -- write the path model implied by the research question:
model1 <- "
  # ...your paths here...
"
# Step 2 -- fit it, choosing the argument(s) the data situation requires:
fit1 <- lavaan(model1, data = anxiety_survey )   # <- what's missing?

summary(fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.content-visible when-profile="teacher"}


```{r}
#| message: false
#| warning: false
model1 <- "
  avoid ~ anxiety
  perf  ~ avoid
"
fit1 <- lavaan(model1, data = anxiety_survey,
               ordered = c("anxiety", "avoid", "perf"),
               auto.var = TRUE, auto.cov.y = FALSE)
summary(fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.callout-note collapse="true" title="Path diagram — coefficients, variances & disturbances"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
path_diagram(fit1, cols = rbind(
  anxiety = c(0, 0), avoid = c(1, 0), perf = c(2, 0)))
```
:::

:::

---

# Scenario 2

**Research question.** Do **prior achievement** and **weekly study time**
predict students' **final exam** score *through* their score on a **practice
(mock) test**?

**The data.** There are four variables in this data (described below) and the sample size is (**N = 250**).

| Variable | Meaning                         |
|----------|---------------------------------|
| `prior`  | prior-achievement score         |
| `study`  | weekly study hours              |
| `mock`   | practice (mock) test score (different instrument from final test)      |
| `final`  | final exam score                |

```{r}
#| code-fold: false
test_cov <- matrix(c(
   91.393, 14.086, 56.418,  52.332,
   14.086,  6.357, 16.931,  16.577,
   56.418, 16.931, 85.823,  75.840,
   52.332, 16.577, 75.840, 117.496),
  nrow = 4, byrow = TRUE,
  dimnames = list(c("prior", "study", "mock", "final"),
                  c("prior", "study", "mock", "final")))

test_cov
```


```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
model2 <- "
  # ...your paths here...
"
fit2 <- lavaan(model2,
               auto.var = TRUE, auto.cov.y = FALSE)   # <- what do you have to do?
summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.content-visible when-profile="teacher"}


```{r}
#| message: false
#| warning: false
model2 <- "
  mock  ~ prior + study
  final ~ mock + prior
"
fit2 <- lavaan(model2,
               sample_cov  = test_cov,
               sample_nobs = 250,
               auto.var = TRUE, auto.cov.y = FALSE)
summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.callout-note collapse="true" title="Path diagram — coefficients, variances & disturbances"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5.5
path_diagram(fit2, cols = rbind(
  prior = c(0, 1), study = c(0, -1),
  mock  = c(1, 0), final = c(2, 0)))
```
:::

:::

---

# Scenario 3

**Research question.** In a district testing dataset, how do students'
**socioeconomic status** and **prior achievement** shape their **final exam**
score -- both directly and *through* their **motivation**, **test anxiety**,
**study time**, and **class attendance**?

**The data.** There are 7 variables, described below. **N = 500**, across seven variables:

| Variable  | Meaning                          |
|-----------|----------------------------------|
| `ses`     | socioeconomic-status index       |
| `prior`   | prior-year achievement score     |
| `motiv`   | academic motivation              |
| `anxiety` | test anxiety                     |
| `study`   | weekly study hours               |
| `attend`  | class attendance (%)             |
| `final`   | final exam score                 |

**The hypothesized structure.** The research team theorizes these relationships
(and *only* these -- everything else is left out on purpose):

- `ses` and `prior` are background variables (they may covary).
- `motiv` is predicted by `ses` and `prior`.
- `anxiety` is predicted by `motiv` and `prior`.
- `study` is predicted by `motiv` and `ses`.
- `attend` is predicted by `motiv` and `ses`.
- `final` is predicted by `prior`, `study`, `attend`, and `anxiety`.

```{r}
#| code-fold: false
big_cov <- matrix(c(
   0.944,   4.796,   4.289,  -1.864,   1.975,   1.274,   5.839,
   4.796, 103.878,  44.578, -29.646,  17.468,   8.459,  81.571,
   4.289,  44.578,  85.216, -26.365,  24.380,  12.595,  73.673,
  -1.864, -29.646, -26.365,  58.352,  -9.027,  -5.672, -49.709,
   1.975,  17.468,  24.380,  -9.027,  16.036,   4.613,  33.118,
   1.274,   8.459,  12.595,  -5.672,   4.613,  27.516,  20.719,
   5.839,  81.571,  73.673, -49.709,  33.118,  20.719, 169.422),
  nrow = 7, byrow = TRUE,
  dimnames = list(
    c("ses", "prior", "motiv", "anxiety", "study", "attend", "final"),
    c("ses", "prior", "motiv", "anxiety", "study", "attend", "final")))

big_cov
```


```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
model3 <- "
  # ...your paths here...
"
fit3 <- lavaan(model3,
               auto.var = TRUE, auto.cov.y = FALSE)   # <- what else do you need?
summary(fit3, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.content-visible when-profile="teacher"}


```{r}
#| message: false
#| warning: false
model3 <- "
  motiv   ~ ses + prior
  anxiety ~ motiv + prior
  study   ~ motiv + ses
  attend  ~ motiv + ses
  final   ~ prior + study + attend + anxiety
"
fit3 <- lavaan(model3,
               sample_cov  = big_cov,
               sample_nobs = 500,
               auto.var = TRUE, auto.cov.y = FALSE)
summary(fit3, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.callout-note collapse="true" title="Path diagram — coefficients, variances & disturbances"}
```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6
path_diagram(fit3, cols = rbind(
  ses     = c(0,  1.2), prior   = c(0, -1.2),
  motiv   = c(1,  2.4), anxiety = c(1,  0.8),
  study   = c(1, -0.8), attend  = c(1, -2.4),
  final   = c(2,  0)))
```
:::

:::

---

# Scenario 4

**Research question.** Does students' **classroom engagement** predict their
**final test** score *through* the amount of **test preparation** they do?

**The data.** A district pulled 400 students' records. Prep & final have missing data, due to students who transferred and absences on test day).

| Variable | Meaning              | Missing? |
|----------|----------------------|----------|
| `engage` | classroom engagement | complete |
| `prep`   | test prep            | ~19% NA  |
| `final`  | final test score     | ~25% NA  |
| `attend` | attendance rate      | complete |
| `prior`  | prior-year test score| complete |

```{r}
#| code-fold: false
set.seed(42)
n3 <- 400
prior  <- rnorm(n3)
attend <- 0.5 * prior + rnorm(n3)
engage <- 0.4 * prior + rnorm(n3)
prep   <- 0.5 * engage + rnorm(n3)
final  <- 0.5 * prep + 0.3 * engage + rnorm(n3)

district <- data.frame(engage, prep, final, attend, prior)

# Missingness that depends on the auxiliary variables (MAR)
district$final[rbinom(n3, 1, plogis(-1.2 - 0.8 * scale(attend))) == 1] <- NA
district$prep[rbinom(n3, 1, plogis(-1.6 - 0.7 * scale(prior))) == 1] <- NA

round(colMeans(is.na(district)) * 100, 1)   # % missing per variable
```


```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
# Step 1 -- write the path model implied by the research question
#           (only engage, prep, final -- attend & prior are NOT in it):
model4 <- "
  # ...your paths here...
"
# Step 2 -- fit it: use FIML AND borrow strength from attend & prior.
fit4 <- lavaan(model4, data = district,
               missing = "fiml",
               auto.var = TRUE, auto.cov.y = FALSE)   # <- add the aux vars how?
summary(fit4, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.content-visible when-profile="teacher"}

```{r}
#| message: false
#| warning: false
model4 <- "
  prep  ~ engage
  final ~ prep + engage
"
fit4 <- lavaan(model4, data = district,
               aux     = c("attend", "prior"),
               missing = "fiml",
               auto.var = TRUE, auto.cov.y = FALSE)
summary(fit4, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

::: {.callout-note collapse="true" title="Path diagram — coefficients, variances & disturbances"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5.5
path_diagram(fit4)
```
:::

:::

---




