---
title: "W07 Lab: Specifying CFA models in lavaan"
date: today
description: '<a href="rlab-downloads/w07_rlab_cfa_specification.qmd" download="w07_rlab_cfa_specification.qmd">Download .qmd</a>'
format: html
code-fold: true
code-summary: "Show code"
---

## Purpose

- Practice writing `lavaan()` model syntax for CFA models

Each scenario below gives you a research question and some background.

In this rlab, for each scenario:

1. Decide which latent variables there are, and which indicators measure each one.
2. Write the model in lavaan syntax (just the model, we won't fit it in this class). 
  - only change the Model, nothing else. 
3. Run the `semPaths()` code to check that the diagram matches the model you intended.
4. Refer to lecture slides for relevant symbols. 

::: {.callout-note appearance="simple"}
These are **fictional scenarios**, and the data are placeholders. Because we never fit the models, the numbers do not matter, only the structure.
:::

## Setup

```{r}
#| message: false
#| code-fold: false
library(lavaan)
library(semPlot)

# Placeholder data: random numbers with the right column names.
# lavaan() needs *some* data to build the model object; since we use
# do.fit = FALSE, the values are never used.
make_dat <- function(vars, n = 200) {
  as.data.frame(setNames(replicate(length(vars), rnorm(n), simplify = FALSE), vars))
}
```

::: {.content-visible when-profile="teacher"}
```{r}
#| include: false
# Teacher-only: one helper that draws the answer diagram, so each scenario
# below only has to supply its model. Stripped from the student .qmd download.
answer_diagram <- function(model, vars) {
  fit0 <- lavaan(model, data = make_dat(vars), do.fit = FALSE)
  semPaths(fit0, what = "path", whatLabels = "name", style = "lisrel",
           layout = "tree", residuals = TRUE, nCharNodes = 0,
           sizeMan = 8, sizeMan2 = 4)
}
```
:::


---

# Scenario 1

**Research question.** Do students' responses to four survey items reflect a single underlying **growth mindset**?

**Background.**

- A district gave 8th graders four Likert items about how they think about ability (1 = strongly disagree, 5 = strongly agree).
- Researchers believe the items are all driven by one construct, growth mindset.
- Each item also has its own unique influences (wording, mood on the day, etc.).
- **Scaling:** set the scale of the latent variable by fixing the **first loading to 1**.

| Variable | Item |
|----------|------|
| `develop`  | "My intelligence can be developed"        |
| `workhard` | "Working hard makes me smarter"           |
| `hardsubj` | "I can get better even at hard subjects"  |
| `mistakes` | "Mistakes help me learn"                  |

```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
vars1 <- c("develop", "workhard", "hardsubj", "mistakes")

model1 <- "
  # ...your model here...
"
fit1 <- lavaan(model1, data = make_dat(vars1), do.fit = FALSE)

semPaths(fit1, what = "path", whatLabels = "name", style = "lisrel",
         layout = "tree", residuals = TRUE, nCharNodes = 0,
         sizeMan = 8, sizeMan2 = 4)
```

::: {.content-visible when-profile="teacher"}
::: {.callout-tip collapse="true" title="Model answer: code (teacher)"}
```{r}
#| code-fold: false
vars1 <- c("develop", "workhard", "hardsubj", "mistakes")
model1 <- "
  mindset =~ 1*develop + workhard + hardsubj + mistakes
  mindset ~~ mindset
  develop ~~ develop
  workhard ~~ workhard
  hardsubj ~~ hardsubj
  mistakes ~~ mistakes
"
```
:::

::: {.callout-tip collapse="true" title="Model answer: diagram (teacher)"}
```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-align: center
answer_diagram(model1, vars1)
```
:::
:::

---

# Scenario 2

**Research question.** Are **reading comprehension** and **listening comprehension** in a second language (L2) distinct but related skills?

**Background.**

- Adult learners of English took a proficiency test with six tasks (scores in points).
- Theory says reading and listening are separate skills, but there is also evidence that they may be related.
- **Scaling:** set the scale of each latent variable by fixing its **first loading to 1**.

| Variable | Item |
|----------|------|
| `dialogue`     | listening to an everyday dialogue   |
| `mainidea`     | finding the main idea of a passage  |
| `lecture`      | listening to an academic lecture    |
| `vocab`        | vocabulary in context               |
| `instructions` | following spoken instructions       |
| `inference`    | inference from a text               |

```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
vars2 <- c("dialogue", "mainidea", "lecture", "vocab", "instructions", "inference")

model2 <- "
  # ...your model here...
"
fit2 <- lavaan(model2, data = make_dat(vars2), do.fit = FALSE)

semPaths(fit2, what = "path", whatLabels = "name", style = "lisrel",
         layout = "tree", residuals = TRUE, nCharNodes = 0,
         sizeMan = 8, sizeMan2 = 4)
```

::: {.content-visible when-profile="teacher"}
::: {.callout-tip collapse="true" title="Model answer: code (teacher)"}
```{r}
#| code-fold: false
vars2 <- c("dialogue", "mainidea", "lecture", "vocab", "instructions", "inference")
model2 <- "
  read   =~ 1*mainidea + vocab + inference
  listen =~ 1*dialogue + lecture + instructions
  read   ~~ read
  listen ~~ listen
  read   ~~ listen
  mainidea ~~ mainidea
  vocab ~~ vocab
  inference ~~ inference
  dialogue ~~ dialogue
  lecture ~~ lecture
  instructions ~~ instructions
"
```
:::

::: {.callout-tip collapse="true" title="Model answer: diagram (teacher)"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center
answer_diagram(model2, vars2)
```
:::
:::

---

# Scenario 3

**Research question.** Are **positive affect** and **negative affect** two separate dimensions of daily emotional experience among outpatients?

**Background.**

- Outpatients rated how often they experienced seven feelings and states over the past week (1 = never, 5 = always).
- Some items are expected to reflect positive affect and some negative affect.
- In this research team's theory, the two dimensions are **independent**: feeling more positive says nothing about feeling more (or less) negative.
- **Scaling:** set the scale of each latent variable by fixing its **variance to 1**.

| Variable | Item |
|----------|------|
| `distressed`   | distressed    |
| `enthusiastic` | enthusiastic  |
| `afraid`       | afraid        |
| `joyful`       | joyful        |
| `irritable`    | irritable     |
| `hungry`       | hungry        |
| `calm`         | calm          |

```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
vars3 <- c("distressed", "enthusiastic", "afraid", "joyful", "irritable",
           "hungry", "calm")

model3 <- "
  # ...your model here...
"
fit3 <- lavaan(model3, data = make_dat(vars3), do.fit = FALSE)

semPaths(fit3, what = "path", whatLabels = "name", style = "lisrel",
         layout = "tree", residuals = TRUE, nCharNodes = 0,
         sizeMan = 8, sizeMan2 = 4)
```

::: {.content-visible when-profile="teacher"}
::: {.callout-tip collapse="true" title="Model answer: code (teacher)"}
```{r}
#| code-fold: false
vars3 <- c("distressed", "enthusiastic", "afraid", "joyful", "irritable",
           "hungry", "calm")
model3 <- "
  # `hungry` is not part of either dimension, so it is left out of the model
  pa =~ NA*enthusiastic + joyful + calm
  na =~ NA*distressed + irritable + afraid
  pa ~~ 1*pa
  na ~~ 1*na
  enthusiastic ~~ enthusiastic
  joyful ~~ joyful
  calm ~~ calm
  distressed ~~ distressed
  irritable ~~ irritable
  afraid ~~ afraid
"
```
:::

::: {.callout-tip collapse="true" title="Model answer: diagram (teacher)"}
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center
answer_diagram(model3, vars3)
```
:::
:::

---

# Scenario 4

**Research question.** How are customers' **trust in a brand**, their **perceived product quality**, and their **purchase intention** related to one another?

**Background.**

- An online survey of shoppers asked shoppers a set of questions on 7-point agreement scales.
- All three constructs are expected to be correlated with one another.
- Two of the purchase-intention items were worded almost identically ("I plan to buy" and "I intend to buy"), so the researchers expect their unique parts (residuals) to **covary** beyond what purchase intention explains.
- **Scaling:** set the scale of each latent variable by fixing its **variance to 1** (so all loadings are free).

| Variable | Item |
|----------|------|
| `reliable`   | "The product is reliable"              |
| `buysoon`    | "I will buy this product soon"         |
| `trustbrand` | "I trust this brand"                   |
| `plantobuy`  | "I plan to buy this product"           |
| `promises`   | "This brand keeps its promises"        |
| `lasts`      | "The product lasts a long time"        |
| `intendbuy`  | "I intend to buy this product"         |
| `honest`     | "This brand is honest with customers"  |
| `wellmade`   | "The product is well made"             |
| `shopfreq`   | "I shop online at least once a week"  |

```{r}
#| eval: false
#| code-fold: false
# YOUR TURN
vars4 <- c("reliable", "buysoon", "trustbrand", "plantobuy", "promises",
           "lasts", "intendbuy", "shopfreq", "honest", "wellmade")

model4 <- "
  # ...your model here...
"
fit4 <- lavaan(model4, data = make_dat(vars4), do.fit = FALSE)

semPaths(fit4, what = "path", whatLabels = "name", style = "lisrel",
         layout = "tree", residuals = TRUE, nCharNodes = 0,
         sizeMan = 8, sizeMan2 = 4)
```

::: {.content-visible when-profile="teacher"}
::: {.callout-tip collapse="true" title="Model answer: code (teacher)"}
```{r}
#| code-fold: false
vars4 <- c("reliable", "buysoon", "trustbrand", "plantobuy", "promises",
           "lasts", "intendbuy", "shopfreq", "honest", "wellmade")
model4 <- "
  # `shopfreq` measures none of the three constructs, so it is left out
  trust  =~ NA*trustbrand + promises + honest
  qual   =~ NA*wellmade + reliable + lasts
  intent =~ NA*buysoon + plantobuy + intendbuy
  trust  ~~ 1*trust
  qual   ~~ 1*qual
  intent ~~ 1*intent
  trust  ~~ qual
  trust  ~~ intent
  qual   ~~ intent
  trustbrand ~~ trustbrand
  promises ~~ promises
  honest ~~ honest
  wellmade ~~ wellmade
  reliable ~~ reliable
  lasts ~~ lasts
  buysoon ~~ buysoon
  plantobuy ~~ plantobuy
  intendbuy ~~ intendbuy
  plantobuy ~~ intendbuy
"
```
:::

::: {.callout-tip collapse="true" title="Model answer: diagram (teacher)"}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: center
answer_diagram(model4, vars4)
```
:::
:::

---
